---
title: "ATAC clustering and visualization: setup"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---


```{r knitr_setup, include = FALSE}
# Echo code, hide warnings/messages, and cache expensive chunks.
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
```

## Purpose

Entry point for the ATAC chromatin-accessibility analysis of Atlantic salmon
(*Salmo salar*) and rainbow trout (*Oncorhynchus mykiss*). It installs/loads the
package stack, defines the `paths` list every downstream notebook reads, and
sources the shared helpers from `99_helpers.R`. Knit this notebook first;
`01..03` expect `paths` and the helper functions to be in scope.

### Where the input data comes from

- **IDR consensus peak BEDs** — fetched on first use from Salmobase
  (`robust_ATAC_peaks/annotated_peaks/`), cached under `cache/`. Nothing is
  shipped in the repo.
- **Per-species consensus expression RDS** — the one input not yet redistributed
  (the tables are large; a Zenodo deposit / regeneration from the
  `robust-open-chromatin/` pipeline is pending). For now they live locally under
  `data/atac/`, materialised from the manuscript-state OneDrive bundles by
  `scripts/extract_legacy_consensus_expression_rdata.R` (see the README).

## R package dependencies

CRAN + Bioconductor packages used across `00..03`: I/O of consensus expression
tables and BED peaks (`data.table`, `plyranges`, `rtracklayer`, `GenomicRanges`),
SOM fitting + dimensionality reduction (`kohonen`, `cluster`, `uwot`, `aweSOM`),
heatmaps (`ComplexHeatmap`, `circlize`), figure assembly (`gtable`, `gridExtra`),
and `tidyverse` + `viridis` for wrangling + palettes. They are auto-installed on
a fresh R session.

```{r r_packages}
# Full dependency set for this notebook tree, split by repository.
.cran_pkgs <- c("circlize", "tidyverse", "viridis", "data.table", "gtable",
                "gridExtra", "kohonen", "uwot", "cluster", "aweSOM")
.bioc_pkgs <- c("ComplexHeatmap", "plyranges", "rtracklayer", "GenomicRanges")

# .ensure_pkgs — install any listed package not already available (CRAN via
# install.packages, Bioconductor via BiocManager); a no-op on a configured machine.
.ensure_pkgs <- function(cran = character(), bioc = character()) {
  # Keep only the CRAN packages that cannot currently be loaded.
  miss_cran <- cran[!vapply(cran, requireNamespace, logical(1), quietly = TRUE)]
  # Same for Bioconductor packages.
  miss_bioc <- bioc[!vapply(bioc, requireNamespace, logical(1), quietly = TRUE)]
  # Install missing CRAN packages from a fixed mirror.
  if (length(miss_cran)) {
    message("[setup] installing CRAN packages: ", paste(miss_cran, collapse = ", "))
    install.packages(miss_cran, repos = "https://cloud.r-project.org")
  }
  # Install missing Bioconductor packages (installing BiocManager first if needed).
  if (length(miss_bioc)) {
    if (!requireNamespace("BiocManager", quietly = TRUE)) {
      install.packages("BiocManager", repos = "https://cloud.r-project.org")
    }
    message("[setup] installing Bioconductor packages: ", paste(miss_bioc, collapse = ", "))
    BiocManager::install(miss_bioc, ask = FALSE, update = FALSE)
  }
}
# Run the installer for this tree's dependency set.
.ensure_pkgs(.cran_pkgs, .bioc_pkgs)

# Attach the packages used unqualified by the helpers and notebooks.
suppressPackageStartupMessages({
  library(circlize)        # colorRamp2() colour scale for the heatmaps
  library(tidyverse)       # dplyr/ggplot2/stringr/readr/...
  library(viridis)         # cividis/inferno palettes
  library(ComplexHeatmap)  # ATAC SOM heatmaps
  library(data.table)      # fast table handling
  library(gtable)          # grob table manipulation for the composite
  library(gridExtra)       # arrange grobs into the composite
  library(kohonen)         # self-organising map fitting
  library(uwot)            # UMAP embeddings
  library(cluster)         # PAM super-clustering
  library(plyranges)       # tidy GRanges (as_granges, %over%)
  library(rtracklayer)     # BED/peak import
})

# Record exact package versions for provenance.
sessionInfo()
```

## Data paths

`paths` lists, for both species: the local consensus-expression RDS files, the
Salmobase IDR-consensus peak BED URLs (one per condition for BodyMap, one per
stage for DevMap), and the output directories. The peak BEDs are downloaded +
cached on first use; the RDS files must already be present under `data/atac/`
(see Purpose).

```{r paths_config}
# Salmobase robust ATAC peak files (per-condition IDR-consensus bigBeds)
# are the authoritative source for the per-context peak sets. The
# canonical lists live at:
#   https://salmobase.org/datafiles/datasets/Aqua-Faang/robust_ATAC_peaks/{salmon,trout}_files.txt
# We hard-code the URL set here so the notebook does not need to parse the
# index on every knit; the index file is also written to data/derived/ for
# provenance.
.peaks_root <- "https://salmobase.org/datafiles/datasets/Aqua-Faang/robust_ATAC_peaks/annotated_peaks"

.salmon_bodymap_conds <- c(
  "Brain_Immature_Female",  "Brain_Immature_Male",
  "Brain_Mature_Female",    "Brain_Mature_Male",
  "Gonad_Immature_Female",  "Gonad_Immature_Male",
  "Gonad_Mature_Female",    "Gonad_Mature_Male",
  "Liver_Immature_Female",  "Liver_Immature_Male",
  "Liver_Mature_Female",    "Liver_Mature_Male",
  "Muscle_Immature_Female", "Muscle_Immature_Male",
  "Muscle_Mature_Female",   "Muscle_Mature_Male"
)
.trout_bodymap_conds <- c(
  "Brain_Immature_Female",  "Brain_Immature_Male",
  "Brain_Mature_Female",    "Brain_Mature_Male",
  "Liver_Immature_Female",  "Liver_Immature_Male",
  "Liver_Mature_Female",    "Liver_Mature_Male",
  "Muscle_Immature_Female", "Muscle_Immature_Male",
  "Muscle_Mature_Female",   "Muscle_Mature_Male"
)
.devmap_stages <- c("LateBlastulation", "MidGastrulation",
                    "EarlySomitogenesis", "MidSomitogenesis",
                    "LateSomitogenesis")

paths <- list(
  consensus_expression_salmon = "data/atac/AtlanticSalmon_consensus_expression.RDS",
  consensus_expression_trout  = "data/atac/RainbowTrout_consensus_expression.RDS",
  idr_bodymap_salmon_urls = sprintf(
    "%s/AtlanticSalmon/BodyMap/AtlanticSalmon_ATAC_%s.bed",
    .peaks_root, .salmon_bodymap_conds),
  idr_devmap_salmon_urls = sprintf(
    "%s/AtlanticSalmon/DevMap/AtlanticSalmon_ATAC_%s.bed",
    .peaks_root, .devmap_stages),
  idr_bodymap_trout_urls = sprintf(
    "%s/RainbowTrout/BodyMap/RainbowTrout_ATAC_%s.bed",
    .peaks_root, .trout_bodymap_conds),
  idr_devmap_trout_urls = sprintf(
    "%s/RainbowTrout/DevMap/RainbowTrout_ATAC_%s.bed",
    .peaks_root, .devmap_stages),
  derived_dir = "data/derived/",
  results_dir = "results/"
)

dir.create(paths$derived_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(paths$results_dir, showWarnings = FALSE, recursive = TRUE)
```

## Helpers

```{r source_helpers}
# Load the shared helper functions (peak loaders, SOM wrappers, palettes,
# heatmap/UMAP builders) used by 01..03.
source("99_helpers.R")
```
