---
title: "ATAC clustering and visualization: setup"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---


```{r knitr_setup, include = FALSE}
# Echo code, hide warnings/messages, and cache the expensive chunks under cache/.
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  cache.path = "cache/00_setup/"
)
```

## Purpose

Entry point for the ATAC chromatin-accessibility analysis of Atlantic salmon
(*Salmo salar*) and rainbow trout (*Oncorhynchus mykiss*). It installs/loads the
package stack, defines the `paths` list every downstream notebook reads, and
sources the shared helpers from `99_helpers.R`. Knit this notebook first;
`01..03` expect `paths` and the helper functions to be in scope.

### Where the input data comes from

All inputs are the deposited manuscript assets shipped in this repository, so the
analysis runs offline and reproduces the published Figure 2B exactly:

- **Per-species consensus-expression tables** — `data/atac/*_consensus_expression.rds`
  (the unified robust-peak set × per-condition ATAC fold-enrichment).
- **Per-context SOM assets** — `data/atac_soms/` (the manuscript peak subsets,
  SOM unit assignments, super-clusters and UMAP coordinates).

## Working directory

The notebooks use paths relative to this folder
(`chromatin-accesibility-clustering-and-visualization/`). Knitting with
`rmarkdown::render()` or the RStudio **Knit** button handles this automatically.
If you instead run the chunks interactively, first set the working directory to
this folder. The check below fails early with a clear message rather than later
with a confusing "file not found".

```{r working_dir_check}
if (!file.exists("99_helpers.R")) {
  stop("Working directory is not the chromatin notebook folder.\n",
       "  Run the notebooks from ",
       "chromatin-accesibility-clustering-and-visualization/ ",
       "(e.g. setwd() to it, or open this .Rmd from that folder and Knit).\n",
       "  Current working directory: ", getwd())
}
```

## R package dependencies

CRAN + Bioconductor packages used across `00..03`: GRanges handling of the peak
coordinates (`plyranges`, `GenomicRanges`), heatmaps (`ComplexHeatmap`,
`circlize`), figure assembly (`gtable`, `gridExtra`), and `tidyverse` + `viridis`
for wrangling + palettes. They are auto-installed on a fresh R session.

```{r r_packages}
# Full dependency set for this notebook tree, split by repository.
.cran_pkgs <- c("circlize", "tidyverse", "viridis", "gtable", "gridExtra")
.bioc_pkgs <- c("ComplexHeatmap", "plyranges", "GenomicRanges")

# .ensure_pkgs — install any listed package not already available (CRAN via
# install.packages, Bioconductor via BiocManager); a no-op on a configured machine.
.ensure_pkgs <- function(cran = character(), bioc = character()) {
  # Keep only the CRAN packages that cannot currently be loaded.
  miss_cran <- cran[!vapply(cran, requireNamespace, logical(1), quietly = TRUE)]
  # Same for Bioconductor packages.
  miss_bioc <- bioc[!vapply(bioc, requireNamespace, logical(1), quietly = TRUE)]
  # Install missing CRAN packages from a fixed mirror.
  if (length(miss_cran)) {
    message("[setup] installing CRAN packages: ", paste(miss_cran, collapse = ", "))
    install.packages(miss_cran, repos = "https://cloud.r-project.org")
  }
  # Install missing Bioconductor packages (installing BiocManager first if needed).
  if (length(miss_bioc)) {
    if (!requireNamespace("BiocManager", quietly = TRUE)) {
      install.packages("BiocManager", repos = "https://cloud.r-project.org")
    }
    message("[setup] installing Bioconductor packages: ", paste(miss_bioc, collapse = ", "))
    BiocManager::install(miss_bioc, ask = FALSE, update = FALSE)
  }
}
# Run the installer for this tree's dependency set.
.ensure_pkgs(.cran_pkgs, .bioc_pkgs)

# Attach the packages used unqualified by the helpers and notebooks.
suppressPackageStartupMessages({
  library(circlize)        # colorRamp2() colour scale for the heatmaps
  library(tidyverse)       # dplyr/ggplot2/stringr/readr/...
  library(viridis)         # cividis/inferno palettes
  library(ComplexHeatmap)  # ATAC SOM heatmaps
  library(gtable)          # grob table manipulation for the composite
  library(gridExtra)       # arrange grobs into the composite
  library(plyranges)       # tidy GRanges (as_granges)
})

# Record exact package versions for provenance.
sessionInfo()
```

## Data paths

`paths` records, for both species, the deposited consensus-expression tables
that `01_atac_loading.Rmd` reads, plus the output directories.

```{r paths_config}
paths <- list(
  consensus_expression_salmon = "data/atac/AtlanticSalmon_consensus_expression.rds",
  consensus_expression_trout  = "data/atac/RainbowTrout_consensus_expression.rds",
  derived_dir = "data/derived/",
  results_dir = "results/"
)

dir.create(paths$derived_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(paths$results_dir, showWarnings = FALSE, recursive = TRUE)
```

## Check the deposited inputs are present

The consensus-expression tables and the SOM assets (read by `01` and `02`) ship
with the repository. Fail early with a clear message if a checkout is missing
them, rather than later with a confusing read error.

```{r check_deposited_inputs}
.needed <- c(paths$consensus_expression_salmon, paths$consensus_expression_trout,
             "data/atac_soms/devmap_salmon_idx.tsv.gz")
if (!all(file.exists(.needed))) {
  stop("deposited inputs are missing: ",
       paste(.needed[!file.exists(.needed)], collapse = ", "),
       ". They ship with the repository — re-checkout to restore them.")
}
```

## Helpers

```{r source_helpers}
# Load the shared helper functions (peak loaders, SOM wrappers, palettes,
# heatmap/UMAP builders) used by 01..03.
source("99_helpers.R")
```
