---
title: "ATAC clustering and visualization: setup"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---


```{r knitr_setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
```

## Purpose

Entry point for the ATAC chromatin-accessibility analysis
of Atlantic salmon (Salmo salar) and rainbow trout (Oncorhynchus mykiss). It
declares the R package dependencies, defines the on-disk paths used by every
downstream notebook, and sources the shared helper functions from
`99_helpers.R`. Knit this notebook first; downstream notebooks expect the
`paths` list and helper functions to be available.

## R package dependencies

The packages below cover I/O of consensus expression tables and IDR consensus
BED files (data.table, plyranges, rtracklayer), self-organising map fitting
and dimensionality reduction (kohonen, cluster, uwot), heatmap rendering
(ComplexHeatmap, circlize) and figure assembly (gtable, gridExtra). The
tidyverse + viridis pair supports data wrangling and the perceptually uniform
colour scales.

```{r r_packages}
.cran_pkgs <- c("circlize", "tidyverse", "viridis", "data.table", "gtable",
                "gridExtra", "kohonen", "uwot", "cluster", "aweSOM")
.bioc_pkgs <- c("ComplexHeatmap", "plyranges", "rtracklayer", "GenomicRanges")
.ensure_pkgs <- function(cran = character(), bioc = character()) {
  miss_cran <- cran[!vapply(cran, requireNamespace,
                            logical(1), quietly = TRUE)]
  miss_bioc <- bioc[!vapply(bioc, requireNamespace,
                            logical(1), quietly = TRUE)]
  if (length(miss_cran)) {
    message("[setup] installing CRAN packages: ",
            paste(miss_cran, collapse = ", "))
    install.packages(miss_cran, repos = "https://cloud.r-project.org")
  }
  if (length(miss_bioc)) {
    if (!requireNamespace("BiocManager", quietly = TRUE)) {
      install.packages("BiocManager", repos = "https://cloud.r-project.org")
    }
    message("[setup] installing Bioconductor packages: ",
            paste(miss_bioc, collapse = ", "))
    BiocManager::install(miss_bioc, ask = FALSE, update = FALSE)
  }
}
.ensure_pkgs(.cran_pkgs, .bioc_pkgs)

suppressPackageStartupMessages({
  library(circlize)
  library(tidyverse)
  library(viridis)
  library(ComplexHeatmap)
  library(data.table)
  library(gtable)
  library(gridExtra)
  library(kohonen)
  library(uwot)
  library(cluster)
  library(plyranges)
  library(rtracklayer)
})

sessionInfo()
```

## Data paths

Two modes. `legacy` reads the local derived assets that produced the
manuscript figures (consensus expression RDS and IDR consensus BED dirs)
and reproduces the figures bit-for-bit. `salmobase` points at the upstream
nf-core align outputs published on Salmobase; the IDR consensus BEDs and
consensus-expression RDS files are derived in the sibling
`robust-open-chromatin/` pipeline rather than republished by Salmobase, so
this mode is the entry point for a from-raw biological replication via
that pipeline.

```{r paths_config}
# Salmobase robust ATAC peak files (per-condition IDR-consensus bigBeds)
# are the authoritative source for the per-context peak sets. The
# canonical lists live at:
#   https://salmobase.org/datafiles/datasets/Aqua-Faang/robust_ATAC_peaks/{salmon,trout}_files.txt
# We hard-code the URL set here so the notebook does not need to parse the
# index on every knit; the index file is also written to data/derived/ for
# provenance.
.peaks_root <- "https://salmobase.org/datafiles/datasets/Aqua-Faang/robust_ATAC_peaks/annotated_peaks"

.salmon_bodymap_conds <- c(
  "Brain_Immature_Female",  "Brain_Immature_Male",
  "Brain_Mature_Female",    "Brain_Mature_Male",
  "Gonad_Immature_Female",  "Gonad_Immature_Male",
  "Gonad_Mature_Female",    "Gonad_Mature_Male",
  "Liver_Immature_Female",  "Liver_Immature_Male",
  "Liver_Mature_Female",    "Liver_Mature_Male",
  "Muscle_Immature_Female", "Muscle_Immature_Male",
  "Muscle_Mature_Female",   "Muscle_Mature_Male"
)
.trout_bodymap_conds <- c(
  "Brain_Immature_Female",  "Brain_Immature_Male",
  "Brain_Mature_Female",    "Brain_Mature_Male",
  "Liver_Immature_Female",  "Liver_Immature_Male",
  "Liver_Mature_Female",    "Liver_Mature_Male",
  "Muscle_Immature_Female", "Muscle_Immature_Male",
  "Muscle_Mature_Female",   "Muscle_Mature_Male"
)
.devmap_stages <- c("LateBlastulation", "MidGastrulation",
                    "EarlySomitogenesis", "MidSomitogenesis",
                    "LateSomitogenesis")

paths <- list(
  consensus_expression_salmon = "data/atac/AtlanticSalmon_consensus_expression.RDS",
  consensus_expression_trout  = "data/atac/RainbowTrout_consensus_expression.RDS",
  idr_bodymap_salmon_urls = sprintf(
    "%s/AtlanticSalmon/BodyMap/AtlanticSalmon_ATAC_%s.bed",
    .peaks_root, .salmon_bodymap_conds),
  idr_devmap_salmon_urls = sprintf(
    "%s/AtlanticSalmon/DevMap/AtlanticSalmon_ATAC_%s.bed",
    .peaks_root, .devmap_stages),
  idr_bodymap_trout_urls = sprintf(
    "%s/RainbowTrout/BodyMap/RainbowTrout_ATAC_%s.bed",
    .peaks_root, .trout_bodymap_conds),
  idr_devmap_trout_urls = sprintf(
    "%s/RainbowTrout/DevMap/RainbowTrout_ATAC_%s.bed",
    .peaks_root, .devmap_stages),
  derived_dir = "data/derived/",
  results_dir = "results/"
)

dir.create(paths$derived_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(paths$results_dir, showWarnings = FALSE, recursive = TRUE)
```

## Helpers

```{r source_helpers}
source("99_helpers.R")
```
