---
title: "ATAC clustering and visualization: setup"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---


```{r knitr_setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
```

## Purpose

Entry point for the ATAC chromatin-accessibility analysis
of Atlantic salmon (Salmo salar) and rainbow trout (Oncorhynchus mykiss). It
declares the R package dependencies, defines the on-disk paths used by every
downstream notebook, and sources the shared helper functions from
`99_helpers.R`. Knit this notebook first; downstream notebooks expect the
`paths` list and helper functions to be available.

## R package dependencies

The packages below cover I/O of consensus expression tables and IDR consensus
BED files (data.table, plyranges, rtracklayer), self-organising map fitting
and dimensionality reduction (kohonen, cluster, uwot), heatmap rendering
(ComplexHeatmap, circlize) and figure assembly (gtable, gridExtra). The
tidyverse + viridis pair supports data wrangling and the perceptually uniform
colour scales.

```{r r_packages}
suppressPackageStartupMessages({
  library(circlize)
  library(tidyverse)
  library(viridis)
  library(ComplexHeatmap)
  library(data.table)
  library(gtable)
  library(gridExtra)
  library(kohonen)
  library(uwot)
  library(cluster)
  library(plyranges)
  library(rtracklayer)
})

sessionInfo()
```

## Data paths

Two modes. `legacy` reads the local derived assets that produced the
manuscript figures (consensus expression RDS and IDR consensus BED dirs)
and reproduces the figures bit-for-bit. `salmobase` points at the upstream
nf-core align outputs published on Salmobase; the IDR consensus BEDs and
consensus-expression RDS files are derived in the sibling
`robust-open-chromatin/` pipeline rather than republished by Salmobase, so
this mode is the entry point for a from-raw biological replication via
that pipeline.

```{r paths_config}
paths <- list(
  mode = "legacy",
  legacy = list(
    consensus_expression_salmon = "data/atac/AtlanticSalmon_consensus_expression.RDS",
    consensus_expression_trout  = "data/atac/RainbowTrout_consensus_expression.RDS",
    idr_bodymap_salmon_dir = "data/atac/bodymap_idr/salmon/",
    idr_devmap_salmon_dir  = "data/atac/devmap_idr/salmon/",
    idr_bodymap_trout_dir  = "data/atac/bodymap_idr/trout/",
    idr_devmap_trout_dir   = "data/atac/devmap_idr/trout/",
    derived_dir            = "data/derived/",
    results_dir            = "results/"
  ),
  salmobase = list(
    # Browse-only: Salmobase publishes nf-core align outputs only.
    salmon_devmap_atac  = "https://salmobase.org/datafiles/datasets/Aqua-Faang/nfcore/AtlanticSalmon/DevMap/ATAC/results/",
    salmon_bodymap_atac = "https://salmobase.org/datafiles/datasets/Aqua-Faang/nfcore/AtlanticSalmon/BodyMap/ATAC/results/",
    trout_devmap_atac   = "https://salmobase.org/datafiles/datasets/Aqua-Faang/nfcore/RainbowTrout/DevMap/ATAC/results/",
    trout_bodymap_atac  = "https://salmobase.org/datafiles/datasets/Aqua-Faang/nfcore/RainbowTrout/BodyMap/ATAC/results/"
  )
)

dir.create(paths$legacy$derived_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(paths$legacy$results_dir, showWarnings = FALSE, recursive = TRUE)
```

## Helpers

```{r source_helpers}
source("99_helpers.R")
```
