---
title: "01 — ATAC loading: consensus expression and IDR peaks"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---


```{r knitr_setup, include = FALSE}
# Echo code, hide warnings/messages, and cache the expensive download/overlap
# chunks so re-knits are fast.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = TRUE)
```

## Purpose

Load the per-species consensus expression tables (RDS) and the per-context IDR
consensus peak BEDs (fetched from Salmobase) for Atlantic salmon and rainbow
trout, build the union of robust ATAC-seq peaks per context, then produce the
scaled fold-enrichment matrices that feed the SOM clustering in
`02_atac_soms.Rmd`. Outputs are written to
`data/derived/atac_matrices_{salmon,trout}.RData`.

Run `00_setup.Rmd` first so the `paths` list and helper functions are available.

```{r source_setup}
# Shared helper functions (I/O, peak union, scaled-FE matrix builder).
source("99_helpers.R")
# If run stand-alone (paths not yet defined), knit 00_setup.Rmd to build them.
if (!exists("paths")) knitr::knit("00_setup.Rmd", output = tempfile())
```

# Salmon

## Load the consensus expression table and IDR peak unions

The consensus expression RDS holds, for every peak in the per-species union, the
ATAC fragment counts across DevMap stages and BodyMap tissues. The IDR consensus
BEDs (downloaded from Salmobase) define the per-tissue / per-stage robust peak
sets; we reduce their union to the regions that pass IDR in at least one sample
of each context.

```{r load_consensus_expression_salmon}
# Read the salmon consensus expression table (one row per peak in the union).
ssal_consensus_expression <- readr::read_rds(paths$consensus_expression_salmon)

# Turn the chr/start/end/strand columns into a GRanges of all salmon peaks.
union_peaks_salmon <- ssal_consensus_expression %>%
  dplyr::rename(seqnames = chr) %>%                      # GRanges expects `seqnames`
  dplyr::mutate(start = as.numeric(start),               # coordinates as numeric
                end   = as.numeric(end),
                strand = stringr::str_replace(strand, stringr::fixed("."), "*")) %>%  # "." -> "*" (unstranded)
  plyranges::as_granges()                                # data frame -> GRanges
```

```{r load_idr_peaks_salmon_bodymap}
# Download + load the salmon BodyMap IDR consensus peak BEDs (one per condition).
bodyMap_idr_salmon <- load_idr_urls(paths$idr_bodymap_salmon_urls,
                                    species_prefix = "AtlanticSalmon_ATAC_")
# Reduce them to a single non-overlapping union of BodyMap robust peaks.
bodyMap_union_salmon <- union_granges(bodyMap_idr_salmon)
```

```{r load_idr_peaks_salmon_devmap}
# Same for the salmon DevMap IDR consensus BEDs (one per stage).
devMap_idr_salmon <- load_idr_urls(paths$idr_devmap_salmon_urls,
                                   species_prefix = "AtlanticSalmon_ATAC_")
devMap_union_salmon <- union_granges(devMap_idr_salmon)
```

## Subset to per-context peak sets and build SOM input matrices

Each context-specific matrix is scaled per region (fold enrichment over the
per-sample average, then z-scaled across samples without centering) so the SOM
operates on accessibility *shape* rather than magnitude.

```{r subset_to_devmap_ssal}
# Indices of union peaks that overlap a DevMap robust peak.
devMap_idx_salmon <- which(union_peaks_salmon %over% devMap_union_salmon)
# Scaled fold-enrichment matrix over the 5 DevMap ATAC columns (1:5).
devMap_in_mat_salmon <- build_scaled_fe_matrix(
  ssal_consensus_expression[devMap_idx_salmon, ],
  atac_col_idx = 1:5
)
```

```{r subset_to_bodymap_ssal}
# Indices of union peaks that overlap a BodyMap robust peak.
bodyMap_idx_salmon <- which(union_peaks_salmon %over% bodyMap_union_salmon)
# Scaled fold-enrichment matrix over the 16 BodyMap ATAC columns (6:21).
bodyMap_in_mat_salmon <- build_scaled_fe_matrix(
  ssal_consensus_expression[bodyMap_idx_salmon, ],
  atac_col_idx = 6:21
)
```

```{r save_atac_matrices_salmon}
# Persist the salmon SOM inputs + peak indices + union for 02 and 03.
save(devMap_in_mat_salmon, bodyMap_in_mat_salmon,
     devMap_idx_salmon, bodyMap_idx_salmon,
     union_peaks_salmon,
     file = file.path(paths$derived_dir, "atac_matrices_salmon.RData"))
```

# Trout

## Load the consensus expression table and IDR peak unions

```{r load_consensus_expression_trout}
# Read the trout consensus expression table.
omyk_consensus_expression <- readr::read_rds(paths$consensus_expression_trout)

# Build the GRanges of all trout peaks (same recipe as salmon).
union_peaks_trout <- omyk_consensus_expression %>%
  dplyr::rename(seqnames = chr) %>%
  dplyr::mutate(start = as.numeric(start),
                end   = as.numeric(end),
                strand = stringr::str_replace(strand, stringr::fixed("."), "*")) %>%
  plyranges::as_granges()
```

```{r load_idr_peaks_trout_bodymap}
# Download + load the trout BodyMap IDR consensus BEDs, then union them.
bodyMap_idr_trout <- load_idr_urls(paths$idr_bodymap_trout_urls,
                                   species_prefix = "RainbowTrout_ATAC_")
bodyMap_union_trout <- union_granges(bodyMap_idr_trout)
```

```{r load_idr_peaks_trout_devmap}
# Download + load the trout DevMap IDR consensus BEDs, then union them.
devMap_idr_trout <- load_idr_urls(paths$idr_devmap_trout_urls,
                                  species_prefix = "RainbowTrout_ATAC_")
devMap_union_trout <- union_granges(devMap_idr_trout)
```

## Subset to per-context peak sets and build SOM input matrices

```{r subset_to_devmap_trout}
# DevMap: overlap indices then the scaled FE matrix over columns 1:5.
devMap_idx_trout <- which(union_peaks_trout %over% devMap_union_trout)
devMap_in_mat_trout <- build_scaled_fe_matrix(
  omyk_consensus_expression[devMap_idx_trout, ],
  atac_col_idx = 1:5
)
```

```{r subset_to_bodymap_trout}
# BodyMap: overlap indices then the scaled FE matrix over columns 6:17
# (trout BodyMap has 12 ATAC columns, vs 16 in salmon).
bodyMap_idx_trout <- which(union_peaks_trout %over% bodyMap_union_trout)
bodyMap_in_mat_trout <- build_scaled_fe_matrix(
  omyk_consensus_expression[bodyMap_idx_trout, ],
  atac_col_idx = 6:17
)
```

```{r save_atac_matrices_trout}
# Persist the trout SOM inputs + peak indices + union for 02 and 03.
save(devMap_in_mat_trout, bodyMap_in_mat_trout,
     devMap_idx_trout, bodyMap_idx_trout,
     union_peaks_trout,
     file = file.path(paths$derived_dir, "atac_matrices_trout.RData"))
```
