---
title: "01 — ATAC loading: consensus expression and peak subsets"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---


```{r knitr_setup, include = FALSE}
# Echo code, hide warnings/messages, and cache the expensive chunks under cache/.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = TRUE, cache.path = "cache/01_atac_loading/")
```

## Purpose

Load the per-species consensus expression tables and the deposited per-context
peak subsets (the manuscript `data/atac_soms/*_idx.tsv.gz`) for Atlantic salmon
and rainbow trout, then produce the scaled fold-enrichment matrices that feed the
SOM clustering in `02_atac_soms.Rmd`. Outputs are written to
`data/derived/atac_matrices_{salmon,trout}.RData`.

Run `00_setup.Rmd` first so the `paths` list and helper functions are available.

```{r source_setup}
# Shared helper functions (I/O, scaled-FE matrix builder).
source("99_helpers.R")
# If run stand-alone (paths not yet defined), knit 00_setup.Rmd to build them.
if (!exists("paths")) knitr::knit("00_setup.Rmd", output = tempfile())
```

# Salmon

## Load the consensus expression table

The consensus expression table holds, for every peak in the per-species unified
set, the ATAC fold-enrichment across DevMap stages and BodyMap tissues. Each
context is then restricted to the deposited manuscript peak subset (the regions
robust in that context).

```{r load_consensus_expression_salmon}
# Read the salmon consensus expression table (one row per peak in the union).
ssal_consensus_expression <- readr::read_rds(paths$consensus_expression_salmon)

# Turn the chr/start/end/strand columns into a GRanges of all salmon peaks.
union_peaks_salmon <- ssal_consensus_expression %>%
  dplyr::rename(seqnames = chr) %>%                      # GRanges expects `seqnames`
  dplyr::mutate(start = as.numeric(start),               # coordinates as numeric
                end   = as.numeric(end),
                strand = stringr::str_replace(strand, stringr::fixed("."), "*")) %>%  # "." -> "*" (unstranded)
  plyranges::as_granges()                                # data frame -> GRanges
```

## Subset to per-context peak sets and build SOM input matrices

Each context-specific matrix is scaled per region (fold enrichment over the
per-sample average, then z-scaled across samples without centering) so the SOM
operates on accessibility *shape* rather than magnitude.

```{r subset_to_devmap_ssal}
# DevMap peak subset: the deposited manuscript indices.
devMap_idx_salmon <- read_atac_som_asset("devmap_salmon", "idx")
# Scaled fold-enrichment matrix over the 5 DevMap ATAC columns (1:5).
devMap_in_mat_salmon <- build_scaled_fe_matrix(
  ssal_consensus_expression[devMap_idx_salmon, ],
  atac_col_idx = 1:5
)
```

```{r subset_to_bodymap_ssal}
# BodyMap peak subset: the deposited manuscript indices.
bodyMap_idx_salmon <- read_atac_som_asset("bodymap_salmon", "idx")
# Scaled fold-enrichment matrix over the 16 BodyMap ATAC columns (6:21).
bodyMap_in_mat_salmon <- build_scaled_fe_matrix(
  ssal_consensus_expression[bodyMap_idx_salmon, ],
  atac_col_idx = 6:21
)
```

```{r save_atac_matrices_salmon}
# Persist the salmon SOM inputs + peak indices + union for 02 and 03.
save(devMap_in_mat_salmon, bodyMap_in_mat_salmon,
     devMap_idx_salmon, bodyMap_idx_salmon,
     union_peaks_salmon,
     file = file.path(paths$derived_dir, "atac_matrices_salmon.RData"))
```

# Trout

## Load the consensus expression table

```{r load_consensus_expression_trout}
# Read the trout consensus expression table.
omyk_consensus_expression <- readr::read_rds(paths$consensus_expression_trout)

# Build the GRanges of all trout peaks (same recipe as salmon).
union_peaks_trout <- omyk_consensus_expression %>%
  dplyr::rename(seqnames = chr) %>%
  dplyr::mutate(start = as.numeric(start),
                end   = as.numeric(end),
                strand = stringr::str_replace(strand, stringr::fixed("."), "*")) %>%
  plyranges::as_granges()
```

## Subset to per-context peak sets and build SOM input matrices

```{r subset_to_devmap_trout}
# DevMap subset: the deposited manuscript indices.
devMap_idx_trout <- read_atac_som_asset("devmap_trout", "idx")
devMap_in_mat_trout <- build_scaled_fe_matrix(
  omyk_consensus_expression[devMap_idx_trout, ],
  atac_col_idx = 1:5
)
```

```{r subset_to_bodymap_trout}
# BodyMap subset: the deposited manuscript indices. Trout BodyMap has 12 ATAC
# columns (6:17), vs 16 in salmon.
bodyMap_idx_trout <- read_atac_som_asset("bodymap_trout", "idx")
bodyMap_in_mat_trout <- build_scaled_fe_matrix(
  omyk_consensus_expression[bodyMap_idx_trout, ],
  atac_col_idx = 6:17
)
```

```{r save_atac_matrices_trout}
# Persist the trout SOM inputs + peak indices + union for 02 and 03.
save(devMap_in_mat_trout, bodyMap_in_mat_trout,
     devMap_idx_trout, bodyMap_idx_trout,
     union_peaks_trout,
     file = file.path(paths$derived_dir, "atac_matrices_trout.RData"))
```
