---
title: "ATAC loading: consensus expression and IDR peaks"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---


```{r knitr_setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = TRUE)
```

## Purpose

Load the per-species consensus expression tables (RDS) and the per-context
IDR consensus BED files for Atlantic salmon and rainbow trout, build the
union of robust ATAC-seq peaks per context, then produce the scaled fold-
enrichment matrices that feed the SOM clustering in `02_atac_soms.Rmd`.
Outputs are written to `data/derived/atac_matrices_{salmon,trout}.RData`.

Run `00_setup.Rmd` first so that the `paths` list and helper functions are
available.

```{r source_setup}
source("99_helpers.R")
# Re-evaluate the paths chunk from 00_setup.Rmd if running this notebook
# stand-alone.
if (!exists("paths")) knitr::knit("00_setup.Rmd", output = tempfile())
```

# Salmon

## Load the consensus expression table and IDR peak unions

The consensus expression RDS holds, for every peak in the per-species union,
the ATAC fragment counts across DevMap stages and BodyMap tissues. The IDR
consensus BEDs define the per-tissue / per-stage robust peak sets; we reduce
their union to the set of regions that pass IDR in at least one sample of
each context.

```{r load_consensus_expression_salmon}
ssal_consensus_expression <- readr::read_rds(paths$legacy$consensus_expression_salmon)

union_peaks_salmon <- ssal_consensus_expression %>%
  dplyr::rename(seqnames = chr) %>%
  dplyr::mutate(start = as.numeric(start),
                end   = as.numeric(end),
                strand = stringr::str_replace(strand, stringr::fixed("."), "*")) %>%
  plyranges::as_granges()
```

```{r load_idr_peaks_salmon_bodymap}
bodyMap_idr_salmon <- load_idr_dir(paths$legacy$idr_bodymap_salmon_dir,
                                   species_prefix = "AtanticSalmon_ATAC_")
bodyMap_union_salmon <- union_granges(bodyMap_idr_salmon)
```

```{r load_idr_peaks_salmon_devmap}
devMap_idr_salmon <- load_idr_dir(paths$legacy$idr_devmap_salmon_dir,
                                  species_prefix = "")
devMap_union_salmon <- union_granges(devMap_idr_salmon)
```

## Subset to per-context peak sets and build SOM input matrices

Each context-specific matrix is scaled per region (fold enrichment over the
per-sample average, then z-scaled across samples without centering) so the
SOM operates on shape rather than magnitude.

```{r subset_to_devmap_ssal}
devMap_idx_salmon <- which(union_peaks_salmon %over% devMap_union_salmon)
devMap_in_mat_salmon <- build_scaled_fe_matrix(
  ssal_consensus_expression[devMap_idx_salmon, ],
  atac_col_idx = 1:5
)
```

```{r subset_to_bodymap_ssal}
bodyMap_idx_salmon <- which(union_peaks_salmon %over% bodyMap_union_salmon)
bodyMap_in_mat_salmon <- build_scaled_fe_matrix(
  ssal_consensus_expression[bodyMap_idx_salmon, ],
  atac_col_idx = 6:21
)
```

```{r save_atac_matrices_salmon}
save(devMap_in_mat_salmon, bodyMap_in_mat_salmon,
     devMap_idx_salmon, bodyMap_idx_salmon,
     union_peaks_salmon,
     file = file.path(paths$legacy$derived_dir, "atac_matrices_salmon.RData"))
```

# Trout

## Load the consensus expression table and IDR peak unions

The trout BodyMap IDR BEDs carry extra ChromHMM annotation columns; we read
them through the dedicated `read_trout_bodymap_bed` helper rather than
rtracklayer::import.

```{r load_consensus_expression_trout}
omyk_consensus_expression <- readr::read_rds(paths$legacy$consensus_expression_trout)

union_peaks_trout <- omyk_consensus_expression %>%
  dplyr::rename(seqnames = chr) %>%
  dplyr::mutate(start = as.numeric(start),
                end   = as.numeric(end),
                strand = stringr::str_replace(strand, stringr::fixed("."), "*")) %>%
  plyranges::as_granges()
```

```{r load_idr_peaks_trout_bodymap}
bodyMap_idr_trout <- load_idr_dir(paths$legacy$idr_bodymap_trout_dir,
                                  species_prefix = "RainbowTrout_ATAC_",
                                  reader = read_trout_bodymap_bed)
bodyMap_union_trout <- union_granges(bodyMap_idr_trout)
```

```{r load_idr_peaks_trout_devmap}
devMap_idr_trout <- load_idr_dir(paths$legacy$idr_devmap_trout_dir,
                                 species_prefix = "")
devMap_union_trout <- union_granges(devMap_idr_trout)
```

## Subset to per-context peak sets and build SOM input matrices

```{r subset_to_devmap_trout}
devMap_idx_trout <- which(union_peaks_trout %over% devMap_union_trout)
devMap_in_mat_trout <- build_scaled_fe_matrix(
  omyk_consensus_expression[devMap_idx_trout, ],
  atac_col_idx = 1:5
)
```

```{r subset_to_bodymap_trout}
bodyMap_idx_trout <- which(union_peaks_trout %over% bodyMap_union_trout)
bodyMap_in_mat_trout <- build_scaled_fe_matrix(
  omyk_consensus_expression[bodyMap_idx_trout, ],
  atac_col_idx = 6:17
)
```

```{r save_atac_matrices_trout}
save(devMap_in_mat_trout, bodyMap_in_mat_trout,
     devMap_idx_trout, bodyMap_idx_trout,
     union_peaks_trout,
     file = file.path(paths$legacy$derived_dir, "atac_matrices_trout.RData"))
```
