---
title: "ATAC SOMs: clustering of accessibility profiles"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---


```{r knitr_setup, include = FALSE}
# Echo code, hide warnings/messages, and cache the expensive chunks under cache/.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = TRUE, cache.path = "cache/02_atac_soms/")
```

## Purpose

Assemble the four context-specific accessibility SOMs (DevMap salmon, BodyMap
salmon, DevMap trout, BodyMap trout) from the deposited manuscript assets under
`data/atac_soms/` — the per-peak unit assignments, PAM super-clusters and UMAP
embeddings used in Figure 2B — together with the scaled fold-enrichment matrices
from `01_atac_loading.Rmd`, then build the ribbon-plot summary frames. Loading
the deposited SOMs (rather than re-fitting) reproduces the published figures
exactly. Outputs are written to `data/derived/atac_soms.RData`.

```{r source_setup}
# Shared helpers (SOM summary wrapper, ribbon-panel builder).
source("99_helpers.R")
# If run stand-alone, knit 00_setup.Rmd to define `paths`.
if (!exists("paths")) knitr::knit("00_setup.Rmd", output = tempfile())
# The four scaled fold-enrichment SOM-input matrices built by 01. Fail early with
# a clear message if 01 has not been run yet.
.mat_files <- file.path(paths$derived_dir,
                        c("atac_matrices_salmon.RData", "atac_matrices_trout.RData"))
if (!all(file.exists(.mat_files))) {
  stop("Missing ", paste(basename(.mat_files[!file.exists(.mat_files)]),
                         collapse = " and "), " in ", paths$derived_dir,
       ". Knit 01_atac_loading.Rmd first (the notebooks run 00 -> 01 -> 02 -> 03).")
}
load(.mat_files[1])
load(.mat_files[2])
```

## Load the deposited SOMs

Reconstructs the four SOM objects from the deposited manuscript unit assignments
under `data/atac_soms/`. Only `$unit.classif` is read downstream (by `get_som`
and the Fig-2B heatmap/UMAP builders), so that is all the reconstructed objects
need to carry — the prototype codes and distance matrices are not required.

```{r load_deposited_atac_soms}
# Inject the four SOMs from data/atac_soms/ (row-aligned to the 01 matrices).
devMap_atac_som_salmon  <- list(unit.classif =
  read_atac_som_asset("devmap_salmon",  "unit_classif"))
bodyMap_atac_som_salmon <- list(unit.classif =
  read_atac_som_asset("bodymap_salmon", "unit_classif"))
devMap_atac_som_trout   <- list(unit.classif =
  read_atac_som_asset("devmap_trout",   "unit_classif"))
bodyMap_atac_som_trout  <- list(unit.classif =
  read_atac_som_asset("bodymap_trout",  "unit_classif"))
```

## PAM super-clusters

The four coarse super-clusters per SOM (PAM `k = 4` on the prototype codes in the
manuscript), labelling the dynamic vs. constitutive units, are loaded from the
deposited assets.

```{r superclasses_pam}
superclasses_devmap_salmon  <- read_atac_som_asset("devmap_salmon",  "superclasses")
superclasses_devmap_trout   <- read_atac_som_asset("devmap_trout",   "superclasses")
superclasses_bodymap_salmon <- read_atac_som_asset("bodymap_salmon", "superclasses")
superclasses_bodymap_trout  <- read_atac_som_asset("bodymap_trout",  "superclasses")
```

## Label units and ribbon summary frames

`get_som` rebuilds, for each SOM, the per-unit label vector, the long tidy
frame and the per-unit mean / SD table used by the ribbon plots and by
`03_fig2B_atac_soms.Rmd`.

```{r build_label_units}
# For each SOM, build the per-unit label vector + long tidy frame + per-unit
# mean/SD table (consumed by the ribbon panels and by 03_fig2B_atac_soms.Rmd).
sum_devmap_salmon  <- get_som(devMap_in_mat_salmon,  5, 5, devMap_atac_som_salmon)
sum_devmap_trout   <- get_som(devMap_in_mat_trout,   5, 5, devMap_atac_som_trout)
sum_bodymap_salmon <- get_som(bodyMap_in_mat_salmon, 5, 5, bodyMap_atac_som_salmon)
sum_bodymap_trout  <- get_som(bodyMap_in_mat_trout,  5, 5, bodyMap_atac_som_trout)
```

## Ribbon panels

The four facet_wrap ribbon panels (mean +/- 2 SD per SOM unit across
samples) for the DevMap and BodyMap contexts of each species.

```{r ribbon_panels}
# One facet_wrap ribbon panel (mean +/- 2 SD per SOM unit) per context/species.
ribbon_devmap_salmon  <- make_ribbon_panel(sum_devmap_salmon$tmp_df,
                                           sum_devmap_salmon$label_unit)
ribbon_devmap_trout   <- make_ribbon_panel(sum_devmap_trout$tmp_df,
                                           sum_devmap_trout$label_unit)
ribbon_bodymap_salmon <- make_ribbon_panel(sum_bodymap_salmon$tmp_df,
                                           sum_bodymap_salmon$label_unit)
ribbon_bodymap_trout  <- make_ribbon_panel(sum_bodymap_trout$tmp_df,
                                           sum_bodymap_trout$label_unit)
```

## UMAP embeddings

The 2-D UMAP embedding of each scaled fold-enrichment matrix (used for the
SOM-coloured scatterplots in Figure 2B) is loaded from the deposited assets.

```{r umap_atac}
devMap_umap_salmon  <- read_atac_som_asset("devmap_salmon",  "umap")
devMap_umap_trout   <- read_atac_som_asset("devmap_trout",   "umap")
bodyMap_umap_salmon <- read_atac_som_asset("bodymap_salmon", "umap")
bodyMap_umap_trout  <- read_atac_som_asset("bodymap_trout",  "umap")
```

## Persist SOM objects

```{r save_atac_som_objects}
# Persist the SOMs, super-clusters, ribbon summaries and UMAPs for 03.
save(devMap_atac_som_salmon,  devMap_atac_som_trout,
     bodyMap_atac_som_salmon, bodyMap_atac_som_trout,
     superclasses_devmap_salmon,  superclasses_devmap_trout,
     superclasses_bodymap_salmon, superclasses_bodymap_trout,
     sum_devmap_salmon,  sum_devmap_trout,
     sum_bodymap_salmon, sum_bodymap_trout,
     ribbon_devmap_salmon,  ribbon_devmap_trout,
     ribbon_bodymap_salmon, ribbon_bodymap_trout,
     devMap_umap_salmon,  devMap_umap_trout,
     bodyMap_umap_salmon, bodyMap_umap_trout,
     file = file.path(paths$derived_dir, "atac_soms.RData"))
```
