---
title: "ATAC SOMs: clustering of accessibility profiles"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---


```{r knitr_setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = TRUE)
```

## Purpose

Fit 5 x 5 hexagonal self-organising maps to the four context-specific
ATAC matrices produced by `01_atac_loading.Rmd` (DevMap salmon, BodyMap
salmon, DevMap trout, BodyMap trout), derive PAM super-clusters on the
prototype codes, build the ribbon-plot summary frames used in Figure 2,
and reduce every matrix to two dimensions with uwot::umap. Outputs are
written to `data/derived/atac_soms.RData`.

```{r source_setup}
source("99_helpers.R")
if (!exists("paths")) knitr::knit("00_setup.Rmd", output = tempfile())
load(file.path(paths$legacy$derived_dir, "atac_matrices_salmon.RData"))
load(file.path(paths$legacy$derived_dir, "atac_matrices_trout.RData"))
```

## Seed lock

Both the somInit initialisation and the kohonen::som training are seeded
with 12345 to match the manuscript figures. The seed is applied inside
`fit_atac_som`; this chunk documents the choice.

```{r seed_lock_atac}
atac_seed <- 12345
atac_seed
```

## SOM fits

Each fit uses the same 5 x 5 hexagonal grid with the legacy somInit
initialisation followed by kohonen::som.

```{r fit_devmap_som_salmon, cache = TRUE}
devMap_atac_som_salmon <- fit_atac_som(devMap_in_mat_salmon,
                                       xdim = 5, ydim = 5,
                                       seed = atac_seed)
```

```{r fit_devmap_som_trout, cache = TRUE}
devMap_atac_som_trout <- fit_atac_som(devMap_in_mat_trout,
                                      xdim = 5, ydim = 5,
                                      seed = atac_seed)
```

```{r fit_bodymap_som_salmon, cache = TRUE}
bodyMap_atac_som_salmon <- fit_atac_som(bodyMap_in_mat_salmon,
                                        xdim = 5, ydim = 5,
                                        seed = atac_seed)
```

```{r fit_bodymap_som_trout, cache = TRUE}
bodyMap_atac_som_trout <- fit_atac_som(bodyMap_in_mat_trout,
                                       xdim = 5, ydim = 5,
                                       seed = atac_seed)
```

## PAM super-clusters

Each SOM's 25 prototype codes are summarised into four super-clusters with
PAM, giving the coarse pattern groupings that label the dynamic vs.
constitutive units.

```{r superclasses_pam}
superclasses_devmap_salmon  <- pam_superclusters(devMap_atac_som_salmon, k = 4)
superclasses_devmap_trout   <- pam_superclusters(devMap_atac_som_trout,  k = 4)
superclasses_bodymap_salmon <- pam_superclusters(bodyMap_atac_som_salmon, k = 4)
superclasses_bodymap_trout  <- pam_superclusters(bodyMap_atac_som_trout,  k = 4)
```

## Label units and ribbon summary frames

`get_som` rebuilds, for each SOM, the per-unit label vector, the long tidy
frame and the per-unit mean / SD table used by the ribbon plots and by
`03_figure2_atac.Rmd`.

```{r build_label_units}
sum_devmap_salmon  <- get_som(devMap_in_mat_salmon,  5, 5, devMap_atac_som_salmon)
sum_devmap_trout   <- get_som(devMap_in_mat_trout,   5, 5, devMap_atac_som_trout)
sum_bodymap_salmon <- get_som(bodyMap_in_mat_salmon, 5, 5, bodyMap_atac_som_salmon)
sum_bodymap_trout  <- get_som(bodyMap_in_mat_trout,  5, 5, bodyMap_atac_som_trout)
```

## Ribbon panels

The four facet_wrap ribbon panels (mean +/- 2 SD per SOM unit across
samples) for the DevMap and BodyMap contexts of each species.

```{r ribbon_panels}
ribbon_devmap_salmon  <- make_ribbon_panel(sum_devmap_salmon$tmp_df,
                                           sum_devmap_salmon$label_unit)
ribbon_devmap_trout   <- make_ribbon_panel(sum_devmap_trout$tmp_df,
                                           sum_devmap_trout$label_unit)
ribbon_bodymap_salmon <- make_ribbon_panel(sum_bodymap_salmon$tmp_df,
                                           sum_bodymap_salmon$label_unit)
ribbon_bodymap_trout  <- make_ribbon_panel(sum_bodymap_trout$tmp_df,
                                           sum_bodymap_trout$label_unit)
```

## UMAP embeddings

`uwot::umap` reduces each scaled fold-enrichment matrix to two dimensions
for the SOM-coloured scatterplots in Figure 2.

```{r umap_atac, cache = TRUE}
set.seed(atac_seed)
devMap_umap_salmon  <- uwot::umap(devMap_in_mat_salmon)
set.seed(atac_seed)
devMap_umap_trout   <- uwot::umap(devMap_in_mat_trout)
set.seed(atac_seed)
bodyMap_umap_salmon <- uwot::umap(bodyMap_in_mat_salmon)
set.seed(atac_seed)
bodyMap_umap_trout  <- uwot::umap(bodyMap_in_mat_trout)
```

## Persist SOM objects

```{r save_atac_som_objects}
save(devMap_atac_som_salmon,  devMap_atac_som_trout,
     bodyMap_atac_som_salmon, bodyMap_atac_som_trout,
     superclasses_devmap_salmon,  superclasses_devmap_trout,
     superclasses_bodymap_salmon, superclasses_bodymap_trout,
     sum_devmap_salmon,  sum_devmap_trout,
     sum_bodymap_salmon, sum_bodymap_trout,
     ribbon_devmap_salmon,  ribbon_devmap_trout,
     ribbon_bodymap_salmon, ribbon_bodymap_trout,
     devMap_umap_salmon,  devMap_umap_trout,
     bodyMap_umap_salmon, bodyMap_umap_trout,
     file = file.path(paths$legacy$derived_dir, "atac_soms.RData"))
```
