---
title: "03 — Figure 2B panels (chromatin accessibility SOMs)"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---


```{r knitr_setup, include = FALSE}
# Echo code, suppress warnings/messages in the rendered HTML, and cache the
# expensive heatmap chunks so re-knits are fast.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      cache = TRUE, cache.path = "cache/03_fig2B/")
```

## Purpose

This notebook builds the **chromatin** panels of the manuscript's Figure 2 —
i.e. **Figure 2B** (ATAC-seq): the four accessibility-SOM heatmaps and their
UMAPs (DevMap salmon, BodyMap salmon, DevMap trout, BodyMap trout). It is the
ATAC counterpart of the transcriptome notebook
`../transcriptome-analysis/03_fig2A_expression_soms.Rmd`. As with Figure 2A, the
final figure is assembled in Illustrator, so every panel is saved as an
individual vector **PDF** under `results/` (a combined composite PDF is also
written for quick reference).

It loads the SOM input matrices, the fitted SOM objects and the UMAP embeddings
produced by `01_atac_loading.Rmd` and `02_atac_soms.Rmd`.

```{r source_setup}
# Shared helper functions (palettes, heatmap/UMAP builders).
source("99_helpers.R")
# If `paths` is not already in the session, knit 00_setup.Rmd to define it.
if (!exists("paths")) knitr::knit("00_setup.Rmd", output = tempfile())
# Derived SOM input matrices + fitted SOMs + UMAP embeddings (from 01 and 02).
# Fail early with a clear message if those notebooks have not been run yet.
.need <- file.path(paths$derived_dir,
                   c("atac_matrices_salmon.RData", "atac_matrices_trout.RData",
                     "atac_soms.RData"))
if (!all(file.exists(.need))) {
  stop("Missing ", paste(basename(.need[!file.exists(.need)]), collapse = ", "),
       " in ", paths$derived_dir, ". Knit 01_atac_loading.Rmd and ",
       "02_atac_soms.Rmd first (run the notebooks in order 00 -> 01 -> 02 -> 03).")
}
for (f in .need) load(f)

# Shared heatmap colour scale: cividis ramp over the 0..2 normalised-intensity
# range used by every ATAC heatmap below.
pal_intensity <- circlize::colorRamp2(c(0:2), viridis::cividis(3)[0:3])

# Ensure the output directory exists, then define two small savers used
# throughout: one for ComplexHeatmap objects, one for ggplot UMAPs.
dir.create(paths$results_dir, showWarnings = FALSE, recursive = TRUE)
# save_hm_pdf — render a ComplexHeatmap to its own PDF.
save_hm_pdf <- function(hm, file, width = 4, height = 7) {
  pdf(file.path(paths$results_dir, file), width = width, height = height)
  ComplexHeatmap::draw(hm)
  dev.off()
}
# save_umap_pdf — save a ggplot UMAP to its own PDF.
save_umap_pdf <- function(p, file, width = 4, height = 4) {
  ggplot2::ggsave(file.path(paths$results_dir, file), p,
                  width = width, height = height)
}
```

# Salmon DevMap

## Palette

The DevMap palette uses inferno restricted to 0.285–0.627 across the 17 dynamic
SOM units, with the 8 constitutive units in grey. `splits_levels_*` defines the
order in which the dynamic units appear down the heatmap; the trailing
constitutive units come after.

```{r palette_devmap_atac_salmon}
# Dynamic-unit order (bottom-to-top) for the salmon DevMap heatmap.
splits_levels_devmap_salmon <- rev(c("25", "24", "14", "20",
                                     "10", "5", "4", "8",
                                     "2", "1", "6", "7",
                                     "11", "17", "16", "22", "21"))
# Full unit order = dynamic units followed by the 8 constitutive units.
splits_levels_c_salmon <- c(splits_levels_devmap_salmon,
                            c("3", "9", "13", "12", "15", "18", "19", "23"))

# Build the inferno+grey palette keyed by raw SOM unit id.
palette_devmap_salmon <- inferno_plus_grey_palette(
  n_dynamic      = 17,
  n_constitutive = 8,
  order_perm     = splits_levels_c_salmon
)
```

## Heatmap → PDF

```{r heatmap_devmap_salmon, cache = TRUE}
# Heatmap of normalised peak intensity for the salmon DevMap dynamic units.
heatmap_devmap_salmon <- make_atac_heatmap(
  in_mat        = devMap_in_mat_salmon,
  som           = devMap_atac_som_salmon,
  splits_levels = splits_levels_devmap_salmon,
  palette       = pal_intensity,
  show_column_names   = TRUE,
  show_heatmap_legend = FALSE
)
# Save this panel on its own.
save_hm_pdf(heatmap_devmap_salmon, "fig2B_heatmap_devmap_salmon.pdf")
```

## UMAP → PDF

Rendered with `flip_y = TRUE` so the colour gradient aligns with the heatmap row
order.

```{r umap_devmap_salmon}
# SOM-coloured UMAP of the salmon DevMap accessibility regions.
umap_devmap_salmon <- make_atac_umap(devMap_umap_salmon,
                                     devMap_atac_som_salmon,
                                     palette = palette_devmap_salmon,
                                     flip_y = TRUE)
save_umap_pdf(umap_devmap_salmon, "fig2B_umap_devmap_salmon.pdf")
```

# Salmon BodyMap

## Heatmap → PDF

The salmon BodyMap columns are reordered to group the gonad samples by
maturity-then-sex (`[,c(1:4,6,8,5,7,9:16)]`) and renamed so immature/mature
ovary and testis labels are explicit.

```{r heatmap_bodymap_salmon, cache = TRUE}
# Rename the four gonad columns to explicit ovary/testis labels.
colnames_bodymap_salmon <- colnames(bodyMap_in_mat_salmon)
colnames_bodymap_salmon[5] <- "Ovary_Inmature_Female_ATAC"
colnames_bodymap_salmon[6] <- "Testes_Inmature_Male_ATAC"
colnames_bodymap_salmon[7] <- "Ovary_Mature_Female_ATAC"
colnames_bodymap_salmon[8] <- "Testes_Mature_Male_ATAC"
colnames(bodyMap_in_mat_salmon) <- colnames_bodymap_salmon
# Reorder columns into the canonical tissue/maturity/sex order.
bodymap_in_mat_reorder_salmon <- bodyMap_in_mat_salmon[, c(1:4, 6, 8, 5, 7, 9:16)]

# Dynamic-unit order (bottom-to-top) for the salmon BodyMap heatmap.
splits_levels_bodymap_salmon <- rev(c("24", "25", "19", "13", "8", "3",
                                      "17", "5", "1", "6", "7", "23", "22"))

# Heatmap of normalised peak intensity for the salmon BodyMap dynamic units.
heatmap_bodymap_salmon <- make_atac_heatmap(
  in_mat        = bodymap_in_mat_reorder_salmon,
  som           = bodyMap_atac_som_salmon,
  splits_levels = splits_levels_bodymap_salmon,
  palette       = pal_intensity,
  show_column_names = TRUE
)
save_hm_pdf(heatmap_bodymap_salmon, "fig2B_heatmap_bodymap_salmon.pdf", 5, 7)
```

## Palette and UMAP → PDF

The BodyMap palette is hand-curated to reflect tissue identity (red for gonads,
blue for somatic non-neural tissues, etc.) and is already indexed by SOM unit,
so no reordering is needed for the UMAP.

```{r umap_bodymap_salmon}
# Hand-curated tissue palette for the salmon BodyMap.
palette_bodymap_salmon <- salmon_bodymap_palette()
# SOM-coloured UMAP of the salmon BodyMap accessibility regions.
umap_bodymap_salmon <- make_atac_umap(bodyMap_umap_salmon,
                                      bodyMap_atac_som_salmon,
                                      palette = palette_bodymap_salmon,
                                      flip_y = FALSE)
save_umap_pdf(umap_bodymap_salmon, "fig2B_umap_bodymap_salmon.pdf")
```

# Trout DevMap

## Palette and heatmap → PDF

```{r heatmap_devmap_trout, cache = TRUE}
# Dynamic-unit order (bottom-to-top) for the trout DevMap heatmap.
splits_levels_devmap_trout <- rev(c("25", "15", "14", "20", "24",
                                    "23", "18", "5",
                                    "3", "2",
                                    "19", "1",
                                    "7", "6", "21", "17", "16"))
# Full unit order = dynamic units followed by the 8 constitutive units.
splits_levels_c_trout <- c(splits_levels_devmap_trout,
                           c("8", "9", "10", "11", "12", "13", "22", "4"))

# Inferno+grey palette for the trout DevMap, keyed by raw SOM unit id.
palette_devmap_trout <- inferno_plus_grey_palette(
  n_dynamic      = 17,
  n_constitutive = 8,
  order_perm     = splits_levels_c_trout
)

# Heatmap of normalised peak intensity for the trout DevMap dynamic units
# (legend shown here; column names hidden as they repeat the salmon panel).
heatmap_devmap_trout <- make_atac_heatmap(
  in_mat        = devMap_in_mat_trout,
  som           = devMap_atac_som_trout,
  splits_levels = splits_levels_devmap_trout,
  palette       = pal_intensity,
  show_column_names = FALSE,
  show_heatmap_legend = TRUE
)
save_hm_pdf(heatmap_devmap_trout, "fig2B_heatmap_devmap_trout.pdf")
```

## UMAP → PDF

```{r umap_devmap_trout}
# SOM-coloured UMAP of the trout DevMap accessibility regions.
umap_devmap_trout <- make_atac_umap(devMap_umap_trout,
                                    devMap_atac_som_trout,
                                    palette = palette_devmap_trout,
                                    flip_y = TRUE)
save_umap_pdf(umap_devmap_trout, "fig2B_umap_devmap_trout.pdf")
```

# Trout BodyMap

## Heatmap → PDF

Trout BodyMap has 12 sample columns; we keep them in input order.

```{r heatmap_bodymap_trout, cache = TRUE}
# Keep trout BodyMap columns in input order (no gonad reshuffle needed).
bodymap_in_mat_reorder_trout <- bodyMap_in_mat_trout[, c(1:4, 5:8, 9:12)]

# Dynamic-unit order (bottom-to-top) for the trout BodyMap heatmap.
splits_levels_bodymap_trout <- rev(c("25", "20", "21", "16", "17", "2", "3"))

# Heatmap of normalised peak intensity for the trout BodyMap dynamic units.
heatmap_bodymap_trout <- make_atac_heatmap(
  in_mat        = bodymap_in_mat_reorder_trout,
  som           = bodyMap_atac_som_trout,
  splits_levels = splits_levels_bodymap_trout,
  palette       = pal_intensity,
  show_column_names = TRUE
)
save_hm_pdf(heatmap_bodymap_trout, "fig2B_heatmap_bodymap_trout.pdf", 5, 7)
```

## Palette and UMAP → PDF

```{r umap_bodymap_trout}
# Hand-curated tissue palette for the trout BodyMap.
palette_bodymap_trout <- trout_bodymap_palette()
# SOM-coloured UMAP of the trout BodyMap accessibility regions.
umap_bodymap_trout <- make_atac_umap(bodyMap_umap_trout,
                                     bodyMap_atac_som_trout,
                                     palette = palette_bodymap_trout,
                                     flip_y = FALSE)
save_umap_pdf(umap_bodymap_trout, "fig2B_umap_bodymap_trout.pdf")
```

# Composite assembly → PDF

For a quick-look reference, the four heatmaps are converted to grobs (so they
can sit in a `grid.arrange` layout) and stacked above a parallel strip of the
four UMAPs. The manuscript's final Fig 2B is assembled in Illustrator from the
individual PDFs above.

```{r assemble_atac_figure2}
# Convert each ComplexHeatmap into a grob so it can be arranged with the UMAPs.
heatmap_grob_devmap_salmon  <- grid.grabExpr(ComplexHeatmap::draw(heatmap_devmap_salmon))
heatmap_grob_bodymap_salmon <- grid.grabExpr(ComplexHeatmap::draw(heatmap_bodymap_salmon))
heatmap_grob_devmap_trout   <- grid.grabExpr(ComplexHeatmap::draw(heatmap_devmap_trout))
heatmap_grob_bodymap_trout  <- grid.grabExpr(ComplexHeatmap::draw(heatmap_bodymap_trout))

# Row of the four heatmaps.
atac_heatmaps <- gridExtra::arrangeGrob(
  heatmap_grob_devmap_salmon,  heatmap_grob_bodymap_salmon,
  heatmap_grob_devmap_trout,   heatmap_grob_bodymap_trout,
  nrow = 1
)
# Row of the four UMAPs.
atac_umaps <- gridExtra::arrangeGrob(
  umap_devmap_salmon,  umap_bodymap_salmon,
  umap_devmap_trout,   umap_bodymap_trout,
  nrow = 1
)

# Write the combined composite PDF (heatmap row over UMAP row).
pdf(file.path(paths$results_dir, "fig2B_atac_composite.pdf"),
    width = 16, height = 10)
gridExtra::grid.arrange(atac_heatmaps, atac_umaps, ncol = 1, heights = c(2, 1))
dev.off()
```
