---
title: "Figure2"
author: "Damir Baranasic"
format: html
editor: visual
---

## Introduction

Salmonids, a group of fish species that includes salmon, play a vital ecological role and are economically significant due to their widespread aquaculture, recreational fishing, and cultural importance. In recent years, functional genomics has emerged as a powerful approach to understanding the complex biological mechanisms underlying the adaptation, reproduction, and overall survival of these species. By unraveling the genomic architecture of salmon, researchers can explore fundamental aspects of their biology, potentially informing strategies for conservation, aquaculture improvement, and disease management.

ATAC-seq (Assay for Transposase-Accessible Chromatin using sequencing) is a valuable technique that enables the assessment of genome-wide chromatin accessibility, providing insights into the regulatory landscape of an organism. In this study, we have generated ATAC-seq data from four tissues (brain, liver, muscle, and gonads) in male and female salmon, encompassing mature and immature specimens. Additionally, we have obtained data from five stages of embryonic development, which will allow us to investigate the dynamic changes in chromatin accessibility during salmon ontogeny.

Our primary objective is to cluster and analyze this rich dataset to uncover developmental dynamics and tissue-specific sex regulatory differences. By examining the global chromatin accessibility patterns, we aim to identify key regulatory elements and factors involved in tissue-specific functions, sexual dimorphism, and developmental transitions. Ultimately, this research will enhance our understanding of salmon biology and contribute to the growing body of knowledge in functional genomics, with potential applications in fisheries management, aquaculture, and conservation efforts.

## SOM clustering

To analyze the obtained data, we will employ Self-Organizing Map (SOM) clustering, a powerful unsupervised machine learning technique that effectively organizes high-dimensional data into lower-dimensional grid structures while preserving topological relationships. For each candidate regulatory region, which is defined as the union of robust ATAC-seq peaks across all analyzed samples, we will count the number of reads within the region. Subsequently, we will calculate the fold change of the counts, normalized by transcripts per million (TPM), over the expected regions. This fold change measurement will provide insights into the relative accessibility of these regions across various tissues, developmental stages, and sexes. After scaling the data per sample to ensure comparability, we will perform SOM clustering to group the candidate regulatory regions based on their chromatin accessibility profiles. This approach will enable us to identify and investigate distinct patterns of chromatin accessibility, uncovering meaningful relationships that can shed light on the complex regulatory landscape of salmon.

```{r}
#| echo: true
#| warning: false
#| message: false
#| cache: true

source("R/utils.R")

the_rds <- "~/data/AQUA-FAANG/Salmon/BodyMap_DevMap_expression/RainbowTrout_consensus_expression.RDS"
ssal_consensus_expression <- read_rds(the_rds)

# blacklist_salmon <- import("https://salmobase.org/datafiles/datasets/Aqua-Faang/blacklist/AtlanticSalmon_blacklist.bed")

union_peaks <- ssal_consensus_expression %>% rename(c("chr" = "seqnames")) %>%
  mutate(start = as.numeric(start), end = as.numeric(end), strand = str_replace(strand, "\\.", "\\*")) %>% as_granges()

# filter_idx <- which(union_peaks %over% blacklist_salmon)
# 
# union_peaks <- union_peaks[-filter_idx]

# fe_ATAC <- ssal_consensus_expression %>% 
#   select(contains("ATAC")) %>%
#   mutate(across(everything(), ~ .x / mean(.x)))

bodyMap_idr_files <- list.files("~/data/AQUA-FAANG/Trout/BodyMap_ATAC_IDR/",
                                pattern = "\\.bed$", full.names = T) 
bodyMap_idr <- lapply(bodyMap_idr_files, function(x){
  read_tsv(x,
           col_names = c("seqnames", "start", "end", "name",
                         "score", "strand", "tickStart", "tickEnd", "itemRgb",
                         "no_idea_1", "no_idea_2", "no_idea_3", "Replicates",
                         "ChromHMM_annotation")) %>%
    makeGRangesFromDataFrame(keep.extra.columns = T,
                             starts.in.df.are.0based = T)
})
names(bodyMap_idr) <- str_remove(str_remove(basename(bodyMap_idr_files),
                                            "_consensus_peaks.bed"),
                                 "RainbowTrout_ATAC_")
bodyMap_union <- as(bodyMap_idr, Class = "GRangesList") %>% unlist() %>% reduce()


devMap_idr_files <- list.files(
  "~/data/AQUA-FAANG/Trout/Trout_DevMap/ATAC_consensus_IDR//",
                                pattern = "\\.bed$", full.names = T) 
devMap_idr <- lapply(devMap_idr_files, import)
names(devMap_idr) <- str_remove(basename(devMap_idr_files),  "_consensus_peaks.bed")
devMap_union <- as(devMap_idr, Class = "GRangesList") %>% unlist() %>% reduce()
```

You can add options to executable code like this

The `echo: false` option disables the printing of code (only output is displayed).

```{r}

devMap_idx <- which(union_peaks %over% devMap_union)
devMap_expression <- ssal_consensus_expression[devMap_idx, ] %>% 
  select(contains("ATAC")) %>% select(1:5)

devMap_fe_ATAC <- devMap_expression %>%
  mutate(across(everything(), ~ .x / mean(.x)))

devMap_in_mat <- devMap_fe_ATAC %>% 
  as.matrix() %>% t() %>% scale(center = F) %>% t()

# set.seed(12345)
# result <- som_clustering(in_mat, 5, 5)
# 
# output_plot <- result$plot
# output_plot

set.seed(12345)

devMap_init <- somInit(traindat = devMap_in_mat, 5, 5)
devMap_atac.som <- kohonen::som(devMap_in_mat, 
                         grid = kohonen::somgrid(5, 5, "hexagonal"), 
                         init = devMap_init)

  devMap_label_unit <- table(devMap_atac.som$unit.classif)
  devMap_tmp_name <- names(devMap_label_unit)
  devMap_label_unit <- str_c("Class ", devMap_tmp_name, " (", devMap_label_unit, ")")
  names(devMap_label_unit) <- devMap_tmp_name
  
  devMap_tmp_df <- devMap_in_mat %>% as.data.frame() %>%
    {cbind(class = devMap_atac.som$unit.classif, .)} %>%
    pivot_longer(cols = !class,
                 names_to = "stage",
                 values_to = "value") %>%
    mutate(stage = factor(stage,
                          levels = colnames(devMap_in_mat)),
           class = factor(class,
                          levels = str_sort(unique(class), numeric = T)))
  devMap_sum_stat <- devMap_tmp_df %>% group_by(class, stage) %>%
    summarize(value = mean(value)) %>% ungroup()
  
  # out_plot <- tmp_df%>%
  #   ggplot(aes(stage, value)) + geom_violin() +
  #   geom_point(data = sum_stat) +
  #   geom_line(data = sum_stat, aes(group = class)) +
  #   facet_wrap(~ class, ncol = 5, labeller = labeller(class = label_unit),
  #              scale = "free_y",as.table = F) +
  #   theme_bw() +
  #   theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
devMap_superclust_pam <- cluster::pam(devMap_atac.som$codes[[1]], 4)
devMap_superclasses_pam <- devMap_superclust_pam$clustering

devMap_tmp_df%>%
  ggplot(aes(stage, value, 
             colour = factor(devMap_superclasses_pam[class]),
             group = "Class")) +
  stat_summary(fun.data  = "mean_sdl",
               fun.args = list(mult = 2)) + 
  stat_summary(fun = mean, geom = "line") +
  facet_wrap(~ class, ncol = 5, 
             labeller = labeller(class = devMap_label_unit),
             as.table = F) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(colour = "cluster")


# devMap_dynamic_idx <- which(!(devMap_atac.som$unit.classif %in% 
#                          which(devMap_superclasses_pam == 2)))
# 
# devMap_umap_in <- devMap_in_mat[devMap_dynamic_idx, ]
# 
# devMap_row_classes <- data.frame(paste0("Super ",
#                               devMap_superclasses_pam[
#                                 devMap_atac.som$unit.classif[
#                                   devMap_dynamic_idx]]),
#                           paste0("SOM ",
#                                  devMap_atac.som$unit.classif[
#                                   devMap_dynamic_idx]))
library(ComplexHeatmap)

devMap_ht1 <- Heatmap(devMap_in_mat, cluster_columns = F, 
               row_split = devMap_atac.som$unit.classif, 
               cluster_row_slices = T, row_dend_reorder = T)

# devMap_class_index <- as.numeric(names(row_order(devMap_ht1)))

# library(scales)
# Heatmap(devMap_umap_in, cluster_columns = F, 
#         row_split = devMap_superclasses_pam[
#                                 devMap_atac.som$unit.classif[
#                                   devMap_dynamic_idx]], 
#         cluster_row_slices = T, row_dend_reorder = T, 
#         row_title_gp = gpar(col = hue_pal()(4)[
#           devMap_superclasses_pam[devMap_class_index]
#           ]))


devMap_umap_atac <- uwot::umap(devMap_in_mat)

devMap_umap_atac %>% as.data.frame %>% ggplot(aes(V1, V2)) + 
  geom_point(aes(colour = factor(devMap_superclasses_pam[
                                devMap_atac.som$unit.classif])), size = .2, alpha = .2) + 
  theme_classic() + labs(colour = "Class")

```

```{r}

bodyMap_idx <- which(union_peaks %over% bodyMap_union)
bodyMap_expression <- ssal_consensus_expression[bodyMap_idx, ] %>% 
  select(contains("ATAC")) %>% select(6:17)

bodyMap_fe_ATAC <- bodyMap_expression %>%
  mutate(across(everything(), ~ .x / mean(.x)))

bodyMap_in_mat <- bodyMap_fe_ATAC %>% 
  as.matrix() %>% t() %>% scale(center = F) %>% t()

# set.seed(12345)
# result <- som_clustering(in_mat, 5, 5)
# 
# output_plot <- result$plot
# output_plot

set.seed(12345)

init <- somInit(traindat = bodyMap_in_mat, 5, 5)
bodyMap_atac.som <- kohonen::som(bodyMap_in_mat, 
                         grid = kohonen::somgrid(5, 5, "hexagonal"), 
                         init = init)

  bodyMap_label_unit <- table(bodyMap_atac.som$unit.classif)
  bodyMap_tmp_name <- names(bodyMap_label_unit)
  bodyMap_label_unit <- str_c("Class ", bodyMap_tmp_name, " (", bodyMap_label_unit, ")")
  names(bodyMap_label_unit) <- bodyMap_tmp_name
  
  bodyMap_tmp_df <- bodyMap_in_mat %>% as.data.frame() %>%
    {cbind(class = bodyMap_atac.som$unit.classif, .)} %>%
    pivot_longer(cols = !class,
                 names_to = "stage",
                 values_to = "value") %>%
    mutate(stage = factor(stage,
                          levels = colnames(bodyMap_in_mat)),
           class = factor(class,
                          levels = str_sort(unique(class), numeric = T)))
  bodyMap_sum_stat <- bodyMap_tmp_df %>% group_by(class, stage) %>%
    summarize(value = mean(value)) %>% ungroup()
  
  # out_plot <- tmp_df%>%
  #   ggplot(aes(stage, value)) + geom_violin() +
  #   geom_point(data = sum_stat) +
  #   geom_line(data = sum_stat, aes(group = class)) +
  #   facet_wrap(~ class, ncol = 5, labeller = labeller(class = label_unit),
  #              scale = "free_y",as.table = F) +
  #   theme_bw() +
  #   theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
bodyMap_superclust_pam <- cluster::pam(bodyMap_atac.som$codes[[1]], 4)
bodyMap_superclasses_pam <- bodyMap_superclust_pam$clustering

bodyMap_tmp_df%>%
  ggplot(aes(stage, value, 
             colour = factor(bodyMap_superclasses_pam[class]),
             group = "Class")) +
  stat_summary(fun.data  = "mean_sdl",
               fun.args = list(mult = 2)) + 
  stat_summary(fun = mean, geom = "line") +
  facet_wrap(~ class, ncol = 5, 
             labeller = labeller(class = bodyMap_label_unit),
             as.table = F) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(colour = "cluster")


# bodyMap_dynamic_idx <- which(!(atac.som$unit.classif == 14))
# 
# bodyMap_umap_in <- bodyMap_in_mat[bodyMap_dynamic_idx, ]
# 
# bodyMap_row_classes <- data.frame(paste0("Super ",
#                               bodyMap_superclasses_pam[
#                                 bodyMap_atac.som$unit.classif[
#                                   bodyMap_dynamic_idx]]),
#                           paste0("SOM ",
#                                  bodyMap_atac.som$unit.classif[
#                                   bodyMap_dynamic_idx]))
library(ComplexHeatmap)

bodyMap_ht1 <- Heatmap(bodyMap_in_mat, cluster_columns = F,
               row_split = bodyMap_atac.som$unit.classif, 
               cluster_row_slices = T, row_dend_reorder = T)

# bodyMap_class_index <- as.numeric(names(row_order(bodyMap_ht1)))

# library(scales)
# Heatmap(bodyMap_umap_in, cluster_columns = F, 
#         row_split = bodyMap_superclasses_pam[
#                                 bodyMap_atac.som$unit.classif[
#                                   bodyMap_dynamic_idx]], 
#         cluster_row_slices = T, row_dend_reorder = T, 
#         row_title_gp = gpar(col = hue_pal()(4)[
#           bodyMap_superclasses_pam[bodyMap_class_index]
#           ]))


bodyMap_umap_atac <- uwot::umap(bodyMap_in_mat)

bodyMap_umap_atac %>% as.data.frame %>% ggplot(aes(V1, V2)) + 
  geom_point(aes(colour = factor(bodyMap_superclasses_pam[
                                bodyMap_atac.som$unit.classif])), size = .2, alpha = .2) + 
  theme_classic() + labs(colour = "Class")

```
