---
title: "00 — Setup"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
---

## What this notebook does

`00_setup.Rmd` is the entry point for the transcriptome analysis. Run it first,
top to bottom: it (1) records which data source you want, (2) sources
`R/setup.R` (which installs/loads packages, defines constants and the `paths`
list, and creates output directories), and (3) prints the resulting
configuration so you can confirm everything is set up correctly. Then run
`01..06` in order, in the same R session.

## Choose a data source

There are two modes. **You do not have to change anything** — the default
(`reproduce`) is already set in the chunk below and is what most users want.

- **`reproduce`** (default) — Use the small derived assets shipped in this repo
  (the normalised expression matrices in `data/norm/` and the SOM cluster
  assignments in `data/soms/`). These are the exact objects behind the
  manuscript, so the notebooks reproduce the published figures. Fast, and the
  only thing downloaded is the orthogroup table.
- **`recompute`** — Re-derive everything from the raw RNA-seq TPM matrices,
  which are downloaded from Salmobase, and re-fit the SOMs from scratch. Use
  this to replicate the analysis from raw data. Note: because SOM fitting is not
  bit-stable across computers, the clustering — and therefore the figures — can
  differ slightly from the manuscript depending on the system used.

The chunk below sets the mode. To use `recompute` instead, comment the first
`options()` line and uncomment the second. `aquafaang.test` controls run length
(see comment).

```{r choose_mode}
# Default: reproduce the published figures from the deposited derived data.
options(aquafaang.mode = "reproduce")
# Alternative: re-derive everything from raw RNA-seq (Salmobase). To use it,
# comment the line above and uncomment the line below.
# options(aquafaang.mode = "recompute")

# Run length: FALSE = full manuscript run; TRUE = fast smoke test (fewer
# permutations/bootstraps in 05). Leave FALSE unless you just want to check the
# pipeline runs end to end.
options(aquafaang.test = FALSE)
```

## Run setup

Everything the downstream notebooks need — the package stack, the helper
functions, the project constants, and the `paths` list — is defined in
`R/setup.R`. Sourcing it reads the options set above and configures the
session accordingly. `sessionInfo()` then records the exact package versions
for provenance.

```{r run_setup}
# Configure the session from the options chosen above (packages, paths, ...).
source("R/setup.R")
# Print R version + attached package versions, for reproducibility records.
sessionInfo()
```

## Configuration in scope

This chunk is a sanity check: it prints the configuration that `R/setup.R` just
resolved, so you can confirm which mode you are in before running `01..06`.
`DATA_MODE` and `TEST_MODE` echo your choices; `N_PERMUTATIONS` is the permutation
count notebook 05 will use; and `str(paths$active, ...)` lists the active data
sources (local directories in reproduce mode, Salmobase URLs in recompute mode).

```{r show_config}
# Which data source is active ("reproduce" or "recompute").
cat("DATA_MODE = ", DATA_MODE, "\n", sep = "")
# Whether the fast smoke-test shortcuts are on.
cat("TEST_MODE = ", TEST_MODE, "\n", sep = "")
# Permutation count notebook 05 will run with (100 in test mode, else 10,000).
cat("N_PERMUTATIONS = ", N_PERMUTATIONS, "\n", sep = "")
# The active set of input locations (top level only, to keep the output short).
str(paths$active, max.level = 1)
```
