---
title: "01 — Data loading and normalisation"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---

## Overview

This notebook produces the four quantile-normalised, z-scaled expression
matrices — `sd_norm` (salmon DevMap), `td_norm` (trout DevMap), `sb_norm`
(salmon BodyMap), `tb_norm` (trout BodyMap) — that feed the SOM fits in
`02_soms.Rmd`. What it actually does depends on `aquafaang.mode` (see
`00_setup.Rmd`):

- **`reproduce` mode** (default): load the deposited matrices directly from
  `data/norm/*.tsv.gz` and skip RNA-seq loading + normalisation entirely. These
  matrices are exactly what produced the manuscript figures, so everything
  downstream reproduces bit-for-bit.
- **`recompute` mode**: download the per-tissue / per-stage TPM matrices from
  Salmobase, merge technical replicates, and re-run `normalize()` from scratch.
  A sibling `data/norm/*_regen.tsv.gz` set is written so the deposited matrices
  are never overwritten.

The four matrices are left in the session for `02..06` to use (the README's
intended "run `01..06` in order in one session" workflow).

```{r setup_chain, include=FALSE}
# Packages, constants, paths, helpers and the deposited-asset loaders.
source("R/setup.R")
```

## Load the orthogroup table

The salmon × trout sub-orthogroup table is the join key for every cross-species
analysis (observed/expected, ohnolog co-clustering, fold-change patterns). It
is always fetched from Salmobase — never shipped in the repo —
and `read_table_anywhere()` caches the download under `cache/` so repeat runs
do not re-hit the network.

```{r load_orthogroups}
# Download (once) and read the orthogroup table into `orthogroups`.
orthogroups <- read_table_anywhere(paths$active$orthogroups)
```

## Load (reproduce) or re-derive (recompute) the normalised matrices

The single `if` below is the fork between the two modes. In `reproduce` mode it
is one call to the shared loader; in `recompute` mode it runs the full
download → merge replicates → normalise → persist pipeline.

```{r normalize_expression}
if (DATA_MODE == "reproduce") {
  # Inject sd_norm / td_norm / sb_norm / tb_norm straight from data/norm/.
  load_deposited_norm()
} else {
  # --- recompute: pull raw TPM from Salmobase --------------------------------
  # DevMap: one merged TPM matrix per species (gene_id, transcript_id(s), reps).
  sd_raw <- read_table_anywhere(paths$active$devmap_salmon)
  td_raw <- read_table_anywhere(paths$active$devmap_trout)
  # BodyMap: seven per-tissue TPM files concatenated into one wide matrix.
  sb_raw <- process_tissue_data("salmon", urls = paths$active$bodymap_salmon_urls)
  tb_raw <- process_tissue_data("trout",  urls = paths$active$bodymap_trout_urls)

  # --- collapse technical replicates to one column per condition -------------
  sd_merged <- merge_reps_devmap(sd_raw)
  td_merged <- merge_reps_devmap(td_raw)
  sb_merged <- merge_reps_bodymap(sb_raw)
  tb_merged <- merge_reps_bodymap(tb_raw)

  # --- quantile-normalise + z-scale each matrix ------------------------------
  sd_norm <- normalize(sd_merged)
  td_norm <- normalize(td_merged)
  sb_norm <- normalize(sb_merged)
  tb_norm <- normalize(tb_merged)

  # --- persist a *_regen.tsv.gz set (never overwrite the deposited files) ----
  out_dir <- "data/norm"
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
  for (tag in c("sd", "td", "sb", "tb")) {
    m  <- get(paste0(tag, "_norm"))                                   # this matrix
    df <- tibble::rownames_to_column(as.data.frame(m), var = "gene_id")  # gene_id col
    readr::write_tsv(df, file.path(out_dir, paste0(tag, "_norm_regen.tsv.gz")))
  }
}
```
