---
title: "01 — Data loading and normalisation"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---

## Overview

Loads the DevMap and BodyMap RNA-seq TPM matrices for Atlantic salmon and
rainbow trout, merges technical replicates, and produces the
quantile-normalised, z-scaled matrices that feed the SOM fits in
`02_soms.Rmd`. A trailing chunk re-saves the four normalised matrices to
`data/original_norm_files/norm.RData` so the downstream notebooks can
skip the recomputation.

```{r setup_chain, include=FALSE}
source("R/setup.R")
```

## Load the orthogroup table

The salmon × trout sub-orthogroup table is the join key for every
cross-species analysis (observed / expected, ohnolog co-clustering,
fold-change patterns). `read_table_anywhere()` accepts either the local
path (`legacy` mode) or the Salmobase URL (`salmobase` mode) and caches
the URL download under `cache/`.

```{r load_orthogroups}
orthogroups <- read_table_anywhere(paths$active$orthogroups)
```

## Load DevMap TPM matrices

DevMap RNA covers 14 developmental stages × technical replicates for both
species. The TSV layout (gene_id, transcript_id(s), then one column per
sample) is identical between the legacy local copy and the Salmobase mirror.

```{r load_devmap_tpm}
sd_raw <- read_table_anywhere(paths$active$devmap_salmon)
td_raw <- read_table_anywhere(paths$active$devmap_trout)
```

## Load BodyMap TPM matrices

BodyMap RNA is published per tissue (one TSV per tissue containing all
maturity × sex × replicate columns for that tissue). `process_tissue_data()`
concatenates the seven per-tissue files into a single gene × sample matrix
in the order that `merge_reps_bodymap()` expects.

```{r load_bodymap_tpm}
if (DATA_MODE == "salmobase") {
  sb_raw <- process_tissue_data("salmon",
                                urls = paths$active$bodymap_salmon_urls)
  tb_raw <- process_tissue_data("trout",
                                urls = paths$active$bodymap_trout_urls)
} else {
  sb_raw <- process_tissue_data("salmon",
                                file_path = paths$active$bodymap_salmon_dir)
  tb_raw <- process_tissue_data("trout",
                                file_path = paths$active$bodymap_trout_dir)
}
```

## Merge technical replicates

`merge_reps_devmap()` collapses the per-stage replicate columns into one
mean-TPM column per stage; `merge_reps_bodymap()` collapses the
Tissue × Maturity × Sex × Replicate naming scheme into one mean-TPM column
per (Tissue × Maturity × Sex) sample and renames Gonad → Ova / Testes by
the sex field.

```{r merge_replicates}
sd_merged <- merge_reps_devmap(sd_raw)
td_merged <- merge_reps_devmap(td_raw)

sb_merged <- merge_reps_bodymap(sb_raw)
tb_merged <- merge_reps_bodymap(tb_raw)
```

## Normalise expression

`normalize()` drops rows with no sample above 1 TPM, quantile-normalises
across samples and z-scales each gene. In `legacy` mode the chunk loads
the deposited matrices; in `salmobase` mode it redoes the normalisation
from scratch so that the resulting matrices reflect the Salmobase input.

```{r normalize_expression}
if (DATA_MODE == "legacy" && file.exists(paths$active$norm)) {
  load(paths$active$norm)
} else {
  sd_norm <- normalize(sd_merged)
  td_norm <- normalize(td_merged)
  sb_norm <- normalize(sb_merged)
  tb_norm <- normalize(tb_merged)
}
```

## Persist normalised matrices

In `salmobase` mode (where the matrices were just re-derived) the chunk
writes a sibling `norm_regen.RData` so the legacy file is not overwritten.

```{r save_norm_rdata}
if (DATA_MODE == "salmobase") {
  out <- sub("\\.RData$", "_regen.RData", paths$legacy$norm)
  save(sd_norm, td_norm, sb_norm, tb_norm, file = out)
  message("Wrote ", out)
}
```
