---
title: "01 — Data loading and normalisation"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---

## Overview

Produces the four quantile-normalised, z-scaled expression matrices
(`sd_norm`, `td_norm`, `sb_norm`, `tb_norm`) that feed the SOM fits in
`02_soms.Rmd`.

- **`legacy` mode** loads the deposited matrices directly from
  `data/norm/*.tsv.gz` and skips the RNA-seq loading + normalisation
  entirely — the deposited matrices are exactly what produced the
  manuscript figures, so the downstream notebooks are reproduced
  bit-for-bit.
- **`salmobase` mode** fetches the per-tissue / per-stage TPM matrices
  from Salmobase, merges technical replicates, and re-runs
  `normalize()`. A sibling `data/norm/*_regen.tsv.gz` set is written so
  the deposited matrices are not overwritten.

```{r setup_chain, include=FALSE}
source("R/setup.R")
```

## Load the orthogroup table

The salmon × trout sub-orthogroup table is the join key for every
cross-species analysis (observed / expected, ohnolog co-clustering,
fold-change patterns). It is always fetched from Salmobase;
`read_table_anywhere()` caches the download under `cache/` so subsequent
runs do not re-hit the network.

```{r load_orthogroups}
orthogroups <- read_table_anywhere(paths$active$orthogroups)
```

## Load (legacy) or re-derive (salmobase) the normalised matrices

```{r normalize_expression}
if (DATA_MODE == "legacy") {
  read_norm <- function(tag) {
    path <- file.path(paths$active$norm_dir, paste0(tag, "_norm.tsv.gz"))
    df <- readr::read_tsv(path, show_col_types = FALSE)
    m  <- as.matrix(df[, -1L])
    rownames(m) <- df$gene_id
    m
  }
  sd_norm <- read_norm("sd")
  td_norm <- read_norm("td")
  sb_norm <- read_norm("sb")
  tb_norm <- read_norm("tb")
} else {
  sd_raw <- read_table_anywhere(paths$active$devmap_salmon)
  td_raw <- read_table_anywhere(paths$active$devmap_trout)
  sb_raw <- process_tissue_data("salmon",
                                urls = paths$active$bodymap_salmon_urls)
  tb_raw <- process_tissue_data("trout",
                                urls = paths$active$bodymap_trout_urls)

  sd_merged <- merge_reps_devmap(sd_raw)
  td_merged <- merge_reps_devmap(td_raw)
  sb_merged <- merge_reps_bodymap(sb_raw)
  tb_merged <- merge_reps_bodymap(tb_raw)

  sd_norm <- normalize(sd_merged)
  td_norm <- normalize(td_merged)
  sb_norm <- normalize(sb_merged)
  tb_norm <- normalize(tb_merged)

  # Persist a sibling *_regen.tsv.gz set so the deposited matrices stay
  # untouched.
  out_dir <- "data/norm"
  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
  for (tag in c("sd", "td", "sb", "tb")) {
    m <- get(paste0(tag, "_norm"))
    df <- tibble::rownames_to_column(as.data.frame(m), var = "gene_id")
    readr::write_tsv(df, file.path(out_dir, paste0(tag, "_norm_regen.tsv.gz")))
  }
}
```
