---
title: "02 — Self-organising maps"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---

## Overview

This notebook provides the four self-organising maps (SOMs) used throughout
the paper: a 4×4 DevMap SOM and a 6×6 BodyMap SOM for **each** species, all on
a hexagonal grid.

**Important — this notebook behaves differently depending on `aquafaang.mode`
(set in `00_setup.Rmd`), and it does *not* always compute the SOMs from
scratch:**

- In **`reproduce`** mode (the default) the SOM fitting chunks are **skipped**
  and the deposited SOM unit assignments under `data/soms/` are loaded instead.
  This is what reproduces the manuscript figures exactly — the published SOMs
  are loaded, not recomputed.
- In **`recompute`** mode the SOMs are **re-fit from scratch** with
  `kohonen::som()` on the normalised matrices from `01_data_loading.Rmd`.
  Because `kohonen::som()` is not bit-stable across BLAS/OS builds, the unit
  numbering — and therefore the clustering and figures — can differ slightly
  from the manuscript.

Which path runs is controlled by the single flag `SOM_LOAD_DEPOSITED` (defined
in `R/setup.R`: `TRUE` in reproduce mode, `FALSE` in recompute mode). Every
chunk below is gated on it, so exactly one of the two paths executes.

This notebook expects `01_data_loading.Rmd` to have been run first **in the
same R session** (the README's intended workflow), so the normalised matrices
`sd_norm` / `td_norm` / `sb_norm` / `tb_norm` are already in scope.

```{r setup_chain, include=FALSE}
# Load packages, constants, paths and the deposited-asset loaders.
source("R/setup.R")
```

## Seed

```{r seed_lock}
# Fixed seed so a recompute-mode re-fit is at least reproducible on a given
# machine (it still differs across BLAS builds; see Overview).
som_seed <- 387334
```

## Fit DevMap SOMs — recompute mode only

The next chunk has `eval = !SOM_LOAD_DEPOSITED`, so it runs **only in recompute
mode**. In reproduce mode it is skipped entirely and the deposited assignments
are loaded further down ("Load deposited SOM objects").

```{r fit_devmap_soms, eval=!SOM_LOAD_DEPOSITED, cache=TRUE}
# Fit the salmon DevMap SOM (4x4 grid) on its normalised matrix.
set.seed(som_seed)
sd_som <- get_som(sd_norm, 4, 4)
# Fit the trout DevMap SOM (4x4 grid).
set.seed(som_seed)
td_som <- get_som(td_norm, 4, 4)
```

## Fit BodyMap SOMs — recompute mode only

Also recompute-only. The BodyMap fits take the column-reordered input so the
unit assignments line up with the tissue × maturity × sex ordering used in
every BodyMap figure.

```{r fit_bodymap_soms, eval=!SOM_LOAD_DEPOSITED, cache=TRUE}
# Fit the salmon BodyMap SOM (6x6 grid) on the column-reordered matrix.
set.seed(som_seed)
sb_som <- get_som_bodymap(sb_norm, 6, 6, order_indices = indices_order_ssal)
# Fit the trout BodyMap SOM (6x6 grid).
set.seed(som_seed)
tb_som <- get_som_bodymap(tb_norm, 6, 6, order_indices = indices_order_omyk)
```

## Load deposited SOM objects — reproduce mode only

This chunk has `eval = SOM_LOAD_DEPOSITED`, so it runs **only in reproduce
mode** (and is skipped in recompute mode, where the chunks above already
produced the SOMs). It reconstructs `sd_som` / `td_som` / `sb_som` / `tb_som`
from the deposited per-gene unit assignments and per-unit summary tables under
`data/soms/`, via the shared `load_deposited_soms()` helper (also used by the
ohnolog scripts, so the two can never drift). The reconstructed objects carry
only the fields the rest of the pipeline reads — `$som$unit.classif` and
`$stat` — not the full kohonen prototype codes / distance matrices, which are
not needed downstream.

```{r load_deposited_soms, eval=SOM_LOAD_DEPOSITED}
# Inject sd_som / td_som / sb_som / tb_som from data/soms/ into this session.
load_deposited_soms()

# Re-level the salmon BodyMap stat table's stage factor into the canonical
# tissue grouping; without this the BodyMap heatmaps/bar panels would render
# with their columns out of order.
lev <- levels(sb_som$stat$stage)
lev_order <- lev[c(3:6, 7:8, 1:2, 9:28)]          # hand-picked canonical order
sb_som$stat$stage <- factor(sb_som$stat$stage, levels = lev_order)
sb_som$stat <- arrange(sb_som$stat, class, stage) # sort rows to match

# Same re-levelling for the trout BodyMap stat table (its own column order).
lev <- levels(tb_som$stat$stage)
lev_order <- lev[c(3:6, 7, 1:2, 8:28)]
tb_som$stat$stage <- factor(tb_som$stat$stage, levels = lev_order)
tb_som$stat <- arrange(tb_som$stat, class, stage)
```

## Per-cluster annotation tables

Runs in both modes (the SOM objects exist either way by this point).
`som_ids()` returns the gene IDs that landed in each SOM unit;
`som_file_info_d()` / `som_file_info_b()` add the per-cluster annotation (peak
stage / tissue, peak mean expression, constitutive flag, and a stage-specific
flag for BodyMap) consumed by the heatmap, UMAP and ohnolog notebooks.

```{r som_info_tables}
# DevMap: gene lists per unit + the per-unit annotation table, per species.
sd_ids <- som_ids(sd_norm, sd_som)
sd_som_info <- som_file_info_d(sd_som)
td_ids <- som_ids(td_norm, td_som)
td_som_info <- som_file_info_d(td_som)

# BodyMap: same, using the BodyMap-specific annotation builder.
sb_ids <- som_ids(sb_norm, sb_som)
sb_som_info <- som_file_info_b(sb_som)
tb_ids <- som_ids(tb_norm, tb_som)
tb_som_info <- som_file_info_b(tb_som)
```

## Persist re-fit SOMs — recompute mode only

When the SOMs were just re-fit (recompute mode), write the per-gene unit
assignments and per-unit stat tables to a sibling `data/soms/*_regen.tsv.gz`
set, so the deposited files stay intact. The full kohonen objects remain
available in the live session for anyone who wants them.

```{r save_som_objects, eval=!SOM_LOAD_DEPOSITED}
# Ensure the output directory exists.
out_dir <- "data/soms"
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

# For each SOM, write the gene->unit map, the per-unit stat table, and the
# stage factor order, mirroring the deposited file format with a _regen suffix.
for (tag in c("sd", "td", "sb", "tb")) {
  som_obj <- get(paste0(tag, "_som"))     # the re-fit SOM object
  norm_m  <- get(paste0(tag, "_norm"))    # its input matrix (for gene order)
  # gene_id -> unit assignment, in the matrix's row order.
  readr::write_tsv(
    tibble::tibble(gene_id = rownames(norm_m),
                   unit    = as.integer(som_obj$som$unit.classif)),
    file.path(out_dir, paste0(tag, "_unit_classif_regen.tsv.gz")))
  # Per-unit summary table (factors flattened to plain text for portability).
  st <- as.data.frame(som_obj$stat)
  st$class <- as.character(st$class); st$stage <- as.character(st$stage)
  readr::write_tsv(st, file.path(out_dir, paste0(tag, "_unit_stats_regen.tsv.gz")))
  # The stage factor order, one level per line.
  writeLines(levels(som_obj$stat$stage),
             file.path(out_dir, paste0(tag, "_stage_levels_regen.txt")))
}
```
