---
title: "02 — Self-organising maps"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---

## Overview

Provides the four self-organising maps used throughout the paper: a 4×4
DevMap SOM and a 6×6 BodyMap SOM for each species, both on a hexagonal
grid. In `legacy` mode the SOMs are loaded from
`data/original_soms/soms.RData` (the canonical objects that produced the
manuscript figures). In `salmobase` mode they are re-fit from the
normalised matrices produced in `01_data_loading.Rmd`; unit assignments
will differ from the manuscript because `kohonen::som` is not bit-stable
across BLAS builds.

```{r setup_chain, include=FALSE}
source("R/setup.R")
```

## Seed

```{r seed_lock}
som_seed <- 387334
```

## Fit DevMap SOMs

```{r fit_devmap_soms, eval=!SOM_LOAD_LEGACY, cache=TRUE}
set.seed(som_seed)
sd_som <- get_som(sd_norm, 4, 4)
set.seed(som_seed)
td_som <- get_som(td_norm, 4, 4)
```

## Fit BodyMap SOMs

The BodyMap fits take the column-reordered input so the unit assignments
line up with the tissue × maturity × sex ordering used in every BodyMap
figure.

```{r fit_bodymap_soms, eval=!SOM_LOAD_LEGACY, cache=TRUE}
set.seed(som_seed)
sb_som <- get_som_bodymap(sb_norm, 6, 6, order_indices = indices_order_ssal)
set.seed(som_seed)
tb_som <- get_som_bodymap(tb_norm, 6, 6, order_indices = indices_order_omyk)
```

## Load deposited SOM objects (legacy mode only)

`soms.RData` carries `sd_som`, `td_som`, `sb_som`, `tb_som`. The BodyMap
stat tables are then reordered so the stage factor reflects the canonical
tissue grouping; without this re-levelling the BodyMap heatmaps and bar
panels render with the columns out of order.

```{r load_legacy_soms, eval=SOM_LOAD_LEGACY}
load(paths$legacy$soms)

lev <- levels(sb_som$stat$stage)
lev_order <- lev[c(3:6, 7:8, 1:2, 9:28)]
sb_som$stat$stage <- factor(sb_som$stat$stage, levels = lev_order)
sb_som$stat <- arrange(sb_som$stat, class, stage)

lev <- levels(tb_som$stat$stage)
lev_order <- lev[c(3:6, 7, 1:2, 8:28)]
tb_som$stat$stage <- factor(tb_som$stat$stage, levels = lev_order)
tb_som$stat <- arrange(tb_som$stat, class, stage)
```

## Per-cluster annotation tables

`som_ids()` returns the gene IDs that landed in each SOM unit.
`som_file_info_d()` and `som_file_info_b()` add the per-cluster
annotation (peak stage / tissue, peak mean expression, constitutive flag,
stage-specific flag for BodyMap) consumed by the heatmap, UMAP, and
ohnolog notebooks.

```{r som_info_tables}
sd_ids <- som_ids(sd_norm, sd_som)
sd_som_info <- som_file_info_d(sd_som)
td_ids <- som_ids(td_norm, td_som)
td_som_info <- som_file_info_d(td_som)

sb_ids <- som_ids(sb_norm, sb_som)
sb_som_info <- som_file_info_b(sb_som)
tb_ids <- som_ids(tb_norm, tb_som)
tb_som_info <- som_file_info_b(tb_som)
```

## Persist SOMs (salmobase mode)

When the SOMs were just re-fit, write them to a sibling `soms_regen.RData`
so the legacy file stays intact.

```{r save_som_objects, eval=!SOM_LOAD_LEGACY}
save(sd_som, td_som, sb_som, tb_som,
     file = sub("\\.RData$", "_regen.RData", paths$legacy$soms))
```
