---
title: "02 — Self-organising maps"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---

## Overview

Provides the four self-organising maps used throughout the paper: a 4×4
DevMap SOM and a 6×6 BodyMap SOM for each species, both on a hexagonal
grid. In `legacy` mode the SOMs are reconstructed from the deposited
per-gene unit assignments and per-unit summary tables under `data/soms/`
— only the fields actually consumed downstream (`som$unit.classif` and
`stat`) are reified, which is enough for every notebook in this tree.
In `salmobase` mode the SOMs are re-fit from the normalised matrices
produced in `01_data_loading.Rmd`; unit assignments will differ from the
manuscript because `kohonen::som` is not bit-stable across BLAS builds.

```{r setup_chain, include=FALSE}
source("R/setup.R")
```

## Seed

```{r seed_lock}
som_seed <- 387334
```

## Fit DevMap SOMs

```{r fit_devmap_soms, eval=!SOM_LOAD_LEGACY, cache=TRUE}
set.seed(som_seed)
sd_som <- get_som(sd_norm, 4, 4)
set.seed(som_seed)
td_som <- get_som(td_norm, 4, 4)
```

## Fit BodyMap SOMs

The BodyMap fits take the column-reordered input so the unit assignments
line up with the tissue × maturity × sex ordering used in every BodyMap
figure.

```{r fit_bodymap_soms, eval=!SOM_LOAD_LEGACY, cache=TRUE}
set.seed(som_seed)
sb_som <- get_som_bodymap(sb_norm, 6, 6, order_indices = indices_order_ssal)
set.seed(som_seed)
tb_som <- get_som_bodymap(tb_norm, 6, 6, order_indices = indices_order_omyk)
```

## Load deposited SOM objects (legacy mode only)

Reconstructs `sd_som`, `td_som`, `sb_som`, `tb_som` from the deposited
per-gene unit assignments (`{tag}_unit_classif.tsv.gz`) and per-unit
summary tables (`{tag}_unit_stats.tsv.gz` + `{tag}_stage_levels.txt`).
The reconstructed objects carry only the fields the rest of the
pipeline reads (`$som$unit.classif` and `$stat`); the kohonen prototype
codes, distance matrices, etc. are not needed downstream and are
omitted. The BodyMap stat tables are then re-levelled so the stage
factor reflects the canonical tissue grouping; without this the
BodyMap heatmaps and bar panels render with the columns out of order.

```{r load_legacy_soms, eval=SOM_LOAD_LEGACY}
load_legacy_som <- function(tag) {
  d <- paths$legacy$soms_dir
  uc <- readr::read_tsv(file.path(d, paste0(tag, "_unit_classif.tsv.gz")),
                        show_col_types = FALSE)
  st <- readr::read_tsv(file.path(d, paste0(tag, "_unit_stats.tsv.gz")),
                        show_col_types = FALSE)
  stage_levels <- readLines(file.path(d, paste0(tag, "_stage_levels.txt")))
  class_levels <- stringr::str_sort(unique(as.character(st$class)),
                                    numeric = TRUE)
  st$class <- factor(st$class, levels = class_levels)
  st$stage <- factor(st$stage, levels = stage_levels)
  list(som = list(unit.classif = as.integer(uc$unit)),
       stat = tibble::as_tibble(st))
}

sd_som <- load_legacy_som("sd")
td_som <- load_legacy_som("td")
sb_som <- load_legacy_som("sb")
tb_som <- load_legacy_som("tb")

lev <- levels(sb_som$stat$stage)
lev_order <- lev[c(3:6, 7:8, 1:2, 9:28)]
sb_som$stat$stage <- factor(sb_som$stat$stage, levels = lev_order)
sb_som$stat <- arrange(sb_som$stat, class, stage)

lev <- levels(tb_som$stat$stage)
lev_order <- lev[c(3:6, 7, 1:2, 8:28)]
tb_som$stat$stage <- factor(tb_som$stat$stage, levels = lev_order)
tb_som$stat <- arrange(tb_som$stat, class, stage)
```

## Per-cluster annotation tables

`som_ids()` returns the gene IDs that landed in each SOM unit.
`som_file_info_d()` and `som_file_info_b()` add the per-cluster
annotation (peak stage / tissue, peak mean expression, constitutive flag,
stage-specific flag for BodyMap) consumed by the heatmap, UMAP, and
ohnolog notebooks.

```{r som_info_tables}
sd_ids <- som_ids(sd_norm, sd_som)
sd_som_info <- som_file_info_d(sd_som)
td_ids <- som_ids(td_norm, td_som)
td_som_info <- som_file_info_d(td_som)

sb_ids <- som_ids(sb_norm, sb_som)
sb_som_info <- som_file_info_b(sb_som)
tb_ids <- som_ids(tb_norm, tb_som)
tb_som_info <- som_file_info_b(tb_som)
```

## Persist SOMs (salmobase mode)

When the SOMs were just re-fit, write the per-gene unit assignments and
per-unit stat tables to a sibling `data/soms/*_regen.tsv.gz` set so the
deposited files stay intact. Anyone who wants the full kohonen objects
in this mode can grab them from `globalenv()` while the session is live.

```{r save_som_objects, eval=!SOM_LOAD_LEGACY}
out_dir <- "data/soms"
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
for (tag in c("sd", "td", "sb", "tb")) {
  som_obj <- get(paste0(tag, "_som"))
  norm_m  <- get(paste0(tag, "_norm"))
  readr::write_tsv(
    tibble::tibble(gene_id = rownames(norm_m),
                   unit    = as.integer(som_obj$som$unit.classif)),
    file.path(out_dir, paste0(tag, "_unit_classif_regen.tsv.gz")))
  st <- as.data.frame(som_obj$stat)
  st$class <- as.character(st$class); st$stage <- as.character(st$stage)
  readr::write_tsv(st, file.path(out_dir, paste0(tag, "_unit_stats_regen.tsv.gz")))
  writeLines(levels(som_obj$stat$stage),
             file.path(out_dir, paste0(tag, "_stage_levels_regen.txt")))
}
```
