---
title: "03 — Figure 2A panels (transcriptome expression SOMs)"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---

## Overview

This notebook produces the **transcriptome** panels of the manuscript's
Figure 2 — i.e. **Figure 2A** (RNA-Seq) — plus their supplementary full-SOM
counterparts. It does **not** produce the whole of Figure 2: the final figure
is assembled in Illustrator from several sources, and the other pieces come
from elsewhere in the project:

| Figure 2 piece | Produced by |
|---|---|
| 2A heatmaps + UMAPs (RNA expression SOMs, DevMap) | **this notebook** |
| 2A light-red observed/expected ratio bands (RNA) | `05_observed_expected.Rmd` |
| 2B heatmaps + UMAPs (ATAC peak intensity, DevMap) | `chromatin-accesibility-clustering-and-visualization/03_fig2B_atac_soms.Rmd` |
| 2B brown synteny bands (ATAC) | chromatin notebooks |
| Full SOM clusters (Figs S18–S21) and BodyMap SOMs (Fig S25) | this notebook (saved separately below) |

Every panel this notebook makes is written to `results/` as an individual
vector **PDF**, ready to be placed into the Illustrator artboard.

This notebook expects `01_data_loading.Rmd` and `02_soms.Rmd` to have been run
first **in the same R session** (the README's intended workflow), so the
normalised matrices (`sd_norm` …) and SOM fits/annotation (`sd_som`,
`sd_som_info` …) are already in scope.

```{r setup_chain, include=FALSE}
# Packages, constants, paths and helpers.
source("R/setup.R")
```

## Initial full heatmaps (diagnostic)

Heatmaps drawn with the SOM unit order as it falls straight out of
`kohonen::som()`. This pass is diagnostic only; the next chunk reorders the
SOM-info tables into the manuscript row order and we redraw.

```{r full_heatmaps_initial}
# Build one ComplexHeatmap per SOM (BodyMap matrices are column-reordered into
# the canonical tissue x maturity x sex order before plotting).
sd_heatmap <- make_heatmap(sd_norm, sd_som, sd_som_info$class_number)
td_heatmap <- make_heatmap(td_norm, td_som, td_som_info$class_number)
sb_heatmap <- make_heatmap(sb_norm[, indices_order_ssal], sb_som, sb_som_info$class_number)
tb_heatmap <- make_heatmap(tb_norm[, indices_order_omyk], tb_som, tb_som_info$class_number)
```

## Reorder SOM-info tables to match the manuscript layout

The DevMap reorderings move two stage-specific units into the position the
manuscript heatmaps expect. The BodyMap edits flip a handful of
stage-specific flags that the automated 2.165 threshold mis-calls, then
re-arrange by constitutive / stage-specific / peak-tissue / peak-mean.

```{r reorder_som_info}
# DevMap: hand-tuned unit reordering for each species.
sd_som_info <- sd_som_info[c(1:9, 11, 12, 10, 13:16), ]
td_som_info <- td_som_info[c(1:8, 11, 10, 9, 12:16), ]

# Salmon BodyMap: correct one mis-called stage-specific flag, then re-sort.
sb_som_info$stage_specific[10] <- "FALSE"
sb_som_info <- arrange(sb_som_info, constitutive_class, desc(stage_specific),
                       class_max, desc(class_max_mean))

# Trout BodyMap: correct several flags, then re-sort the same way.
tb_som_info$stage_specific[c(5, 4, 9)] <- "FALSE"
tb_som_info$stage_specific[c(25, 27)]  <- "TRUE"
tb_som_info <- arrange(tb_som_info, constitutive_class, desc(stage_specific),
                       class_max, desc(class_max_mean))
```

## Final full heatmaps → supplementary PDFs (Figs S18–21, S25)

These full heatmaps (all clusters, including the constitutive/housekeeping
ones) are the **supplementary** SOM figures, not the curated Fig 2A panels.
Each is saved to its own PDF.

```{r full_heatmaps_final}
# Rebuild the heatmaps now the SOM-info row order is final.
sd_heatmap <- make_heatmap(sd_norm, sd_som, sd_som_info$class_number)
td_heatmap <- make_heatmap(td_norm, td_som, td_som_info$class_number)
sb_heatmap <- make_heatmap(sb_norm[, indices_order_ssal], sb_som, sb_som_info$class_number)
tb_heatmap <- make_heatmap(tb_norm[, indices_order_omyk], tb_som, tb_som_info$class_number)

# Save each full heatmap to its own supplementary PDF. draw() renders a
# ComplexHeatmap to the currently-open (pdf) graphics device.
save_heatmap_pdf <- function(hm, file, width = 5, height = 7) {
  pdf(file, width = width, height = height)  # open a PDF device
  ComplexHeatmap::draw(hm)                    # render the heatmap into it
  dev.off()                                   # close/flush the device
}
save_heatmap_pdf(sd_heatmap, "results/suppS18_full_som_devmap_salmon.pdf")
save_heatmap_pdf(td_heatmap, "results/suppS19_full_som_devmap_trout.pdf")
save_heatmap_pdf(sb_heatmap, "results/suppS25_full_som_bodymap_salmon.pdf")
save_heatmap_pdf(tb_heatmap, "results/suppS25_full_som_bodymap_trout.pdf")

# Also keep the legacy combined file for backward compatibility.
pdf("results/heatmaps.pdf")
plot(sd_heatmap); plot(td_heatmap); plot(sb_heatmap); plot(tb_heatmap)
dev.off()
```

## UMAP projections

UMAP run on the normalised expression matrices with `init = "pca"` so the
embeddings are deterministic given the input.

```{r umaps_all}
# One 2-D UMAP embedding per normalised matrix (rows = genes).
sd_umap <- sd_norm %>% uwot::umap(init = "pca")
td_umap <- td_norm %>% uwot::umap(init = "pca")
sb_umap <- sb_norm %>% uwot::umap(init = "pca")
tb_umap <- tb_norm %>% uwot::umap(init = "pca")
```

## Palettes

`palette_devmap()` assigns inferno to the stage-specific units and grey to the
constitutive ones, in the manuscript order. `palette_bodymap()` keys on the
peak tissue and stage-specific flag for each unit.

```{r palettes}
# DevMap palettes (n_stage_specific, n_constitutive, unit order).
sd_pal <- palette_devmap(12, 4, sd_som_info$class_number)
td_pal <- palette_devmap(11, 5, td_som_info$class_number)

# BodyMap palettes keyed on the annotation table.
sb_pal <- palette_bodymap(sb_som_info)
tb_pal <- palette_bodymap(tb_som_info)
```

## BodyMap UMAP + stage-specific panels → supplementary PDFs

The salmon/trout BodyMap UMAPs (coloured by SOM unit) are saved for the
supplementary BodyMap figure.

```{r umap_panels_bodymap}
# Salmon BodyMap UMAP coloured by SOM unit.
sb_umap_cols <- sb_umap %>% as.data.frame() %>%
  ggplot(aes(V1, -V2)) +
  geom_point(aes(colour = factor(sb_som$som$unit.classif,
                                 levels = sb_som_info$class_number)),
             alpha = .4, size = .1) +
  scale_color_manual(values = sb_pal) +
  theme_classic() +
  theme(legend.position = "none",
        axis.line = element_blank(), axis.text = element_blank(),
        axis.title = element_blank(), axis.ticks = element_blank())

# Trout BodyMap UMAP coloured by SOM unit.
tb_umap_cols <- tb_umap %>% as.data.frame() %>%
  ggplot(aes(-V1, -V2)) +
  geom_point(aes(colour = factor(tb_som$som$unit.classif,
                                 levels = tb_som_info$class_number)),
             alpha = .4, size = .1) +
  scale_color_manual(values = tb_pal) +
  theme_classic() +
  theme(legend.position = "none",
        axis.line = element_blank(), axis.text = element_blank(),
        axis.title = element_blank(), axis.ticks = element_blank())

# Save both BodyMap UMAPs as supplementary PDFs.
ggsave("results/suppS25_umap_bodymap_salmon.pdf", sb_umap_cols, width = 4, height = 4)
ggsave("results/suppS25_umap_bodymap_trout.pdf",  tb_umap_cols, width = 4, height = 4)
```

## Stage-specific heatmaps → Fig 2A / supplementary PDFs

The full heatmaps include the constitutive units, which compress the dynamic
range. Fig 2A shows only the stage-specific units; the panels below restrict to
those, per map, and save each as a PDF.

```{r stage_specific_heatmaps}
# DevMap stage-specific heatmaps (the Fig 2A heatmaps), per species.
sd_heatmap_specific <- make_heatmap(
  sd_norm[which(sd_som$som$unit.classif %in% sd_som_info$class_number[c(1:9, 11, 12)]), ],
  sd_som, sd_som_info$class_number[c(1:9, 11, 12)])
td_heatmap_specific <- make_heatmap(
  td_norm[which(td_som$som$unit.classif %in% td_som_info$class_number[c(1:8, 10, 11)]), ],
  td_som, td_som_info$class_number[c(1:8, 10, 11)])

# Save the two DevMap stage-specific heatmaps as Fig 2A PDFs.
save_heatmap_pdf(sd_heatmap_specific, "results/fig2A_heatmap_devmap_salmon.pdf", 4, 6)
save_heatmap_pdf(td_heatmap_specific, "results/fig2A_heatmap_devmap_trout.pdf",  4, 6)

# Capture them as grobs so they can be arranged into the composite below.
sd_heatmap_specific_grob <- grid.grabExpr(draw(sd_heatmap_specific))
td_heatmap_specific_grob <- grid.grabExpr(draw(td_heatmap_specific))
# Side-by-side grob (not drawn yet) of the two DevMap heatmaps.
specific_heatmaps <- arrangeGrob(sd_heatmap_specific_grob,
                                 td_heatmap_specific_grob, nrow = 1)

# BodyMap stage-specific heatmaps (supplementary).
sb_heatmap_specific <- make_heatmap(
  sb_norm[which(sb_som$som$unit.classif %in%
                sb_som_info$class_number[which(sb_som_info$stage_specific == TRUE)]),
          indices_order_ssal],
  sb_som,
  sb_som_info$class_number[which(sb_som_info$stage_specific == TRUE)])

tb_heatmap_specific <- make_heatmap(
  tb_norm[which(tb_som$som$unit.classif %in%
                tb_som_info$class_number[which(tb_som_info$stage_specific == TRUE)]),
          indices_order_omyk],
  tb_som,
  tb_som_info$class_number[which(tb_som_info$stage_specific == TRUE)])

# Save the BodyMap stage-specific heatmaps as supplementary PDFs.
save_heatmap_pdf(sb_heatmap_specific, "results/suppS25_heatmap_bodymap_salmon.pdf", 5, 6)
save_heatmap_pdf(tb_heatmap_specific, "results/suppS25_heatmap_bodymap_trout.pdf",  5, 6)
```

## Final inferno UMAPs → Fig 2A PDFs

The published Figure 2A UMAPs use a `desired_order` that maps the kohonen unit
IDs to the chronological cluster ordering used in the manuscript (M1..P2, then
C1..C4). `match(1:16, desired_order)` inverts the mapping so the palette stays
keyed by raw unit ID. These are the UMAP panels that go into Fig 2A.

```{r inferno_palettes_final}
# Salmon DevMap: 12 inferno colours for stage-specific units + 4 greys.
stage_specific_cols_ssal <- inferno(12, begin = 0.15, end = 0.9)
constitutive <- c("#becacaff")
final_colors_ssal <- c(stage_specific_cols_ssal, rep(constitutive, 4))

# Reorder the palette so palette[i] is the colour for raw kohonen unit i.
desired_order <- c(13, 10, 15, 14, 16, 12, 4, 2, 1, 6, 9, 5, 3, 8, 11, 7)
color_order_for_objects <- match(1:16, desired_order)
ordered_palette_salmon_devmap <- final_colors_ssal[color_order_for_objects]

# Salmon DevMap UMAP, coloured by the published palette.
sd_umap_cols <- sd_umap %>% as.data.frame() %>%
  ggplot(aes(-V1, -V2)) +
  geom_point(aes(colour = factor(sd_som$som$unit.classif)),
             alpha = .4, size = .8) +
  scale_color_manual(values = ordered_palette_salmon_devmap) +
  theme_classic() +
  theme(legend.position = "none") + ylim(c(7, -6)) + xlim(c(-8, 9))

# Trout DevMap: 11 inferno colours + 5 greys, same inversion trick.
stage_specific_cols_omyk <- inferno(11, begin = 0.15, end = 0.9)
final_colors_omyk <- c(stage_specific_cols_omyk, rep(constitutive, 5))

desired_order_trout <- c(13, 14, 9, 5, 1, 2, 3, 8, 12, 15, 16, 7, 4, 11, 6, 10)
color_order_for_objects <- match(1:16, desired_order_trout)
ordered_palette_trout_devmap <- final_colors_omyk[color_order_for_objects]

# Trout DevMap UMAP, coloured by the published palette.
td_umap_cols <- td_umap %>% as.data.frame() %>%
  ggplot(aes(-V1, -V2)) +
  geom_point(aes(colour = factor(td_som$som$unit.classif)),
             alpha = .4, size = 1) +
  scale_color_manual(values = ordered_palette_trout_devmap) +
  theme_classic() +
  theme(legend.position = "none") + ylim(7, -6) + xlim(8, -8)

# Save the two DevMap UMAPs as Fig 2A PDFs.
ggsave("results/fig2A_umap_devmap_salmon.pdf", sd_umap_cols, width = 4, height = 4)
ggsave("results/fig2A_umap_devmap_trout.pdf",  td_umap_cols, width = 4, height = 4)

# Side-by-side grob of the two published UMAPs, for the composite.
combined_umaps <- arrangeGrob(sd_umap_cols, td_umap_cols, nrow = 1)
```

## Figure 2A composite → PDF

A convenience composite stacking the two DevMap UMAPs over the two DevMap
stage-specific heatmaps. The manuscript's final Fig 2A is assembled in
Illustrator from the individual PDFs above; this composite is a quick-look
preview of how they fit together.

```{r fig2_composite}
# Stack UMAP row (1 unit tall) above heatmap row (3 units tall).
dev_rna_panel <- arrangeGrob(combined_umaps, specific_heatmaps,
                             ncol = 1, heights = c(1, 3))
# Save the composite, and also draw it so it appears in the knitted HTML.
ggsave("results/fig2A_devmap_composite.pdf", dev_rna_panel, width = 8, height = 10)
grid::grid.newpage(); grid::grid.draw(dev_rna_panel)
```
