---
title: "03 — Figure 2 (clustering and visualisation)"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---


## Overview

Builds the panels of Figure 2 of the manuscript: full and stage-specific
heatmaps for each SOM, UMAP projections coloured by SOM unit, and the
final composite. Expects the normalised matrices and SOM fits from
01–02 to be in scope.

```{r setup_chain, include=FALSE}
source("R/setup.R")
```

## Initial full heatmaps

Heatmaps drawn with the SOM unit order as it falls out of `kohonen::som()`.
This pass is diagnostic; the next chunk reorders the SOM-info tables so the
rows read in the manuscript order, and we redraw.

```{r full_heatmaps_initial}
sd_heatmap <- make_heatmap(sd_norm, sd_som, sd_som_info$class_number)
td_heatmap <- make_heatmap(td_norm, td_som, td_som_info$class_number)
sb_heatmap <- make_heatmap(sb_norm[, indices_order_ssal], sb_som, sb_som_info$class_number)
tb_heatmap <- make_heatmap(tb_norm[, indices_order_omyk], tb_som, tb_som_info$class_number)

pdf("./results/heatmaps.pdf")
plot(sd_heatmap); plot(td_heatmap); plot(sb_heatmap); plot(tb_heatmap)
dev.off()
```

## Reorder SOM-info tables to match the manuscript layout

The DevMap reorderings (`c(1:9,11,12,10,13:16)` for salmon and
`c(1:8,11,10,9,12:16)` for trout) move two stage-specific units into the
position that the manuscript heatmaps expect. The BodyMap edits flip a
handful of stage-specific flags that the automated 2.165 threshold either
misses or over-calls, then re-arrange by constitutive / stage-specific /
peak-tissue / peak-mean.

```{r reorder_som_info}
sd_som_info <- sd_som_info[c(1:9, 11, 12, 10, 13:16), ]
td_som_info <- td_som_info[c(1:8, 11, 10, 9, 12:16), ]

sb_som_info$stage_specific[10] <- "FALSE"
sb_som_info <- arrange(sb_som_info, constitutive_class, desc(stage_specific),
                       class_max, desc(class_max_mean))

tb_som_info$stage_specific[c(5, 4, 9)] <- "FALSE"
tb_som_info$stage_specific[c(25, 27)]  <- "TRUE"
tb_som_info <- arrange(tb_som_info, constitutive_class, desc(stage_specific),
                       class_max, desc(class_max_mean))
```

## Final full heatmaps

```{r full_heatmaps_final}
sd_heatmap <- make_heatmap(sd_norm, sd_som, sd_som_info$class_number)
td_heatmap <- make_heatmap(td_norm, td_som, td_som_info$class_number)
sb_heatmap <- make_heatmap(sb_norm[, indices_order_ssal], sb_som, sb_som_info$class_number)
tb_heatmap <- make_heatmap(tb_norm[, indices_order_omyk], tb_som, tb_som_info$class_number)

pdf("results/heatmaps.pdf")
plot(sd_heatmap); plot(td_heatmap); plot(sb_heatmap); plot(tb_heatmap)
dev.off()
```

## UMAP projections

UMAP run on the normalised expression matrices with `init = "pca"` so the
embeddings are deterministic given the input.

```{r umaps_all}
sd_umap <- sd_norm %>% uwot::umap(init = "pca")
td_umap <- td_norm %>% uwot::umap(init = "pca")
sb_umap <- sb_norm %>% uwot::umap(init = "pca")
tb_umap <- tb_norm %>% uwot::umap(init = "pca")
```

## Palettes

`palette_devmap()` assigns inferno to the 11 stage-specific units and grey to
the constitutive ones, in the manuscript order. `palette_bodymap()` keys on
the peak tissue and stage-specific flag for each unit.

```{r palettes}
sd_pal <- palette_devmap(12, 4, sd_som_info$class_number)
td_pal <- palette_devmap(11, 5, td_som_info$class_number)

sb_pal <- palette_bodymap(sb_som_info)
tb_pal <- palette_bodymap(tb_som_info)
```

## UMAP panels (initial palette pass)

```{r umap_panels}
sd_umap_cols <- sd_umap %>% as.data.frame() %>%
  ggplot(aes(V1, -V2)) +
  geom_point(aes(colour = factor(sd_som$som$unit.classif)),
             alpha = .4, size = .1) +
  scale_color_manual(values = sd_pal) +
  theme_classic() +
  theme(legend.position = "none",
        axis.line = element_blank(), axis.text = element_blank(),
        axis.title = element_blank(), axis.ticks = element_blank()) +
  ylim(c(7, -6)) + xlim(c(-8, 9))

td_umap_cols <- td_umap %>% as.data.frame() %>%
  ggplot(aes(-V1, -V2)) +
  geom_point(aes(colour = factor(td_som$som$unit.classif)),
             alpha = .4, size = .1) +
  scale_color_manual(values = td_pal) +
  ylim(-7, 6) + xlim(8, -8) +
  theme_classic() +
  theme(legend.position = "none",
        axis.line = element_blank(), axis.text = element_blank(),
        axis.title = element_blank(), axis.ticks = element_blank()) +
  ylim(7, -6)

combined_umaps <- grid.arrange(sd_umap_cols, td_umap_cols, nrow = 1)

sb_umap_cols <- sb_umap %>% as.data.frame() %>%
  ggplot(aes(V1, -V2)) +
  geom_point(aes(colour = factor(sb_som$som$unit.classif,
                                 levels = sb_som_info$class_number)),
             alpha = .4, size = .1) +
  scale_color_manual(values = sb_pal) +
  theme_classic() +
  theme(legend.position = "none",
        axis.line = element_blank(), axis.text = element_blank(),
        axis.title = element_blank(), axis.ticks = element_blank())

tb_umap_cols <- tb_umap %>% as.data.frame() %>%
  ggplot(aes(-V1, -V2)) +
  geom_point(aes(colour = factor(tb_som$som$unit.classif,
                                 levels = tb_som_info$class_number)),
             alpha = .4, size = .1) +
  scale_color_manual(values = tb_pal) +
  theme_classic() +
  theme(legend.position = "none",
        axis.line = element_blank(), axis.text = element_blank(),
        axis.title = element_blank(), axis.ticks = element_blank())
```

## Stage-specific heatmaps

The full heatmaps include the constitutive units, which compress the dynamic
range of the stage-specific signal. The panels below restrict to the
stage-specific units for each map.

```{r stage_specific_heatmaps}
sd_heatmap_specific <- make_heatmap(
  sd_norm[which(sd_som$som$unit.classif %in% sd_som_info$class_number[c(1:9, 11, 12)]), ],
  sd_som, sd_som_info$class_number[c(1:9, 11, 12)])
td_heatmap_specific <- make_heatmap(
  td_norm[which(td_som$som$unit.classif %in% td_som_info$class_number[c(1:8, 10, 11)]), ],
  td_som, td_som_info$class_number[c(1:8, 10, 11)])

sd_heatmap_specific_grob <- grid.grabExpr(draw(sd_heatmap_specific))
td_heatmap_specific_grob <- grid.grabExpr(draw(td_heatmap_specific))

specific_heatmaps <- grid.arrange(sd_heatmap_specific_grob,
                                  td_heatmap_specific_grob, nrow = 1)

sb_heatmap_specific <- make_heatmap(
  sb_norm[which(sb_som$som$unit.classif %in%
                sb_som_info$class_number[which(sb_som_info$stage_specific == TRUE)]),
          indices_order_ssal],
  sb_som,
  sb_som_info$class_number[which(sb_som_info$stage_specific == TRUE)])

tb_heatmap_specific <- make_heatmap(
  tb_norm[which(tb_som$som$unit.classif %in%
                tb_som_info$class_number[which(tb_som_info$stage_specific == TRUE)]),
          indices_order_omyk],
  tb_som,
  tb_som_info$class_number[which(tb_som_info$stage_specific == TRUE)])
```

## Figure 2 composite

```{r fig2_composite}
dev_rna_panel <- grid.arrange(combined_umaps, specific_heatmaps,
                              ncol = 1, heights = c(1, 3))
```

## Final inferno palettes for the published UMAP panels

The published Figure 2 UMAPs use a `desired_order` that maps the kohonen
unit IDs to the chronological cluster ordering used in the manuscript
(M1..P2, then C1..C4). `match(1:16, desired_order)` inverts the mapping so
the palette stays keyed by raw unit ID.

```{r inferno_palettes_final}
stage_specific_cols_ssal <- inferno(12, begin = 0.15, end = 0.9)
constitutive <- c("#becacaff")
final_colors_ssal <- c(stage_specific_cols_ssal, rep(constitutive, 4))

desired_order <- c(13, 10, 15, 14, 16, 12, 4, 2, 1, 6, 9, 5, 3, 8, 11, 7)
color_order_for_objects <- match(1:16, desired_order)
ordered_palette_salmon_devmap <- final_colors_ssal[color_order_for_objects]

sd_umap_cols <- sd_umap %>% as.data.frame() %>%
  ggplot(aes(-V1, -V2)) +
  geom_point(aes(colour = factor(sd_som$som$unit.classif)),
             alpha = .4, size = .8) +
  scale_color_manual(values = ordered_palette_salmon_devmap) +
  theme_classic() +
  theme(legend.position = "none") + ylim(c(7, -6)) + xlim(c(-8, 9))

stage_specific_cols_omyk <- inferno(11, begin = 0.15, end = 0.9)
final_colors_omyk <- c(stage_specific_cols_omyk, rep(constitutive, 5))

desired_order_trout <- c(13, 14, 9, 5, 1, 2, 3, 8, 12, 15, 16, 7, 4, 11, 6, 10)
color_order_for_objects <- match(1:16, desired_order_trout)
ordered_palette_trout_devmap <- final_colors_omyk[color_order_for_objects]

td_umap_cols <- td_umap %>% as.data.frame() %>%
  ggplot(aes(-V1, -V2)) +
  geom_point(aes(colour = factor(td_som$som$unit.classif)),
             alpha = .4, size = 1) +
  scale_color_manual(values = ordered_palette_trout_devmap) +
  ylim(-7, 6) + xlim(8, -8) +
  theme_classic() +
  theme(legend.position = "none") + ylim(7, -6)
```
