---
title: "04 — Ohnolog SOM co-clustering"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
---

## Overview

Per-cluster composition and per-pair co-clustering of AORe and LORe Atlantic
salmon ohnolog pairs across the DevMap (16-unit) and BodyMap (36-unit) SOMs.
The analysis is split into ten scripts under `scripts/`, chained through
RData outputs in `results/ohnolog/data/` and PDF / PNG panels in
`results/ohnolog/panels/`. Each section below sources one script; knitting
end-to-end reproduces every data object and figure panel.

Unlike `03`/`05`/`06`, this notebook is **self-contained**: the sourced scripts
load their own inputs (the deposited SOMs/matrices via the shared
`load_deposited_*()` loaders, the orthogroup table + raw TPM from Salmobase), so
`04` can be run without `01`/`02` having run first in the session.

```{r setup_chain, include=FALSE}
source("R/setup.R")
scripts_dir <- "scripts"
# Generated intermediates, figure panels and composites for this notebook.
for (d in c("results/ohnolog/data", "results/ohnolog/panels",
            "results/ohnolog/composite")) {
  dir.create(d, showWarnings = FALSE, recursive = TRUE)
}
```

## 1. Pair-level co-clustering (03_ohnolog_som_coclustering.R)

Loads the salmon SOM fits and the orthogroup table, extracts 2:2 AORe/LORe
salmon ohnolog pairs, joins each pair to its DevMap and BodyMap SOM
assignments, and runs a 2×2 Fisher's Exact Test on AORe vs LORe co-clustering
rates. The trout copy-number `nOmyk` is intentionally not constrained because
rediploidisation occurs independently in each species post-Ss4R, so the
AORe/LORe timing of a salmon pair does not depend on how many trout copies
survived. Writes `c1_coclustering_results.RData`.

```{r run_pair_level}
source(file.path(scripts_dir, "03_ohnolog_som_coclustering.R"))
```

## 2. Gene-level per-cluster breakdown (03b_ohnolog_som_extended.R)

Complements section 1 by asking, within each SOM cluster, what fraction of
resident ohnolog genes are AORe vs LORe, and what fraction have their
partner also in the same cluster. Computes per-cluster Fisher's tests for
AORe vs LORe partner-here rates (BH-adjusted, clusters with n ≥ 20), the
chi-squared test for uniform LORe proportion, and a 4-category composition
(Early / Late / Singleton / Other) per cluster. The composition baseline
uses the SOM-local pool to avoid the low-expression filter artefact.
Writes `c1_extended_results.RData`.

```{r run_gene_level}
source(file.path(scripts_dir, "03b_ohnolog_som_extended.R"))
```

## 3. Bar-chart panels (03c_ohnolog_som_figures.R)

Emits six paired PDF/PNG panels per map: per-cluster co-clustering rate
(all and no-constitutive variants), aggregated co-clustering by stage or
tissue group, and per-cluster LORe proportion. Cluster P* (raw SOM unit 9)
is excluded because the manuscript treats it separately. The const/noconst
split lets the composite stage (section 9) draw from the noconst variant.

```{r make_bar_panels}
source(file.path(scripts_dir, "03c_ohnolog_som_figures.R"))
```

## 4. Sankey panels (03d_ohnolog_sankey.R)

Builds the Sankey/alluvial panels showing which DevMap cluster the partner
of a salmon ohnolog ends up in, separately for AORe (Early) and LORe (Late)
pairs. Same const/noconst split as section 3.

```{r make_sankey}
source(file.path(scripts_dir, "03d_ohnolog_sankey.R"))
```

## 5. UpSet panels (03e_ohnolog_upset.R)

UpSet plots over the set of clusters that contain at least one AORe (or
LORe) co-clustered pair, showing the intersection structure between
clusters. Per the UpSetR colour-passthrough bug, palette membership is
derived from `unique(unlist(intersections))` rather than the `sets`
parameter.

```{r make_upset}
source(file.path(scripts_dir, "03e_ohnolog_upset.R"))
```

## 6. Fold-change pattern (03f_ohnolog_fc_pattern.R)

Computes the per-pair fold-change pattern over developmental stages and
tissues for AORe vs LORe ohnologs, and writes the associated panels and
support tables.

```{r make_fc_pattern}
source(file.path(scripts_dir, "03f_ohnolog_fc_pattern.R"))
```

## 7. Early/Late ratio panel (03g_early_late_ratio_panel.R)

Per-cluster Late-to-Early co-clustering rate ratio with 95% CIs, the panel
that recapitulates the Fig. 3 constraint pattern (ratio rises with
developmental progression as Early ohnologs accumulate constraint).

```{r make_el_ratio}
source(file.path(scripts_dir, "03g_early_late_ratio_panel.R"))
```

## 8. LORe ratio panel (03g_lore_ratio_panel.R)

Per-cluster LORe-fraction panel, paired with the Early/Late ratio in
section 7. The two `03g_` scripts share infrastructure but write distinct
PDF outputs.

```{r make_lore_ratio}
source(file.path(scripts_dir, "03g_lore_ratio_panel.R"))
```

## 9. Composite supplementary figures (04_composite_figures.R)

Stitches the panels produced in sections 3-5 into the two supplementary
composite figures (DevMap + BodyMap). The composite reads PNG sources
when present because UpSetR PDFs rasterise unreliably via pdftools;
PDFs are kept alongside for downstream LaTeX inclusion.

```{r build_composites}
source(file.path(scripts_dir, "04_composite_figures.R"))
```

## 10. Fold-change composite (04b_composite_fc_pattern.R)

Final composite tying the fold-change pattern panels (section 6) into the
matching supplementary figure.

```{r build_fc_composite}
source(file.path(scripts_dir, "04b_composite_fc_pattern.R"))
```
