---
title: "05 — Observed vs expected ortholog co-clustering"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---


## Overview

For each pair of (salmon SOM cluster, trout SOM cluster) this notebook counts
the number of shared sub-orthogroups (observed), permutes the trout SOM
assignments `N_PERMUTATIONS` times (10,000 in production) to build an expected distribution under random
co-clustering (expected), and reports the per-cell empirical p-value and
observed-to-expected ratio. Heatmaps and Sankey panels visualise the
resulting matrices. The same pipeline runs in parallel for DevMap (16×16)
and BodyMap (36×36).

```{r setup_chain, include=FALSE}
source("R/setup.R")
# Expects sd_norm / sb_norm / sd_som / sb_som and `orthogroups` in scope
# (produced by 01 and 02). The reorder vectors used below are tied to the
# manuscript SOM ordering.
```

## Build SOM × orthogroup lists

`sd_ids_som[[i]]` is the set of sub-orthogroup IDs for the salmon genes in
DevMap unit i; `td_ids_som`, `sb_ids_som`, `tb_ids_som` analogous. The
`_filtered` variants drop many-copy orthogroups (nSsal < 3 & nOmyk < 3) and
are kept available for sensitivity checks even though the main analysis
uses the unfiltered lists.

```{r build_som_orthogroup_lists}
# DevMap (16 units): for each unit, map its member genes (sd_ids/td_ids from 02)
# to their sub-orthogroup IDs. These subOG lists are what we intersect to count
# shared orthogroups between salmon and trout clusters.
sd_ids_som <- list(); td_ids_som <- list()
for (i in 1:16) {
  sd_ids_som[[i]] <- orthogroups$subOG[which(orthogroups$geneID %in% sd_ids[[i]])]
  td_ids_som[[i]] <- orthogroups$subOG[which(orthogroups$geneID %in% td_ids[[i]])]
}

# BodyMap (36 units): same mapping.
sb_ids_som <- list(); tb_ids_som <- list()
for (i in 1:36) {
  sb_ids_som[[i]] <- orthogroups$subOG[which(orthogroups$geneID %in% sb_ids[[i]])]
  tb_ids_som[[i]] <- orthogroups$subOG[which(orthogroups$geneID %in% tb_ids[[i]])]
}

# `_filtered` variants: same, but restricted to low-copy orthogroups
# (nSsal < 3 & nOmyk < 3). Kept for sensitivity checks; not used in the main run.
sd_ids_som_filtered <- list(); td_ids_som_filtered <- list()
for (i in 1:16) {
  sd_ids_som_filtered[[i]] <- orthogroups$subOG[which(
    orthogroups$geneID %in% sd_ids[[i]] &
      orthogroups$nSsal < 3 & orthogroups$nOmyk < 3)]
  td_ids_som_filtered[[i]] <- orthogroups$subOG[which(
    orthogroups$geneID %in% td_ids[[i]] &
      orthogroups$nSsal < 3 & orthogroups$nOmyk < 3)]
}
sb_ids_som_filtered <- list(); tb_ids_som_filtered <- list()
for (i in 1:36) {
  sb_ids_som_filtered[[i]] <- orthogroups$subOG[which(
    orthogroups$geneID %in% sb_ids[[i]] &
      orthogroups$nSsal < 3 & orthogroups$nOmyk < 3)]
  tb_ids_som_filtered[[i]] <- orthogroups$subOG[which(
    orthogroups$geneID %in% tb_ids[[i]] &
      orthogroups$nSsal < 3 & orthogroups$nOmyk < 3)]
}
```

## Observed shared-orthogroup matrices

Cell (i, j) is the number of sub-orthogroups present in both salmon unit i
and trout unit j. Two matrices: DevMap (16×16) and BodyMap (36×36).

```{r initial_observed_matrices}
# DevMap observed matrix: cell (i,j) = number of subOGs shared between salmon
# DevMap unit i and trout DevMap unit j (size of the subOG-list intersection).
o_t <- data.frame(matrix(nrow = 16, ncol = 16))
rownames(o_t) <- 1:16; colnames(o_t) <- 1:16
for (i in 1:16) for (j in 1:16) {
  o_t[i, j] <- length(sd_ids_som[[i]][which(sd_ids_som[[i]] %in% td_ids_som[[j]])])
}

# BodyMap observed matrix (36x36): salmon BodyMap unit i vs trout BodyMap unit j.
o_t.b <- data.frame(matrix(nrow = 36, ncol = 36))
rownames(o_t.b) <- 1:36; colnames(o_t.b) <- 1:36
for (i in 1:36) for (j in 1:36) {
  o_t.b[i, j] <- length(sb_ids_som[[i]][which(sb_ids_som[[i]] %in% tb_ids_som[[j]])])
}
```

## Permutation setup

Each replicate randomly re-assigns the trout SOM labels across the trout
gene list while preserving the per-cluster cluster-size distribution. The
join to `orthogroups[, c("subOG", "geneID")]` carries the subOG label through the
permutation. `N_PERMUTATIONS` is 10,000 in production and 100 in test
mode (`TEST_MODE = TRUE` in `R/setup.R`).

```{r permutation_setup, cache=TRUE}
# Build N_PERMUTATIONS shuffled trout label assignments. Each replicate:
# randomly permutes the trout gene order, re-attaches the (unchanged) SOM unit
# vector somID, and joins each gene to its subOG — so SOM labels are shuffled
# relative to orthogroup identity while the per-unit cluster sizes are preserved.
td_norm_rownames <- as.data.frame(rownames(td_norm))
colnames(td_norm_rownames) <- "geneID"
tenk_list <- list()
for (i in seq_len(N_PERMUTATIONS)) {
  tenk_list[[i]] <- as.data.frame(sample(td_norm_rownames$geneID))
  colnames(tenk_list[[i]]) <- "geneID"
  tenk_list[[i]] <- left_join(tenk_list[[i]], orthogroups[, c("subOG", "geneID")],
                              by = c("geneID" = "geneID"))
  tenk_list[[i]]$somID <- td_som$som$unit.classif
  tenk_list[[i]] <- tenk_list[[i]] %>% na.omit()
}

tb_norm_rownames <- as.data.frame(rownames(tb_norm))
colnames(tb_norm_rownames) <- "geneID"
tenk_list.b <- list()
for (i in seq_len(N_PERMUTATIONS)) {
  tenk_list.b[[i]] <- as.data.frame(sample(tb_norm_rownames$geneID))
  colnames(tenk_list.b[[i]]) <- "geneID"
  tenk_list.b[[i]] <- dplyr::left_join(tenk_list.b[[i]], orthogroups[, c("subOG", "geneID")],
                                       by = c("geneID" = "geneID"))
  tenk_list.b[[i]]$somID <- tb_som$som$unit.classif
  tenk_list.b[[i]] <- tenk_list.b[[i]] %>% na.omit()
}
```

## Random observations per cell

Re-runs the observed-matrix calculation for each shuffled trout label
vector, producing a (256 × `N_PERMUTATIONS`) DevMap distribution and a
(1296 × `N_PERMUTATIONS`) BodyMap distribution.

```{r permutation_observations, cache=TRUE}
# Recompute the observed shared-subOG count for every (salmon unit j, trout
# unit l) cell under each shuffle, giving the null distribution per cell. Cells
# are flattened to rows via the index 16*(j-1)+l (DevMap) / 36*(j-1)+l (BodyMap).
random_observations <- data.frame(matrix(nrow = 256, ncol = N_PERMUTATIONS))
salnames <- paste(rep("S", 16), rep(1:16, each = 16), sep = "")
omnames  <- paste(rep("T", 16), rep(1:16, 16), sep = "")
rownames(random_observations) <- paste(salnames, omnames, sep = "x")

for (i in seq_len(N_PERMUTATIONS)) {
  for (j in 1:16) for (l in 1:16) {
    salmon_som_orthos <- sd_ids_som[[j]]
    trout_som_orthos <- tenk_list[[i]]$subOG[which(tenk_list[[i]]$somID == l)]
    random_observations[16 * (j - 1) + l, i] <- sum(salmon_som_orthos %in% trout_som_orthos)
  }
}

random_observations.b <- data.frame(matrix(nrow = 1296, ncol = N_PERMUTATIONS))
salnames.b <- paste(rep("S", 36), rep(1:36, each = 36), sep = "")
omnames.b  <- paste(rep("T", 36), rep(1:36, 36), sep = "")
rownames(random_observations.b) <- paste(salnames.b, omnames.b, sep = "x")

for (i in seq_len(N_PERMUTATIONS)) {
  for (j in 1:36) for (l in 1:36) {
    salmon_som_orthos.b <- sb_ids_som[[j]]
    trout_som_orthos.b <- tenk_list.b[[i]]$subOG[which(tenk_list.b[[i]]$somID == l)]
    random_observations.b[36 * (j - 1) + l, i] <-
      sum(salmon_som_orthos.b %in% trout_som_orthos.b)
  }
}
```

## Empirical p-values

`-log10(p)` with a 1e-6 floor so cells where every permutation undershoots
the observed do not become `Inf`.

```{r empirical_pvalues}
# Per-cell empirical p = fraction of permutations whose count exceeds the
# observed; reported as -log10(p) with a 1e-6 floor (so cells never observed to
# be exceeded don't become Inf). Higher = more significant co-clustering.
pvals_mat_log <- data.frame(matrix(nrow = 16, ncol = 16))
for (i in 1:16) for (j in 1:16) {
  pvals_mat_log[i, j] <- -log10(0.000001 +
    length(which(random_observations[16 * (i - 1) + j, ] > o_t[i, j])) / N_PERMUTATIONS)
}

pvals_mat_log.b <- data.frame(matrix(nrow = 36, ncol = 36))
for (i in 1:36) for (j in 1:36) {
  pvals_mat_log.b[i, j] <- -log10(0.000001 +
    length(which(random_observations.b[36 * (i - 1) + j, ] > o_t.b[i, j])) / N_PERMUTATIONS)
}
```

## Observed / expected ratios

```{r observed_expected_ratios}
# Observed-over-expected ratio per cell = observed count / mean permuted count.
# These ratios are the light-red bands overlaid on Figure 2A.
oe_mat <- data.frame(matrix(nrow = 16, ncol = 16))
random_obs_mean <- rowMeans(random_observations)
rand_obs_wide   <- matrix(random_obs_mean, nrow = 16, ncol = 16, byrow = TRUE)
for (i in 1:16) for (j in 1:16) {
  oe_mat[i, j] <- o_t[i, j] / rand_obs_wide[i, j]
}

oe_mat.b <- data.frame(matrix(nrow = 36, ncol = 36))
random_obs_mean.b <- rowMeans(random_observations.b)
rand_obs_wide.b   <- matrix(random_obs_mean.b, nrow = 36, ncol = 36, byrow = TRUE)
for (i in 1:36) for (j in 1:36) {
  oe_mat.b[i, j] <- o_t.b[i, j] / rand_obs_wide.b[i, j]
}
```

## Reorder matrices to manuscript SOM order

The DevMap reorderings move the rows/cols into the M1..P2 sequence used in
the manuscript heatmaps; the BodyMap reorderings consume the
`class_number` order from `sb_som_info` / `tb_som_info`.

```{r reorder_matrices}
pvals_mat_order_log <- pvals_mat_log[
  c(13, 10, 15, 14, 16, 12, 4, 2, 1, 5, 9, 6, 7, 11, 8, 3),
  c(13, 14, 9, 5, 1, 2, 3, 8, 16, 12, 15, 6, 10, 7, 4, 11)]

pvals_mat_order_log.b <- pvals_mat_log.b[
  as.numeric(sb_som_info$class_number),
  as.numeric(tb_som_info$class_number)]

oe_mat <- oe_mat[
  c(13, 10, 15, 14, 16, 12, 4, 2, 1, 9, 6, 5, 7, 11, 8, 3),
  c(13, 14, 9, 5, 1, 2, 3, 8, 15, 12, 16, 6, 10, 7, 4, 11)]

oe_mat.b <- oe_mat.b[
  as.numeric(sb_som_info$class_number),
  as.numeric(tb_som_info$class_number)]
```

## Heatmap panels

```{r heatmap_panels}
# Build the ComplexHeatmaps of the empirical p-values and the log2 o/e ratios,
# in manuscript SOM order (full + stage-specific subsets, DevMap + BodyMap).
# `_specific` variants drop the constitutive units. These are visualised in the
# HTML; the o/e ratio is what feeds the Fig 2A bands.
pval_heatmap <- Heatmap(as.matrix(pvals_mat_order_log), name = "Empirical p-val",
  cluster_column_slices = FALSE, cluster_row_slices = FALSE,
  show_row_dend = FALSE, show_row_names = TRUE,
  col = colorRamp2(c(1, 0.5, 0), c("red", "yellow", "blue")),
  show_column_names = TRUE, show_column_dend = FALSE,
  use_raster = TRUE, raster_by_magick = TRUE,
  cluster_columns = FALSE, cluster_rows = FALSE,
  rect_gp = gpar(col = "white", lwd = 2))

o_e_heatmap <- Heatmap(log2(as.matrix(oe_mat) + 0.1), name = "o/e ratio",
  cluster_column_slices = FALSE, cluster_row_slices = FALSE,
  show_row_dend = FALSE, show_row_names = TRUE,
  col = colorRamp2(c(-3, 0, 3), c("skyblue", "seashell", "lightcoral")),
  show_column_names = TRUE, show_column_dend = FALSE,
  use_raster = TRUE, raster_by_magick = TRUE,
  cluster_columns = FALSE, cluster_rows = FALSE,
  rect_gp = gpar(col = "white", lwd = 2))

o_e_heatmap_specific <- Heatmap(log2(as.matrix(oe_mat[c(1:9, 11, 12), c(1:8, 10, 11)]) + 0.1),
  name = "o/e ratio",
  cluster_column_slices = FALSE, cluster_row_slices = FALSE,
  show_row_dend = FALSE, show_row_names = TRUE,
  col = colorRamp2(c(-3, 0, 3), c("skyblue", "seashell", "lightcoral")),
  show_column_names = TRUE, show_column_dend = FALSE,
  use_raster = TRUE, raster_by_magick = TRUE,
  cluster_columns = FALSE, cluster_rows = FALSE,
  rect_gp = gpar(col = "white", lwd = 2))

o_e_heatmap.b <- Heatmap(log2(as.matrix(oe_mat.b) + 0.1), name = "o/e ratio",
  cluster_column_slices = FALSE, cluster_row_slices = FALSE,
  show_row_dend = FALSE, show_row_names = TRUE,
  col = colorRamp2(c(-3, 0, 3), c("skyblue", "seashell", "lightcoral")),
  show_column_names = TRUE, show_column_dend = FALSE,
  use_raster = TRUE, raster_by_magick = TRUE,
  cluster_columns = FALSE, cluster_rows = FALSE,
  rect_gp = gpar(col = "white", lwd = 2))

o_e_heatmap_specific.b <- Heatmap(log2(as.matrix(oe_mat.b[1:20, 1:21]) + 0.1),
  name = "o/e ratio",
  cluster_column_slices = FALSE, cluster_row_slices = FALSE,
  show_row_dend = FALSE, show_row_names = TRUE,
  col = colorRamp2(c(-3, 0, 3), c("skyblue", "seashell", "lightcoral")),
  show_column_names = TRUE, show_column_dend = FALSE,
  use_raster = TRUE, raster_by_magick = TRUE,
  cluster_columns = FALSE, cluster_rows = FALSE,
  rect_gp = gpar(col = "white", lwd = 2))

# Save each heatmap as its own vector PDF (the o/e ratio heatmaps are the
# Fig 2A panels; the p-value + BodyMap variants are supplementary). draw()
# renders a ComplexHeatmap into the open pdf() device.
save_hm <- function(hm, file, w = 5, h = 6) {
  pdf(file.path("results", file), width = w, height = h)
  ComplexHeatmap::draw(hm); dev.off()
}
save_hm(o_e_heatmap,            "fig2A_oe_ratio_devmap.pdf")
save_hm(o_e_heatmap_specific,   "fig2A_oe_ratio_devmap_specific.pdf")
save_hm(pval_heatmap,           "suppS_pvalue_devmap.pdf")
save_hm(o_e_heatmap.b,          "suppS_oe_ratio_bodymap.pdf", 7, 7)
save_hm(o_e_heatmap_specific.b, "suppS_oe_ratio_bodymap_specific.pdf", 7, 7)
```

## Sankey panel — RNA shared orthogroups

The DevMap Sankey uses the reordered, stage-specific subset of the
observation matrix (rows 1:9, 11, 12; cols 1:8, 10, 11). Edge colour is the
log2 o/e ratio mapped onto a skyblue → seashell → lightcoral palette.

```{r sankey_rna_overlap}
o_st <- data.frame(matrix(nrow = 16, ncol = 16))
rownames(o_st) <- 1:16; colnames(o_st) <- 1:16
for (i in 1:16) for (j in 1:16) {
  o_st[i, j] <- length(sd_ids_som[[i]][which(sd_ids_som[[i]] %in% td_ids_som[[j]])])
}
o_st_order <- o_st[
  c(13, 10, 15, 14, 16, 12, 4, 2, 1, 9, 6, 5, 7, 11, 8, 3),
  c(13, 14, 9, 5, 1, 2, 3, 8, 15, 12, 16, 6, 10, 7, 4, 11)]
o_st_order_trim <- o_st_order[c(1:9, 11, 12), c(1:8, 10, 11)]

salmon_sum <- rowSums(o_st_order_trim)
trout_sum  <- colSums(o_st_order_trim)

o_st_long <- pivot_longer(o_st_order_trim,
                          names_to = "trout_clusters",
                          values_to = "shared orthologues",
                          cols = everything())
o_st_long$salmon_clusters <- rep(c(13, 10, 15, 14, 16, 12, 4, 2, 1, 6, 5),
                                 each = 10)

oe_mat_specific      <- log2(as.matrix(oe_mat[c(1:9, 11, 12), c(1:8, 10, 11)]) + 0.1)
oe_mat_specific_long <- pivot_longer(as.data.frame(oe_mat_specific),
                                     everything(),
                                     names_to = "trout_cluster",
                                     values_to = "o_eratio")

data <- data.frame(
  source = factor(paste("s", o_st_long$salmon_clusters, sep = ""),
                  levels = c("s13", "s10", "s15", "s14",
                             "s16", "s12", "s4", "s2",
                             "s1", "s6", "s5")),
  target = factor(paste("t", o_st_long$trout_clusters, sep = ""),
                  levels = c("t13", "t14", "t9", "t5",
                             "t1", "t2", "t3", "t8",
                             "t12", "t16")),
  count = o_st_long$`shared orthologues`,
  color = oe_mat_specific_long$o_eratio)

color_palette <- colorRampPalette(c("skyblue", "seashell", "lightcoral"))
rescaled_values <- rescale(data$color, to = c(0, 1))
data$colors <- color_palette(nrow(data))[as.integer(round(rescaled_values * 99)) + 1]

# DevMap RNA o/e Sankey — the Fig 2A observed/expected ratio bands. Assign,
# display, and save to its own PDF.
rna_sankey_devmap <- ggplot(data, aes(axis1 = source, axis2 = target, y = count)) +
  geom_alluvium(aes(fill = I(colors)), alpha = 0.7) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal()
print(rna_sankey_devmap)
ggsave("results/fig2A_oe_sankey_devmap.pdf", rna_sankey_devmap, width = 8, height = 6)

# BodyMap equivalent
o_st.b <- data.frame(matrix(nrow = 36, ncol = 36))
rownames(o_st.b) <- 1:36; colnames(o_st.b) <- 1:36
for (i in 1:36) for (j in 1:36) {
  o_st.b[i, j] <- length(sb_ids_som[[i]][which(sb_ids_som[[i]] %in% tb_ids_som[[j]])])
}
o_st_order.b <- o_st.b[as.numeric(sb_som_info$class_number),
                       as.numeric(tb_som_info$class_number)]
o_st_order_trim.b <- o_st_order.b[1:20, 1:21]

o_st_long.b <- pivot_longer(o_st_order_trim.b,
                            names_to = "trout_clusters",
                            values_to = "shared orthologues",
                            cols = everything())
o_st_long.b$salmon_clusters <- rep(as.numeric(rownames(o_st_order_trim.b)),
                                   each = length(o_st_order_trim.b))

oe_mat_specific.b      <- log2(as.matrix(oe_mat.b[1:20, 1:21]) + 0.1)
oe_mat_specific_long.b <- pivot_longer(as.data.frame(oe_mat_specific.b),
                                       everything(),
                                       names_to = "trout_cluster",
                                       values_to = "o_eratio")

data.b <- data.frame(
  source = factor(paste("s", o_st_long.b$salmon_clusters, sep = ""),
                  levels = unique(paste("s", o_st_long.b$salmon_clusters, sep = ""))),
  target = factor(paste("t", o_st_long.b$trout_clusters, sep = ""),
                  levels = unique(paste("t", o_st_long.b$trout_clusters, sep = ""))),
  count = o_st_long.b$`shared orthologues`,
  color = oe_mat_specific_long.b$o_eratio)

color_palette.b <- colorRampPalette(c("skyblue", "seashell", "lightcoral"))
rescaled_values.b <- rescale(data.b$color, to = c(0, 1))
data.b$colors <- color_palette.b(nrow(data.b))[as.integer(round(rescaled_values.b * nrow(data.b))) + 1]

# BodyMap RNA o/e Sankey (supplementary). Assign, display, and save.
rna_sankey_bodymap <- ggplot(data.b, aes(axis1 = source, axis2 = target, y = count)) +
  geom_alluvium(aes(fill = I(colors)), alpha = 0.7) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal()
print(rna_sankey_bodymap)
ggsave("results/suppS_oe_sankey_bodymap.pdf", rna_sankey_bodymap, width = 8, height = 6)
```

## Sankey panel — ATAC peak overlap (DevMap)

ATAC peak SOM overlap counts come from upstream Lars-Gronvold-pipeline TSVs.
`Ssal_clusters` / `Omyk_clusters` filter to the manuscript SOM-cluster
selection.

```{r sankey_atac_overlap_dev}
overlapping_peaks_devmap <- read_tsv("data/DevMapSOMpeakOverlapCount.tsv")
Ssal_clusters <- c(21, 22, 16, 17, 11, 7, 6, 1, 2, 8, 4, 5, 10, 20, 14, 24, 25)
Omyk_clusters <- c(16, 17, 21, 6, 7, 1, 19, 2, 3, 5, 18, 23, 24, 20, 14, 15, 25)

overlapping_peaks_devmap_fs <- overlapping_peaks_devmap[
  which(overlapping_peaks_devmap$cluster_Ssal %in% Ssal_clusters), ]
overlapping_peaks_fst <- overlapping_peaks_devmap_fs[
  which(overlapping_peaks_devmap_fs$cluster_Omyk %in% Omyk_clusters), ]

overlapping_peaks_fst$cluster_Omyk <- factor(
  paste("t", overlapping_peaks_fst$cluster_Omyk, sep = ""),
  levels = paste("t", Omyk_clusters, sep = ""))
overlapping_peaks_fst$cluster_Ssal <- factor(
  paste("s", overlapping_peaks_fst$cluster_Ssal, sep = ""),
  levels = paste("s", Ssal_clusters, sep = ""))

# DevMap ATAC peak-overlap Sankey. Assign, display, and save.
atac_sankey_devmap <- ggplot(overlapping_peaks_fst,
       aes(axis1 = cluster_Ssal, axis2 = cluster_Omyk, y = n)) +
  geom_alluvium(aes(fill = n), alpha = 0.7) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal() +
  scale_fill_gradient(low = "lightgrey", high = "coral4")
print(atac_sankey_devmap)
ggsave("results/atac_peak_overlap_sankey_devmap.pdf", atac_sankey_devmap, width = 8, height = 6)
```

## Sankey panel — ATAC peak overlap (BodyMap)

```{r sankey_atac_overlap_body}
overlapping_peaks_bodymap <- read_tsv("data/BodyMapSOMpeakOverlapCount.tsv")
Ssal_clusters_b <- c(22, 23, 7, 6, 1, 13, 19, 25, 24)
Omyk_cluster_b  <- c(3, 2, 17, 16, 21, 20, 25)

overlapping_peaks_bodymap_fs <- overlapping_peaks_bodymap[
  which(overlapping_peaks_bodymap$cluster_Ssal %in% Ssal_clusters_b), ]
overlapping_peaks_bodymap_fst <- overlapping_peaks_bodymap_fs[
  which(overlapping_peaks_bodymap_fs$cluster_Omyk %in% Omyk_cluster_b), ]

overlapping_peaks_bodymap_fst$cluster_Ssal <- factor(
  paste("s", overlapping_peaks_bodymap_fst$cluster_Ssal, sep = ""),
  levels = paste("s", Ssal_clusters_b, sep = ""))
overlapping_peaks_bodymap_fst$cluster_Omyk <- factor(
  paste("t", overlapping_peaks_bodymap_fst$cluster_Omyk, sep = ""),
  levels = paste("t", Omyk_cluster_b, sep = ""))

# BodyMap ATAC peak-overlap Sankey. Assign, display, and save.
atac_sankey_bodymap <- ggplot(overlapping_peaks_bodymap_fst,
       aes(axis1 = cluster_Ssal, axis2 = cluster_Omyk, y = n)) +
  geom_alluvium(aes(fill = n), alpha = 0.7) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal() +
  scale_fill_gradient(low = "lightgrey", high = "coral4")
print(atac_sankey_bodymap)
ggsave("results/atac_peak_overlap_sankey_bodymap.pdf", atac_sankey_bodymap, width = 8, height = 6)
```
