---
title: "06 — GO enrichment across SOM clusters"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---


## Overview

GO biological-process enrichment per SOM cluster, run separately for the
DevMap and BodyMap SOMs of each species. The DevMap clusters are passed in
their stage-specific manuscript order; the BodyMap clusters are grouped by
the peak tissue with a single "constitutive" bucket. clusterProfiler's
`compareCluster` does the enrichment, using the per-species OrgDB packages
built in `99a_make_orgdb_salmon.Rmd` / `99b_make_orgdb_trout.Rmd`.

```{r setup_chain, include=FALSE}
source("R/setup.R")
```

## Packages

GO enrichment uses clusterProfiler together with the two custom OrgDB
packages built in `99a_make_orgdb_salmon.Rmd` / `99b_make_orgdb_trout.Rmd`.

```{r packages_go, message=FALSE, warning=FALSE}
library(clusterProfiler)   # compareCluster / enrichGO
library(org.Ssalar.eg.db)  # salmon OrgDB (built by 99a)
library(org.Omykiss.eg.db) # trout OrgDB (built by 99b)
```

# Salmon DevMap

## Build gene clusters

`specific_sd` is the stage-specific cluster ordering; `constitutive_sd`
carries the four constitutive units that are appended as a single
constitutive group when needed.

```{r build_gene_clusters_sd}
constitutive_sd <- c(9, 7, 11, 8, 9)                                  # constitutive units
specific_sd     <- c(13, 10, 15, 14, 16, 12, 4, 2, 1, 6, 5, 9)        # stage-specific order
universe        <- rownames(sd_norm)                                  # all genes (gene IDs)

# One gene-list per SOM unit (the genes assigned to that unit), in specific_sd order.
geneClusters_sd <- list()
for (stage in specific_sd) {
  stage_num <- match(stage, specific_sd)                              # position in the ordered list
  geneClusters_sd[[stage_num]] <- universe[which(sd_som$som$unit.classif == stage)]
}
names(geneClusters_sd) <- specific_sd
```

## Run compareCluster

The `keyType = "GID"` argument is required by the custom OrgDB packages,
where the salmon and trout gene IDs are stored under the `GID` key rather
than the default `ENTREZID`. Both readable and raw outputs are kept so the
downstream tables can carry SYMBOL labels and the genuine gene IDs.

```{r run_compareCluster_sd, cache=TRUE}
# GO BP over-representation per cluster (BH-adjusted). keyType="GID" because the
# custom OrgDBs key genes under GID, not the default ENTREZID. Two runs:
# readable=TRUE maps IDs to gene SYMBOLs; readable=FALSE keeps the raw gene IDs.
compareClusterObj_sd <- compareCluster(
  geneClusters_sd, fun = "enrichGO",
  OrgDb = org.Ssalar.eg.db, keyType = "GID",
  ont = "BP", pAdjustMethod = "BH", readable = TRUE)

compareClusterObj_sd_nr <- compareCluster(
  geneClusters_sd, fun = "enrichGO",
  OrgDb = org.Ssalar.eg.db, keyType = "GID",
  ont = "BP", pAdjustMethod = "BH", readable = FALSE)
```

# Salmon BodyMap

## Build gene clusters

Tissue-specific clusters are split by `tissue`; the constitutive units are
collapsed into a single `constitutive` entry. The list is reordered
alphabetically by name for stable downstream plotting.

```{r build_gene_clusters_sb}
# Gene -> SOM unit lookup (named vector: gene ID -> unit).
gene_class <- sb_som$som$unit.classif
names(gene_class) <- rownames(sb_norm)

# Stage-specific, non-constitutive units and their peak tissue.
tissue_df <- sb_som_info[
  sb_som_info$stage_specific == TRUE &
    sb_som_info$constitutive_class == FALSE,
  c("tissue", "class_number")]

# The constitutive units, collapsed into one group below.
constitutive_classes <- as.numeric(
  sb_som_info[sb_som_info$constitutive_class == TRUE, "class_number"])

# One gene-list per tissue (union of that tissue's units' genes)...
geneClusters_sb <- lapply(
  split(tissue_df$class_number, tissue_df$tissue),
  function(cls) names(gene_class)[gene_class %in% cls])

# ...plus a single "constitutive" group.
geneClusters_sb[["constitutive"]] <- names(gene_class)[
  gene_class %in% constitutive_classes]

# Alphabetical by group name for stable downstream plotting.
geneClusters_sb <- geneClusters_sb[order(names(geneClusters_sb))]
```

## Run compareCluster

```{r run_compareCluster_sb, cache=TRUE}
compareClusterObj_sb <- compareCluster(
  geneClusters_sb, fun = "enrichGO",
  OrgDb = org.Ssalar.eg.db, keyType = "GID",
  ont = "BP", pAdjustMethod = "BH", readable = TRUE)

compareClusterObj_sb_nr <- compareCluster(
  geneClusters_sb, fun = "enrichGO",
  OrgDb = org.Ssalar.eg.db, keyType = "GID",
  ont = "BP", pAdjustMethod = "BH", readable = FALSE)
```

# Trout DevMap

## Build gene clusters

```{r build_gene_clusters_td}
# Trout DevMap: same recipe as salmon DevMap (one gene-list per unit, in order).
constitutive_td <- c(6, 10, 7, 4, 11)
specific_td     <- c(13, 14, 9, 5, 1, 2, 3, 8, 15, 12, 16)
universe        <- rownames(td_norm)

geneClusters_td <- list()
for (stage in specific_td) {
  stage_num <- match(stage, specific_td)
  geneClusters_td[[stage_num]] <- universe[which(td_som$som$unit.classif == stage)]
}
names(geneClusters_td) <- specific_td
```

## Run compareCluster

```{r run_compareCluster_td, cache=TRUE}
compareClusterObj_td <- compareCluster(
  geneClusters_td, fun = "enrichGO",
  OrgDb = org.Omykiss.eg.db, keyType = "GID",
  ont = "BP", pAdjustMethod = "BH", readable = TRUE)

compareClusterObj_td_nr <- compareCluster(
  geneClusters_td, fun = "enrichGO",
  OrgDb = org.Omykiss.eg.db, keyType = "GID",
  ont = "BP", pAdjustMethod = "BH", readable = FALSE)
```

# Trout BodyMap

## Build gene clusters

```{r build_gene_clusters_tb}
# Trout BodyMap: same tissue-grouping recipe as salmon BodyMap above.
gene_class <- tb_som$som$unit.classif
names(gene_class) <- rownames(tb_norm)

tissue_df <- tb_som_info[
  tb_som_info$stage_specific == TRUE &
    tb_som_info$constitutive_class == FALSE,
  c("tissue", "class_number")]

constitutive_classes <- as.numeric(
  tb_som_info[tb_som_info$constitutive_class == TRUE, "class_number"])

geneClusters_tb <- lapply(
  split(tissue_df$class_number, tissue_df$tissue),
  function(cls) names(gene_class)[gene_class %in% cls])

geneClusters_tb[["constitutive"]] <- names(gene_class)[
  gene_class %in% constitutive_classes]

geneClusters_tb <- geneClusters_tb[order(names(geneClusters_tb))]
```

## Run compareCluster

```{r run_compareCluster_tb, cache=TRUE}
compareClusterObj_tb <- compareCluster(
  geneClusters_tb, fun = "enrichGO",
  OrgDb = org.Omykiss.eg.db, keyType = "GID",
  ont = "BP", pAdjustMethod = "BH", readable = TRUE)

compareClusterObj_tb_nr <- compareCluster(
  geneClusters_tb, fun = "enrichGO",
  OrgDb = org.Omykiss.eg.db, keyType = "GID",
  ont = "BP", pAdjustMethod = "BH", readable = FALSE)
```

# Persist compareCluster tables

Writes the readable and raw tables for the four compareCluster objects.
File naming follows the convention used elsewhere in the project so
downstream consumers can swap sources without renaming.

```{r save_compareCluster_csvs}
# Write each compareCluster result to CSV: *_readable (SYMBOL labels) and
# *_raw (gene IDs), for all four species x context objects.
write.csv(as.data.frame(compareClusterObj_sd), "compareCluster_sd_readable.csv", row.names = FALSE)
write.csv(as.data.frame(compareClusterObj_sd_nr), "compareCluster_sd_raw.csv", row.names = FALSE)

write.csv(as.data.frame(compareClusterObj_sb), "compareCluster_sb_readable.csv", row.names = FALSE)
write.csv(as.data.frame(compareClusterObj_sb_nr), "compareCluster_sb_raw.csv", row.names = FALSE)

write.csv(as.data.frame(compareClusterObj_td), "compareCluster_td_readable.csv", row.names = FALSE)
write.csv(as.data.frame(compareClusterObj_td_nr), "compareCluster_td_raw.csv", row.names = FALSE)

write.csv(as.data.frame(compareClusterObj_tb), "compareCluster_tb_readable.csv", row.names = FALSE)
write.csv(as.data.frame(compareClusterObj_tb_nr), "compareCluster_tb_raw.csv", row.names = FALSE)
```
