---
title: "07 — Figure 3: transcriptome divergence (JSD)"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
---

## Overview

Figure 3: expression divergence between paired genes, measured as the square root
of the Jensen-Shannon divergence (JSD) between the two members of a pair, with a
100-sample bootstrap over pairs giving the mean and standard deviation per sample.
Two pairings are compared:

- **Ohnolog pairs** (salmon 4R duplicates, split by Early = AORe / Late = LORe
  rediploidization): panels a (DevMap) and b (BodyMap).
- **Ortholog pairs** (salmon vs trout), split into pairs retained as ohnologs in
  both species (`ortho22`) vs singletons where one copy was lost in the ancestor
  (`ortho11`): panels c (DevMap) and d (BodyMap).

This notebook produces exactly the published items from this analysis: **Figure 3a-d**
and **Supplementary Fig. 29** (the ohnolog-vs-ortholog and AORe-vs-LORe JSD
correlation scatterplots). Nothing else.

Mode handling (see `00_setup.Rmd`):

- **`reproduce` mode** (default): load the deposited per-sample JSD tables
  (`data/fig3/divergence_output.rds`) and re-plot. These are exactly the values behind
  the manuscript figures, so Figure 3 and Supplementary Fig. 29 reproduce bit-for-bit.
  The original bootstrap was unseeded, so the published numbers can only be reproduced
  by loading them.
- **`recompute` mode**: rebuild the JSD tables from the merged TPM matrices created by
  `01_data_loading.Rmd` (`sd_merged`, `td_merged`, `sb_merged`, `tb_merged`) and the
  `orthogroups` table, under `set.seed(FIG3_SEED)`. A `divergence_output_regen.rds` is
  written so the deposited asset is never overwritten.

```{r setup_chain, include=FALSE}
# Packages, constants, paths, helpers and the deposited-asset loaders.
source("R/setup.R")
library(philentropy)   # JSD()
library(gridExtra)     # grid.arrange()
library(ggpubr)        # ggscatter(), stat_cor()
if (!dir.exists("data/fig3")) dir.create("data/fig3", recursive = TRUE)

# Seed for the recompute bootstrap. The published run was unseeded, so this only
# makes re-derivation deterministic; it does not recover the exact published draws,
# which are shipped in data/fig3/divergence_output.rds for reproduce mode.
FIG3_SEED <- 1

# The four pairings that appear in the paper (Fig 3a-d and Supplementary Fig. 29).
PAIRINGS <- c("a22s", "l22s", "ortho11", "ortho22")

# BodyMap tissue plotting order.
TISSUE_LEVELS <- c("Brain", "Testes", "Gill", "DistalIntestine",
                   "HeadKidney", "Ova", "Muscle", "Liver")
```

## Pair selection and the JSD statistic

```{r fig3_functions}
# Select a gene-pair list from the orthogroup table:
#   a22s     salmon AORe (Early) 2:2 ohnolog pairs (two salmon copies per subOG)
#   l22s     salmon LORe (Late)  2:2 ohnolog pairs
#   ortho22  salmon-trout orthologs where both are retained as 2:2 ohnologs
#   ortho11  1:1 salmon-trout orthologs where the pair is a singleton
get_gene_pairs <- function(phyl) {
  if (phyl == "a22s") {
    orthogroups %>% filter(spc == "Ssal" & nSsal == 2 & nOmyk == 2 & phylogeny == "AORe") %>% dplyr::select(1, 3)
  } else if (phyl == "l22s") {
    orthogroups %>% filter(spc == "Ssal" & nSsal == 2 & nOmyk == 2 & phylogeny == "LORe") %>% dplyr::select(1, 3)
  } else if (phyl == "ortho11") {
    orthogroups %>%
      filter(nSsal == 1 & nOmyk == 1 & proxiPhylogeny == "AORe" & reciprocal == "FALSE") %>%
      dplyr::select(geneID, subOG)
  } else if (phyl == "ortho22") {
    validSynteny <- unique(na.omit(orthogroups$proxiSynteny))
    orthogroups %>%
      filter(nSsal == 2 & nOmyk == 2 & phylogeny == "AORe" & synteny %in% validSynteny) %>%
      mutate(AB = strsplit(gsub("[()]", "", synteny), split = ",")) %>%
      rowwise() %>% mutate(AB = match(chr, AB)) %>% ungroup() %>%
      mutate(AB = c("A", "A", "B", "B")[AB]) %>%
      dplyr::select(geneID, subOG)
  } else {
    stop("unknown pairing: ", phyl)
  }
}

# sqrt(JSD) between two expression profiles: quantile-normalise the pair of columns,
# convert each to a probability vector, then Jensen-Shannon divergence.
JSDfun <- function(x1, x2) {
  x <- cbind(x1, x2) %>% as.matrix() %>% preprocessCore::normalize.quantiles()
  x[, 1] <- x[, 1] / sum(x[, 1])
  x[, 2] <- x[, 2] / sum(x[, 2])
  x <- t(x)
  suppressMessages(d <- sqrt(JSD(x, unit = "log")))
  d
}
```

## Compute (recompute) or load (reproduce) the per-sample JSD tables

```{r fig3_divergence}
if (DATA_MODE == "reproduce") {
  # The exact per-sample JSD tables behind Fig 3a-d and Supplementary Fig. 29.
  divergence_output <- readRDS("data/fig3/divergence_output.rds")
  stages <- divergence_output$a22s$devmap$data$stage
} else {
  set.seed(FIG3_SEED)

  # Merge the four TPM matrices built by 01_data_loading.Rmd into one DevMap and one
  # BodyMap matrix (ortholog pairs span both species, so salmon and trout rows share a
  # matrix). BodyMap collapses sex within each tissue x maturity.
  tb_merged$Testes_Mature_Male <- NA
  devmap  <- rbind(sd_merged, td_merged)
  bodymap <- rbind(sb_merged, tb_merged) %>%
    rownames_to_column(var = "geneID") %>%
    pivot_longer(cols = -geneID, names_to = "sample", values_to = "tpm") %>%
    separate(sample, into = c("Tissue", "Maturity", "Sex"), sep = "_") %>%
    unite("Name", Tissue:Maturity, sep = "_") %>%
    group_by(geneID, Name) %>%
    summarise(tpm = mean(tpm, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(values_from = tpm, names_from = Name) %>%
    column_to_rownames(var = "geneID")

  stages  <- colnames(devmap)
  tissues <- colnames(bodymap)

  # One JSD table (mean + SD over 100 bootstraps of the pairs) per sample.
  jsd_table <- function(pairing, expression_data, samples) {
    gene_pairs <- get_gene_pairs(pairing)
    pairs_split <- gene_pairs %>%
      group_by(subOG) %>%
      summarize(geneA = geneID[1], geneB = geneID[2], .groups = "drop")
    n_pairs <- nrow(pairs_split)

    expr <- expression_data %>%
      rownames_to_column(var = "geneID") %>%
      inner_join(gene_pairs, by = "geneID") %>%
      dplyr::select(-subOG)

    out <- data.frame(sample = samples, mean_JSD = 0, sd_JSD = 0)
    for (s in seq_along(samples)) {
      boot <- vector()
      for (i in 1:100) {
        bp <- pairs_split[sample(1:n_pairs, replace = TRUE), ]
        a  <- merge(bp, expr, by.x = "geneA", by.y = "geneID")
        ab <- merge(a, expr, by.x = "geneB", by.y = "geneID", suffixes = c("_A", "_B"))
        cols <- ab[, c(paste(samples[s], "A", sep = "_"), paste(samples[s], "B", sep = "_"))]
        try(boot[i] <- JSDfun(cols[, 1], cols[, 2]), silent = TRUE)
      }
      out$mean_JSD[s] <- mean(boot)
      out$sd_JSD[s]   <- sd(boot)
    }
    out
  }

  divergence_output <- list()
  for (p in PAIRINGS) {
    dev <- jsd_table(p, devmap, stages)
    names(dev)[1] <- "stage"
    bod <- jsd_table(p, bodymap, tissues) %>%
      rename(tissue = sample) %>%
      separate(tissue, into = c("tissue", "maturity"), sep = "_") %>%
      na.omit()
    divergence_output[[p]] <- list(devmap = list(data = dev),
                                   bodymap = list(data = bod))
  }

  # Never overwrite the deposited asset: write a *_regen sibling.
  saveRDS(divergence_output, "data/fig3/divergence_output_regen.rds")
}
```

## Figure 3, panels a-d

```{r fig3_abcd}
dev_ohno <- bind_rows(a22 = divergence_output$a22s$devmap$data,
                      l22 = divergence_output$l22s$devmap$data, .id = "set")
bod_ohno <- bind_rows(a22 = divergence_output$a22s$bodymap$data,
                      l22 = divergence_output$l22s$bodymap$data, .id = "set")
dev_orth <- bind_rows(o22 = divergence_output$ortho22$devmap$data,
                      o11 = divergence_output$ortho11$devmap$data, .id = "set")
bod_orth <- bind_rows(o22 = divergence_output$ortho22$bodymap$data,
                      o11 = divergence_output$ortho11$bodymap$data, .id = "set")

line_panel <- function(dat, cols, ylims) {
  ggplot(dat, aes(x = factor(stage, levels = unique(stage)),
                  y = mean_JSD, group = set, color = set, fill = set)) +
    geom_ribbon(aes(ymin = mean_JSD - sd_JSD, ymax = mean_JSD + sd_JSD, fill = set), alpha = 0.2) +
    geom_line(linewidth = 2, color = "black") + geom_line(linewidth = 1.5) +
    theme_classic() +
    scale_color_manual(values = cols) + scale_fill_manual(values = cols) +
    ylim(ylims) +
    theme(axis.title = element_blank(), axis.text.y = element_text(size = 12),
          legend.position = "none")
}
point_panel <- function(dat, cols, ylims) {
  ggplot(dat, aes(x = factor(tissue, levels = TISSUE_LEVELS),
                  y = mean_JSD, shape = maturity, fill = set)) +
    geom_errorbar(aes(ymin = mean_JSD - sd_JSD, ymax = mean_JSD + sd_JSD),
                  width = 0.2, position = position_dodge(width = 0.4), size = 0.6) +
    geom_point(size = 4, alpha = 0.9, position = position_dodge(width = 0.4)) +
    scale_shape_manual(values = c(24, 22)) +
    scale_color_manual(values = cols) + scale_fill_manual(values = cols) +
    ylim(ylims) + theme_classic() +
    theme(axis.title = element_blank(), axis.text = element_text(size = 12),
          legend.position = "none")
}

ohno_cols <- c(a22 = "firebrick", l22 = "navyblue")   # Early = red, Late = blue
orth_cols <- c(o22 = "firebrick", o11 = "palegreen")  # ohnolog = red, singleton = green

fig3a <- line_panel(dev_ohno, ohno_cols, c(0.15, 0.45))
fig3b <- point_panel(bod_ohno, ohno_cols, c(0, 0.5))
fig3c <- line_panel(dev_orth, orth_cols, c(0.2, 0.37))
fig3d <- point_panel(bod_orth, orth_cols, c(0.2, 0.48))

fig3abcd <- grid.arrange(fig3a, fig3b, fig3c, fig3d, nrow = 2)
ggsave("results/fig3_ABCD_divergence.pdf", fig3abcd, width = 10, height = 7)
```

## Supplementary Fig. 29 (JSD correlation scatterplots)

```{r suppfig29}
# Correlate per-sample mean JSD across pairings. Join on sample (stage or
# tissue x maturity) so rows always align, regardless of dropped tissues.
scatter <- function(xdat, ydat, by, xlab, ylab, title) {
  d <- dplyr::inner_join(
    dplyr::select(xdat, dplyr::all_of(by), x = mean_JSD),
    dplyr::select(ydat, dplyr::all_of(by), y = mean_JSD),
    by = by)
  d$lab <- do.call(paste, c(d[by], sep = "_"))
  ggscatter(d, x = "x", y = "y", label = "lab",
            add = "reg.line", add.params = list(color = "black", fill = "lightgray"),
            conf.int = TRUE, xlab = xlab, ylab = ylab, title = title) +
    stat_cor(method = "pearson")
}

x_lab <- "Early rediploidization\nohnolog pairs"
# Top row: Early (AORe) vs Late (LORe) ohnolog JSD.
s_topL <- scatter(divergence_output$a22s$devmap$data,  divergence_output$l22s$devmap$data,
                  "stage", x_lab, "Late rediploidization\nohnolog pairs", "DevMap")
s_topR <- scatter(divergence_output$a22s$bodymap$data, divergence_output$l22s$bodymap$data,
                  c("tissue", "maturity"), x_lab, "Late rediploidization\nohnolog pairs", "BodyMap")
# Bottom row: ohnolog vs ortholog JSD.
s_botL <- scatter(divergence_output$a22s$devmap$data,  divergence_output$ortho22$devmap$data,
                  "stage", x_lab, "Ortholog pairs\n(Early rediploidization ohnologs)", "DevMap")
s_botR <- scatter(divergence_output$a22s$bodymap$data, divergence_output$ortho22$bodymap$data,
                  c("tissue", "maturity"), x_lab, "Ortholog pairs\n(Early rediploidization ohnologs)", "BodyMap")

suppfig29 <- grid.arrange(s_topL, s_topR, s_botL, s_botR, nrow = 2)
ggsave("results/suppFig29_JSD_correlations.pdf", suppfig29, width = 9, height = 8)
```
