---
title: "99a — Build Atlantic salmon OrgDB"
author: "AQUA-FAANG salmonid analysis"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    cache: true
---


## Overview

Builds the `org.Ssalar.eg.db` annotation package that the GO enrichment
notebook (`06_go.Rmd`) consumes. Ensembl BioMart provides gene IDs, GO
mappings, gene symbols, descriptions, and chromosomes; AnnotationForge
turns them into an installable OrgDB. Run once per Ensembl release.

## Packages

```{r r_packages, message=FALSE, warning=FALSE}
library(AnnotationForge)  # makeOrgPackage()
library(biomaRt)          # getBM() Ensembl query
library(tidyverse)        # table wrangling
```

## Fetch annotation from BioMart

One query for everything; we split into the three tables AnnotationForge
needs in the next chunk.

```{r fetch_biomart, cache=TRUE}
# One BioMart query pulling gene IDs, GO terms + evidence, symbols,
# descriptions and chromosomes for the salmon Ensembl dataset.
tbl <- getBM(
  attributes = c("ensembl_gene_id", "go_id", "go_linkage_type",
                 "external_gene_name", "description", "chromosome_name"),
  mart = useDataset(dataset = "ssalar_gene_ensembl",
                    mart = useMart("ENSEMBL_MART_ENSEMBL")))
```

## Split into the three AnnotationForge tables

The gene-info table requires non-empty SYMBOL and GENENAME, the GO table
requires non-empty GO and evidence; both filters are applied here so the
package builder does not silently drop rows.

```{r split_tables}
# Gene-info table (GID -> SYMBOL, GENENAME); drop rows with empty symbol/name.
gene_info_tbl <- tbl %>%
  dplyr::select(GID = ensembl_gene_id,
                SYMBOL = external_gene_name,
                GENENAME = description) %>%
  distinct() %>%
  filter(SYMBOL != "", GENENAME != "")

# Gene -> chromosome table.
gene_chromosome_tbl <- tbl %>%
  distinct(GID = ensembl_gene_id, CHROMOSOME = chromosome_name)

# Gene -> GO table with evidence; drop rows with empty GO/evidence.
gene_GO_tbl <- tbl %>%
  distinct(GID = ensembl_gene_id, GO = go_id, EVIDENCE = "go_linkage_type") %>%
  filter(GO != "", EVIDENCE != "")
```

## Build the OrgDB package

```{r build_orgdb}
# Assemble the three tables into an installable OrgDb source package
# (org.Ssalar.eg.db) in the working directory. tax_id 8030 = Salmo salar.
AnnotationForge::makeOrgPackage(
  gene_info  = gene_info_tbl,
  chromosome = gene_chromosome_tbl,
  go         = gene_GO_tbl,
  version    = "0.1",
  maintainer = "Some One <so@someplace.org>",
  author     = "Some One <so@someplace.org>",
  outputDir  = ".",
  tax_id     = "8030",
  genus      = "Salmo",
  species    = "salar",
  goTable    = "go")
```

## Install the built package

```{r install_orgdb}
# Install the freshly-built package into the R library so 06_go.Rmd can load it.
install.packages("./org.Ssalar.eg.db", repos = NULL, type = "source")
```
