---
title: "Data loading and Normalization"
author: Diego Perojil Morata
date: "Timestamp: `r format(Sys.time())`"
output:
  html_document:
    df_print: paged
    toc: yes
    number_sections: true
---

# Install required packages
```{r,include=FALSE}
install.packages("tidyverse")
install.packages("kohonen")
install.packages("data.table")
install.packages("foreach")
```

# Load required libraries
```{r,include=FALSE}
library(tidyverse)
library(kohonen)
library(data.table)
library(foreach)
```

# Global variables
```{r}
stage_names <- c("LC", # late cleavage
                 "EB","MB","LB", # early, mid and late blastulation
                 "EG","MG","LG", # early, mid, and late gastrulation
                 "ES","EMS","MS","MLS","LS", # early to late somitogenesis
                 "EP","LP") # early and late pharyngula

```

# Orthogroup table

```{r,include=FALSE}
orthogroups <- read_tsv(file = "../data/orthogroups/SalmonTroutOrthologs.tsv")
```

# Functions

```{r}
source("functions/Data_Loading_and_Normalization_Functions.R")
```

# Load TPM replicate data

```{r}
# Devmap
sd_raw <- # salmon
  read_tsv("../data/devmap/salmon/rsem.merged.gene_tpm.tsv", show_col_types = F)

td_raw <- # trout
  read_tsv("../data/devmap/trout/rsem.merged.gene_tpm.tsv", show_col_types = F)

# Bodymap
sb_raw <- process_tissue_data("salmon") # Salmon
tb_raw <- process_tissue_data("trout") # Trout
```

# Merge replicates

```{r}
# Devmap
sd_merged <- merge_reps_devmap(sd_raw) # Salmon
td_merged <- merge_reps_devmap(td_raw) # Trout

#Bodymap
sb_merged <- merge_reps_bodymap(sb_raw) # Salmon
tb_merged <- merge_reps_bodymap(tb_raw) # Trout
```

# Normalize expression data
```{r}

load("../data/original_norm_files/norm.RData")

# Original code below

# # Devmap
# sd_norm <- normalize(sd_merged) # Salmon
# td_norm <- normalize(td_merged) # Trout
# 
# # Bodymap
# sb_norm <- normalize(sb_merged) # Salmon
# tb_norm <- normalize(tb_merged) # Trout
```

# Clustering using self organizing maps. Original clustering performed in the following system:

```
platform       x86_64-w64-mingw32          
arch           x86_64                      
os             mingw32                    
system         x86_64, mingw32            
status                                    
major          4                          
minor          1.3                        
year           2022                        
month          03                          
day            10                          
svn rev        81868                      
language       R                          
version.string R version 4.1.3 (2022-03-10)
nickname       One Push-Up
```

```{r}


indices_order_ssal <- c(1,2,3,4,10,
                   12,9,11,21,22,23,24,17,18,
                   19,20,13,14,15,16,5,6,
                   7,8,25,26,27,28)

indices_order_omyk <- c(1,2,3,4,10,9,11,
                        20,21,22,23,16,
                        17,18,19,12,13,
                        14,15,5,6,7,8,24,25,26,27)
# # Devmap
# set.seed(387334)
# sd_som <- get_som(sd_norm,4,4) # Salmon
# set.seed(387334)
# td_som <- get_som(td_norm,4,4) # Trout
# 
# # Bodymap
# set.seed(387334)
# sb_som <- get_som_bodymap(sb_norm,6,6,order_indices = indices_order_ssal) # Salmon
# set.seed(387334)
# tb_som <- get_som_bodymap(tb_norm,6,6,order_indices = indices_order_omyk) # Trout
```

# Load original SOM clustering files

```{r}
load(file = "../data/original_soms/soms.RData")
lev <- levels(sb_som$stat$stage)
lev_order <- lev[c(3:6,7:8,1:2,9:28)]
sb_som$stat$stage <- factor(sb_som$stat$stage,levels = lev_order)
sb_som$stat <- arrange(sb_som$stat,class,stage)

lev <- levels(tb_som$stat$stage)
lev_order <- lev[c(3:6,7,1:2,8:28)]
tb_som$stat$stage <- factor(tb_som$stat$stage,levels = lev_order)
tb_som$stat <- arrange(tb_som$stat,class,stage)
```

# Get gene ID's of each SOM class and max expression stage/tissue
```{r}
# Devmap
sd_ids <- som_ids(sd_norm,sd_som) # List with gene ID's for each SOM class
sd_som_info <- som_file_info_d(sd_som)
td_ids <- som_ids(td_norm,td_som)
td_som_info <- som_file_info_d(td_som)

# Bodymap
sb_ids <- som_ids(sb_norm,sb_som)
sb_som_info <- som_file_info_b(sb_som)
tb_ids <- som_ids(tb_norm,tb_som)
tb_som_info <- som_file_info_b(tb_som)

```








