---
title: "Make_OrgDB"
author: "Lars Gronvold"
date: "`r Sys.Date()`"
output: html_document
---


```{r}
library(AnnotationForge)
library(biomaRt)
library(tidyverse)
```

Fetch annotation from biomart
gene_info_tbl <-

```{r}
tbl <- getBM(attributes = c("ensembl_gene_id","go_id","go_linkage_type","external_gene_name",
                            "description","chromosome_name"),
             mart = useDataset(dataset = "ssalar_gene_ensembl", 
                               mart = useMart("ENSEMBL_MART_ENSEMBL")))
```

Split into separate tables

```{r}

  gene_info_tbl <- tbl %>%
  dplyr::select(GID=ensembl_gene_id,SYMBOL=external_gene_name,GENENAME=description) %>% 
  distinct() %>% 
  filter( SYMBOL != "", GENENAME != "")


gene_chromosome_tbl <-
  tbl %>% distinct(GID=ensembl_gene_id, CHROMOSOME=chromosome_name)

gene_GO_tbl <-
  tbl %>% 
  distinct(GID=ensembl_gene_id, GO=go_id, EVIDENCE="go_linkage_type") %>% 
  filter( GO != "", EVIDENCE != "")

```

Turn into OrgDB package

```{r}

AnnotationForge::makeOrgPackage(gene_info=gene_info_tbl, chromosome=gene_chromosome_tbl, go=gene_GO_tbl,
               version="0.1",
               maintainer="Some One <so@someplace.org>",
               author="Some One <so@someplace.org>",
               outputDir = ".",
               tax_id="8030",
               genus="Salmo",
               species="salar",
               goTable="go")
```

Install package 

```{r}
install.packages("./org.Ssalar.eg.db", repos=NULL,type = "source")
```




