---
title: "Make_OrgDB"
author: "Lars Gronvold"
date: "`r Sys.Date()`"
output: html_document
---


```{r}
library(AnnotationForge)
library(biomaRt)
library(tidyverse)
```

Fetch annotation from biomart
gene_info_tbl <-

```{r}
tbl_trout <- getBM(attributes = c("ensembl_gene_id","go_id","go_linkage_type","external_gene_name",
                            "description","chromosome_name"),
             mart = useDataset(dataset = "omykiss_gene_ensembl", 
                               mart = useMart("ENSEMBL_MART_ENSEMBL")))


```

Split into separate tables

```{r}

  gene_info_tbl_trout <- tbl_trout %>%
  dplyr::select(GID=ensembl_gene_id,SYMBOL=external_gene_name,GENENAME=description) %>% 
  distinct() %>% 
  filter( SYMBOL != "", GENENAME != "")


gene_chromosome_tbl_trout <-
  tbl_trout %>% distinct(GID=ensembl_gene_id, CHROMOSOME=chromosome_name)

gene_GO_tbl_trout <-
  tbl_trout %>% 
  distinct(GID=ensembl_gene_id, GO=go_id, EVIDENCE="go_linkage_type") %>% 
  filter( GO != "", EVIDENCE != "")

```

Turn into OrgDB package

```{r}

AnnotationForge::makeOrgPackage(
  gene_info   = gene_info_tbl_trout,        # data.frame with columns: gene_id, symbol, gene_name
  chromosome  = gene_chromosome_tbl_trout,  # gene_id, chromosome, start, end  (optional strand, etc.)
  go          = gene_GO_tbl_trout,          # gene_id, go_id, evidence
  version     = "0.1",
  maintainer  = "Some One <so@someplace.org>",
  author      = "Some One <so@someplace.org>",
  outputDir   = ".",
  tax_id      = "8022",               # rainbow trout
  genus       = "Oncorhynchus",
  species     = "mykiss",
  goTable     = "go"
)

```

Install package 

```{r}
install.packages("./org.Omykiss.eg.db", repos=NULL,type = "source")
```




