---
title: "Observed to expected matrix"
output: html_notebook
---

# Get SOM cluster IDs (Already in data loading and normalization)
```{r}
sd_ids <- som_ids(sd_norm,sd_som) # List with gene ID's for each SOM class
sd_som_info <- som_file_info_d(sd_som)
td_ids <- som_ids(td_norm,td_som)
td_som_info <- som_file_info_d(td_som)

# Bodymap
sb_ids <- som_ids(sb_norm,sb_som)
sb_som_info <- som_file_info_b(sb_som)
tb_ids <- som_ids(tb_norm,tb_som)
tb_som_info <- som_file_info_b(tb_som)
```

# Get SOM orthogroup IDs
```{r}
sd_ids_som <- list()
td_ids_som <- list()

for (i in 1:16) {
  sd_ids_som[[i]] <- orthogroups$subOG[which(orthogroups$geneID %in% sd_ids[[i]])]
  td_ids_som[[i]] <- orthogroups$subOG[which(orthogroups$geneID %in% td_ids[[i]])]
}

# Bodymap 
sb_ids_som <- list()
tb_ids_som <- list()

for (i in 1:36) {
  sb_ids_som[[i]] <- orthogroups$subOG[which(orthogroups$geneID %in% sb_ids[[i]])]
  tb_ids_som[[i]] <- orthogroups$subOG[which(orthogroups$geneID %in% tb_ids[[i]])]
}

```

# Filter many copy genes from SOM orthogroup IDs

```{r}
sd_ids_som_filtered <- list()
td_ids_som_filtered <- list()

for (i in 1:16) {
  sd_ids_som_filtered[[i]] <- 
    orthogroups$subOG[which(orthogroups$geneID %in% sd_ids[[i]] & orthogroups$nSsal < 3 & orthogroups$nOmyk < 3)]
  td_ids_som_filtered[[i]] <-
    orthogroups$subOG[which(orthogroups$geneID %in% td_ids[[i]] & orthogroups$nSsal < 3 & orthogroups$nOmyk < 3)]
}

# Bodymap

sb_ids_som_filtered <- list()
tb_ids_som_filtered <- list()

for (i in 1:36) {
  sb_ids_som_filtered[[i]] <- 
    orthogroups$subOG[which(orthogroups$geneID %in% sb_ids[[i]] & orthogroups$nSsal < 3 & orthogroups$nOmyk < 3)]
  tb_ids_som_filtered[[i]] <-
    orthogroups$subOG[which(orthogroups$geneID %in% tb_ids[[i]] & orthogroups$nSsal < 3 & orthogroups$nOmyk < 3)]
}
```

# Calculate initial observed values

```{r}
o_t <- data.frame(matrix(nrow = 16, ncol = 16))
rownames(o_t) <- c(1:16)
colnames(o_t) <- c(1:16)

for (i in 1:16) {
  for (j in 1:16) {
    o_t[i,j] <- length(sd_ids_som[[i]][which(sd_ids_som[[i]] %in% td_ids_som[[j]])])
  }
}

# Bodymap

o_t.b <- data.frame(matrix(nrow = 36, ncol = 36))
rownames(o_t.b) <- c(1:36)
colnames(o_t.b) <- c(1:36)

for (i in 1:36) {
  for (j in 1:36) {
    o_t.b[i,j] <- length(sb_ids_som[[i]][which(sb_ids_som[[i]] %in% tb_ids_som[[j]])])
  }
}
```

# Generate 10k random gene ID to SOM ID assignments

```{r}

td_norm_rownames <- as.data.frame(rownames(td_norm))
colnames(td_norm_rownames) <- c("geneID")
tenk_list <- list()

for (i in 1:10000) {
  tenk_list[[i]] <- as.data.frame(sample(td_norm_rownames$geneID))
  colnames(tenk_list[[i]]) <- c("geneID")
  tenk_list[[i]] <- left_join(tenk_list[[i]],orthogroups[,c(1,3)], by = c("geneID" = "geneID"))
  tenk_list[[i]]$somID <- td_som$som$unit.classif
  tenk_list[[i]] <- tenk_list[[i]] %>% na.omit()
  print(i)
}

# Bodymap

tb_norm_rownames <- as.data.frame(rownames(tb_norm))
colnames(tb_norm_rownames) <- c("geneID")
tenk_list.b <- list()

for (i in 1:10000) {
  tenk_list.b[[i]] <- as.data.frame(sample(tb_norm_rownames$geneID))
  colnames(tenk_list.b[[i]]) <- c("geneID")
  tenk_list.b[[i]] <- dplyr::left_join(tenk_list.b[[i]],orthogroups[,c(1,3)], by = c("geneID" = "geneID"))
  tenk_list.b[[i]]$somID <- tb_som$som$unit.classif
  tenk_list.b[[i]] <- tenk_list.b[[i]] %>% na.omit()
  print(i)
}
```

# Get observations from random subsets

```{r}

random_observations <- data.frame(matrix(nrow = 256, ncol = 10000))
salnames <- paste(c(rep("S",16)), c(rep(1:16,each=16)),sep = "")
omnames <- paste(c(rep("T",16)), c(rep(1:16,16)),sep = "")
rownames_random_obs <- paste(salnames,omnames,sep = "x")

rownames(random_observations) <- rownames_random_obs

for (i in 1:10000) {
  for (j in 1:16) {
    for (l in 1:16) {
      salmon_som_orthos <- sd_ids_som[[j]]
      trout_som_orthos <- tenk_list[[i]]$subOG[which(tenk_list[[i]]$somID == l)]
      obs <- sum(salmon_som_orthos %in% trout_som_orthos)
      random_observations[16*(j-1)+l,i] <- obs
        
    }
  }
  print(i)
}

# Bodymaps

random_observations.b <- data.frame(matrix(nrow = 1296, ncol = 10000))
salnames.b <- paste(c(rep("S",36)), c(rep(1:36,each=36)),sep = "")
omnames.b <- paste(c(rep("T",36)), c(rep(1:36,36)),sep = "")
rownames_random_obs.b <- paste(salnames.b,omnames.b,sep = "x")

rownames(random_observations.b) <- rownames_random_obs.b

for (i in 1:10000) {
  for (j in 1:36) {
    for (l in 1:36) {
      salmon_som_orthos.b <- sb_ids_som[[j]]
      trout_som_orthos.b <- tenk_list.b[[i]]$subOG[which(tenk_list.b[[i]]$somID == l)]
      obs.b <- sum(salmon_som_orthos.b %in% trout_som_orthos.b)
      random_observations.b[36*(j-1)+l,i] <- obs.b
        
    }
  }
  print(i)
}
```

Populate matrix of inferred p_values

```{r}

pvals_mat_log <- data.frame(matrix(nrow = 16, ncol = 16))

for (i in 1:16) {
  for (j in 1:16) {
    pvals_mat_log[i,j] <- -log10(0.000001 + length(which(random_observations[16*(i-1)+j,] > o_t[i,j])) / 10000)
  }
}

# Bodymap

pvals_mat_log.b <- data.frame(matrix(nrow = 36, ncol = 36))

for (i in 1:36) {
  for (j in 1:36) {
    pvals_mat_log.b[i,j] <- -log10(0.000001 + length(which(random_observations.b[36*(i-1)+j,] > o_t.b[i,j])) / 10000)
  }
}


```

Populate matrix of observed/expected ratio

```{r}
oe_mat <- data.frame(matrix(nrow = 16, ncol = 16))
random_obs_mean <- rowMeans(random_observations)
rand_obs_wide <- matrix(random_obs_mean, nrow=16, ncol=16, byrow=TRUE)


for (i in 1:16) {
  for (j in 1:16) {
    oe_mat[i,j] <- o_t[i,j] / rand_obs_wide[i,j]
  }
}

# Bodymap

oe_mat.b <- data.frame(matrix(nrow = 36, ncol = 36))
random_obs_mean.b <- rowMeans(random_observations.b)
rand_obs_wide.b <- matrix(random_obs_mean.b, nrow=36, ncol=36, byrow=TRUE)


for (i in 1:36) {
  for (j in 1:36) {
    oe_mat.b[i,j] <- o_t.b[i,j] / rand_obs_wide.b[i,j]
  }
}
```

Reorder pval matrix to reflect som cluster order
```{r}

# pvals_mat_order <- pvals_mat[c(13,10,15,14,16,12,4,2,1,5,9,6,7,11,8,3),
#                              c(13,14,9,5,1,2,3,8,16,12,15,6,10,7,4,11)]

pvals_mat_order_log <- pvals_mat_log[c(13,10,15,14,16,12,4,2,1,5,9,6,7,11,8,3),
                             c(13,14,9,5,1,2,3,8,16,12,15,6,10,7,4,11)]

# bodymap

pvals_mat_order_log.b <- pvals_mat_log.b[as.numeric(sb_som_info$class_number),
                             as.numeric(tb_som_info$class_number)]
```

Reorder o/e ratio matrix to reflect som cluster order

```{r}
oe_mat <- oe_mat[c(13,10,15,14,16,12,4,2,1,9,6,5,7,11,8,3),
                             c(13,14,9,5,1,2,3,8,15,12,16,6,10,7,4,11)]

# Bodymap

oe_mat.b <- oe_mat.b[as.numeric(sb_som_info$class_number),
                             as.numeric(tb_som_info$class_number)]
```


```{r}
# test <- Heatmap(as.matrix(pvals_mat_order), name = "Empirical p-val",
#                 cluster_column_slices = F,
#                 cluster_row_slices = F,
#                 show_row_dend = F,
#                 show_row_names = T,
#                 col = colorRamp2(0:1,c("red","blue")),
#                 show_column_names = T,
#                 show_column_dend = F,
#                 use_raster = T,
#                 raster_by_magick = T,
#                 cluster_columns = F,
#                 cluster_rows = F,rect_gp = gpar(col = "white", lwd = 2))

pval_heatmap <- Heatmap(as.matrix(pvals_mat_order_log), name = "Empirical p-val",
                cluster_column_slices = F,
                cluster_row_slices = F,
                show_row_dend = F,
                show_row_names = T,
                col = colorRamp2(c(1,0.5,0),c("red","yellow","blue")),
                show_column_names = T,
                show_column_dend = F,
                use_raster = T,
                raster_by_magick = T,
                cluster_columns = F,
                cluster_rows = F,rect_gp = gpar(col = "white", lwd = 2))

o_e_heatmap <- Heatmap(log2(as.matrix(oe_mat)+0.1), name = "o/e ratio",
                cluster_column_slices = F,
                cluster_row_slices = F,
                show_row_dend = F,
                show_row_names = T,
                col = colorRamp2(c(-3,0,3),c("skyblue","seashell","lightcoral")),
                show_column_names = T,
                show_column_dend = F,
                use_raster = T,
                raster_by_magick = T,
                cluster_columns = F,
                cluster_rows = F,rect_gp = gpar(col = "white", lwd = 2))

o_e_heatmap_specific <- Heatmap(log2(as.matrix(oe_mat[c(1:9,11,12),c(1:8,10,11)])+0.1), name = "o/e ratio",
                cluster_column_slices = F,
                cluster_row_slices = F,
                show_row_dend = F,
                show_row_names = T,
                col = colorRamp2(c(-3,0,3),c("skyblue","seashell","lightcoral")),
                show_column_names = T,
                show_column_dend = F,
                use_raster = T,
                raster_by_magick = T,
                cluster_columns = F,
                cluster_rows = F,rect_gp = gpar(col = "white", lwd = 2))

# test2 <- Heatmap(pvals_adjust, name = "Empirical p-val",
#                 cluster_column_slices = F,
#                 cluster_row_slices = F,
#                 show_row_dend = F,
#                 show_row_names = T,
#                 col = colorRamp2(0:1,c("red","blue")),
#                 show_column_names = T,
#                 show_column_dend = F,
#                 use_raster = T,
#                 raster_by_magick = T,
#                 cluster_columns = F,
#                 cluster_rows = F,rect_gp = gpar(col = "white", lwd = 2))
  
```
# Bodymap heatmap

```{r}
o_e_heatmap.b <- Heatmap(log2(as.matrix(oe_mat.b)+0.1), name = "o/e ratio",
                cluster_column_slices = F,
                cluster_row_slices = F,
                show_row_dend = F,
                show_row_names = T,
                col = colorRamp2(c(-3,0,3),c("skyblue","seashell","lightcoral")),
                show_column_names = T,
                show_column_dend = F,
                use_raster = T,
                raster_by_magick = T,
                cluster_columns = F,
                cluster_rows = F,rect_gp = gpar(col = "white", lwd = 2))


o_e_heatmap_specific.b <- Heatmap(log2(as.matrix(oe_mat.b[c(1:20),c(1:21)])+0.1), name = "o/e ratio",
                cluster_column_slices = F,
                cluster_row_slices = F,
                show_row_dend = F,
                show_row_names = T,
                col = colorRamp2(c(-3,0,3),c("skyblue","seashell","lightcoral")),
                show_column_names = T,
                show_column_dend = F,
                use_raster = T,
                raster_by_magick = T,
                cluster_columns = F,
                cluster_rows = F,rect_gp = gpar(col = "white", lwd = 2))


```


# reorder and trim real observations matrix for sankey plot
```{r}

o_st <- data.frame(matrix(nrow = 16, ncol = 16))
rownames(o_st) <- c(1:16)
colnames(o_st) <- c(1:16)

for (i in 1:16) {
  for (j in 1:16) {
    o_st[i,j] <- length(sd_ids_som[[i]][which(sd_ids_som[[i]] %in% td_ids_som[[j]])])
  }
}

o_st_order <- o_st[c(13,10,15,14,16,12,4,2,1,9,6,5,7,11,8,3),
                             c(13,14,9,5,1,2,3,8,15,12,16,6,10,7,4,11)]

o_st_order_trim <- o_st_order[c(1:9,11,12),c(1:8,10,11)]

salmon_sum <- rowSums(o_st_order_trim)
trout_sum <- colSums(o_st_order_trim)



o_st_long <- pivot_longer(o_st_order_trim,names_to = "trout_clusters",values_to = "shared orthologues",cols = everything())

o_st_long$salmon_clusters <- rep(c(13,10,15,14,16,12,4,2,1,6,5),each = 10)



# Bodymap 

o_st.b <- data.frame(matrix(nrow = 36, ncol = 36))
rownames(o_st.b) <- c(1:36)
colnames(o_st.b) <- c(1:36)

for (i in 1:36) {
  for (j in 1:36) {
    o_st.b[i,j] <- length(sb_ids_som[[i]][which(sb_ids_som[[i]] %in% tb_ids_som[[j]])])
  }
}

o_st_order.b <- o_st.b[as.numeric(sb_som_info$class_number),
                             as.numeric(tb_som_info$class_number)]

o_st_order_trim.b <- o_st_order.b[c(1:20),c(1:21)]

salmon_sum.b <- rowSums(o_st_order_trim.b)
trout_sum.b <- colSums(o_st_order_trim.b)



o_st_long.b <- pivot_longer(o_st_order_trim.b,names_to = "trout_clusters",values_to = "shared orthologues",cols = everything())

o_st_long.b$salmon_clusters <- rep(as.numeric(rownames(o_st_order_trim.b)),each = length(o_st_order_trim.b))
```

```{r}
library(ggalluvial)
library(scales)

data <- data.frame(
  source = factor(paste("s",o_st_long$salmon_clusters,sep = ""),
                  levels = c("s13","s10","s15","s14",
                             "s16","s12","s4","s2",
                             "s1","s6","s5")),
  target = factor(paste("t",o_st_long$trout_clusters,sep = ""),
                  levels = c("t13","t14","t9","t5",
                             "t1","t2","t3","t8",
                             "t12","t16")),
  count = o_st_long$`shared orthologues`,
  color = oe_mat_specific_long$o_eratio)

# Create color palette function
color_palette <- colorRampPalette(c("skyblue", "seashell", "lightcoral"))

# Rescale values to [0, 1]
rescaled_values <- rescale(data$color, to = c(0, 1))

# Apply color palette to rescaled values
data$colors <- color_palette(nrow(data))[as.integer(round(rescaled_values * 99)) + 1]

ggplot(data = data,
       aes(axis1 = source, axis2 = target, y = count)) +
  geom_alluvium(aes(fill = I(colors)),alpha = 0.7) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal()


# Bodymap
oe_mat_specific.b <- log2(as.matrix(oe_mat.b[c(1:20),c(1:21)])+0.1)
oe_mat_specific_long.b <- pivot_longer(as.data.frame(oe_mat_specific.b), everything(),names_to = "trout_cluster",values_to = "o_eratio")

data.b <- data.frame(
  source = factor(paste("s",o_st_long.b$salmon_clusters,sep = ""),
                  levels = unique(paste("s",o_st_long.b$salmon_clusters,sep = ""))),
  target = factor(paste("t",o_st_long.b$trout_clusters,sep = ""),
                  levels = unique(paste("t",o_st_long.b$trout_clusters,sep = ""))),
  count = o_st_long.b$`shared orthologues`,
  color = oe_mat_specific_long.b$o_eratio)

# Create color palette function
color_palette.b <- colorRampPalette(c("skyblue", "seashell", "lightcoral"))

# Rescale values to [0, 1]
rescaled_values.b <- rescale(data.b$color, to = c(0, 1))

# Apply color palette to rescaled values
data.b$colors <- color_palette.b(nrow(data.b))[as.integer(round(rescaled_values.b * nrow(data.b))) + 1]

ggplot(data = data.b,
       aes(axis1 = source, axis2 = target, y = count)) +
  geom_alluvium(aes(fill = I(colors)),alpha = 0.7) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal()
```

```{r}
"#87CEEB" "#87CEEB" "#88CEEB" "#88CEEB" "#89CEEB" "#89CEEB" "#8ACFEB" "#8BCFEB" "#8BCFEB" "#8CCFEB" "#8CCFEB" "#8DD0EB" "#8DD0EB" "#8ED0EB" "#8FD0EB"
 [16] "#8FD0EB" "#90D0EB" "#90D1EB" "#91D1EB" "#91D1EB" "#92D1EB" "#93D1EB" "#93D2EB" "#94D2EB" "#94D2EB" "#95D2EB" "#95D2EB" "#96D3EB" "#97D3EB" "#97D3EB"
 [31] "#98D3EB" "#98D3EB" "#99D3EB" "#99D4EB" "#9AD4EB" "#9BD4EB" "#9BD4EB" "#9CD4EB" "#9CD5EB" "#9DD5EB" "#9DD5EB" "#9ED5EB" "#9FD5EB" "#9FD6EB" "#A0D6EB"
 [46] "#A0D6EB" "#A1D6EB" "#A1D6EB" "#A2D6EB" "#A3D7EB" "#A3D7EB" "#A4D7EB" "#A4D7EB" "#A5D7EB" "#A5D8EB" "#A6D8EB" "#A7D8EB" "#A7D8EB" "#A8D8EB" "#A8D8EB"
 [61] "#A9D9EB" "#A9D9EB" "#AAD9EB" "#ABD9EB" "#ABD9EB" "#ACDAEB" "#ACDAEB" "#ADDAEB" "#ADDAEB" "#AEDAEB" "#AFDBEC" "#AFDBEC" "#B0DBEC" "#B0DBEC" "#B1DBEC"
 [76] "#B1DBEC" "#B2DCEC" "#B3DCEC" "#B3DCEC" "#B4DCEC" "#B4DCEC" "#B5DDEC" "#B5DDEC" "#B6DDEC" "#B7DDEC" "#B7DDEC" "#B8DEEC" "#B8DEEC" "#B9DEEC" "#B9DEEC"
 [91] "#BADEEC" "#BBDEEC" "#BBDFEC" "#BCDFEC" "#BCDFEC" "#BDDFEC" "#BDDFEC" "#BEE0EC" "#BFE0EC" "#BFE0EC" "#C0E0EC" "#C0E0EC" "#C1E0EC" "#C1E1EC" "#C2E1EC"
[106] "#C3E1EC" "#C3E1EC" "#C4E1EC" "#C4E2EC" "#C5E2EC" "#C6E2EC" "#C6E2EC" "#C7E2EC" "#C7E3EC" "#C8E3EC" "#C8E3EC" "#C9E3EC" "#CAE3EC" "#CAE3EC" "#CBE4EC"
[121] "#CBE4EC" "#CCE4EC" "#CCE4EC" "#CDE4EC" "#CEE5EC" "#CEE5EC" "#CFE5EC" "#CFE5EC" "#D0E5EC" "#D0E6EC" "#D1E6EC" "#D2E6EC" "#D2E6EC" "#D3E6EC" "#D3E6EC"
[136] "#D4E7EC" "#D4E7EC" "#D5E7EC" "#D6E7EC" "#D6E7EC" "#D7E8ED" "#D7E8ED" "#D8E8ED" "#D8E8ED" "#D9E8ED" "#DAE8ED" "#DAE9ED" "#DBE9ED" "#DBE9ED" "#DCE9ED"
[151] "#DCE9ED" "#DDEAED" "#DEEAED" "#DEEAED" "#DFEAED" "#DFEAED" "#E0EBED" "#E0EBED" "#E1EBED" "#E2EBED" "#E2EBED" "#E3EBED" "#E3ECED" "#E4ECED" "#E4ECED"
[166] "#E5ECED" "#E6ECED" "#E6EDED" "#E7EDED" "#E7EDED" "#E8EDED" "#E8EDED" "#E9EEED" "#EAEEED" "#EAEEED" "#EBEEED" "#EBEEED" "#ECEEED" "#ECEFED" "#EDEFED"
[181] "#EEEFED" "#EEEFED" "#EFEFED" "#EFF0ED" "#F0F0ED" "#F0F0ED" "#F1F0ED" "#F2F0ED" "#F2F0ED" "#F3F1ED" "#F3F1ED" "#F4F1ED" "#F4F1ED" "#F5F1ED" "#F6F2ED"
[196] "#F6F2ED" "#F7F2ED" "#F7F2ED" "#F8F2ED" "#F8F3ED" "#F9F3ED" "#FAF3ED" "#FAF3ED" "#FBF3ED" "#FBF3ED" "#FCF4ED" "#FCF4ED" "#FDF4ED" "#FEF4ED" "#FEF4ED"
[211] "#FEF4ED" "#FEF4ED" "#FEF3EC" "#FEF3EC" "#FEF2EB" "#FEF1EB" "#FEF1EA" "#FEF0EA" "#FEF0E9" "#FEEFE9" "#FEEFE8" "#FEEEE7" "#FEEEE7" "#FEEDE6" "#FDECE6"
[226] "#FDECE5" "#FDEBE5" "#FDEBE4" "#FDEAE4" "#FDEAE3" "#FDE9E3" "#FDE8E2" "#FDE8E2" "#FDE7E1" "#FDE7E1" "#FDE6E0" "#FDE6E0" "#FDE5DF" "#FCE5DF" "#FCE4DE"
[241] "#FCE3DD" "#FCE3DD" "#FCE2DC" "#FCE2DC" "#FCE1DB" "#FCE1DB" "#FCE0DA" "#FCE0DA" "#FCDFD9" "#FCDED9" "#FCDED8" "#FCDDD8" "#FBDDD7" "#FBDCD7" "#FBDCD6"
[256] "#FBDBD6" "#FBDBD5" "#FBDAD5" "#FBD9D4" "#FBD9D4" "#FBD8D3" "#FBD8D2" "#FBD7D2" "#FBD7D1" "#FBD6D1" "#FBD6D0" "#FAD5D0" "#FAD4CF" "#FAD4CF" "#FAD3CE"
[271] "#FAD3CE" "#FAD2CD" "#FAD2CD" "#FAD1CC" "#FAD0CC" "#FAD0CB" "#FACFCB" "#FACFCA" "#FACECA" "#FACEC9" "#F9CDC8" "#F9CDC8" "#F9CCC7" "#F9CBC7" "#F9CBC6"
[286] "#F9CAC6" "#F9CAC5" "#F9C9C5" "#F9C9C4" "#F9C8C4" "#F9C8C3" "#F9C7C3" "#F9C6C2" "#F9C6C2" "#F8C5C1" "#F8C5C1" "#F8C4C0" "#F8C4C0" "#F8C3BF" "#F8C3BF"
[301] "#F8C2BE" "#F8C1BD" "#F8C1BD" "#F8C0BC" "#F8C0BC" "#F8BFBB" "#F8BFBB" "#F8BEBA" "#F7BDBA" "#F7BDB9" "#F7BCB9" "#F7BCB8" "#F7BBB8" "#F7BBB7" "#F7BAB7"
[316] "#F7BAB6" "#F7B9B6" "#F7B8B5" "#F7B8B5" "#F7B7B4" "#F7B7B3" "#F7B6B3" "#F6B6B2" "#F6B5B2" "#F6B5B1" "#F6B4B1" "#F6B3B0" "#F6B3B0" "#F6B2AF" "#F6B2AF"
[331] "#F6B1AE" "#F6B1AE" "#F6B0AD" "#F6B0AD" "#F6AFAC" "#F6AEAC" "#F5AEAB" "#F5ADAB" "#F5ADAA" "#F5ACAA" "#F5ACA9" "#F5ABA8" "#F5ABA8" "#F5AAA7" "#F5A9A7"
[346] "#F5A9A6" "#F5A8A6" "#F5A8A5" "#F5A7A5" "#F5A7A4" "#F4A6A4" "#F4A5A3" "#F4A5A3" "#F4A4A2" "#F4A4A2" "#F4A3A1" "#F4A3A1" "#F4A2A0" "#F4A2A0" "#F4A19F"
[361] "#F4A09E" "#F4A09E" "#F49F9D" "#F49F9D" "#F39E9C" "#F39E9C" "#F39D9B" "#F39D9B" "#F39C9A" "#F39B9A" "#F39B99" "#F39A99" "#F39A98" "#F39998" "#F39997"
[376] "#F39897" "#F39896" "#F39796" "#F29695" "#F29695" "#F29594" "#F29593" "#F29493" "#F29492" "#F29392" "#F29291" "#F29291" "#F29190" "#F29190" "#F2908F"
[391] "#F2908F" "#F28F8E" "#F18F8E" "#F18E8D" "#F18D8D" "#F18D8C" "#F18C8C" "#F18C8B" "#F18B8B" "#F18B8A" "#F18A89" "#F18A89" "#F18988" "#F18888" "#F18887"
[406] "#F18787" "#F08786" "#F08686" "#F08685" "#F08585" "#F08584" "#F08484" "#F08383" "#F08383" "#F08282" "#F08282" "#F08181" "#F08181" "#F08080" "#F08080"
```
Getting sankey plots for atac som overlap (from lars files)

- devmap 
```{r}
overlapping_peaks_devmap <- read_tsv("../DevMapSOMpeakOverlapCount.tsv")
Ssal_clusters <- c(21,22,16,17,11,7,6,1,2,8,4,5,10,20,14,24,25)
Omyk_clusters <- c(16,17,21,6,7,1,19,2,3,5,18,23,24,20,14,15,25)

overlapping_peaks_devmap_fs <- overlapping_peaks_devmap[which(overlapping_peaks_devmap$cluster_Ssal %in% Ssal_clusters),]
overlapping_peaks_fst <- overlapping_peaks_devmap_fs[which(overlapping_peaks_devmap_fs$cluster_Omyk %in% Omyk_clusters),]
overlapping_peaks_fst$cluster_Omyk <- factor(paste("t",overlapping_peaks_fst$cluster_Omyk,sep = ""),levels = paste("t",c(16,17,21,6,7,1,19,2,3,5,18,23,24,20,14,15,25),sep = ""))
overlapping_peaks_fst$cluster_Ssal <- factor(paste("s",overlapping_peaks_fst$cluster_Ssal,sep = ""),levels = paste("s",c(21,22,16,17,11,7,6,1,2,8,4,5,10,20,14,24,25),sep = ""))

ggplot(data = overlapping_peaks_fst,
       aes(axis1 = cluster_Ssal, 
           axis2 = cluster_Omyk, y = n)) +
  geom_alluvium(aes(fill = n),alpha = 0.7) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal() + scale_fill_gradient(low = "lightgrey",high = "coral4")

```
- bodymap

```{r}
overlapping_peaks_bodymap <- read_tsv("../BodyMapSOMpeakOverlapCount.tsv")
Ssal_clusters_b <- c(22,23,7,6,1,
                    # 5,17,3,8,
                     13,19,25,24)
Omyk_cluster_b <- c(3,2,17,16,21,20,25)

overlapping_peaks_bodymap_fs <- overlapping_peaks_bodymap[which(overlapping_peaks_bodymap$cluster_Ssal %in% Ssal_clusters_b),]
overlapping_peaks_bodymap_fst <- overlapping_peaks_bodymap_fs[which(overlapping_peaks_bodymap_fs$cluster_Omyk %in% Omyk_cluster_b),]

overlapping_peaks_bodymap_fst$cluster_Ssal <- factor(paste("s",overlapping_peaks_bodymap_fst$cluster_Ssal,sep = ""),levels = paste("s",Ssal_clusters_b,sep = ""))
overlapping_peaks_bodymap_fst$cluster_Omyk <- factor(paste("t",overlapping_peaks_bodymap_fst$cluster_Omyk,sep = ""),levels = paste("t",Omyk_cluster_b,sep = ""))

ggplot(data = overlapping_peaks_bodymap_fst,
       aes(axis1 = cluster_Ssal, 
           axis2 = cluster_Omyk, y = n)) +
  geom_alluvium(aes(fill = n),alpha = 0.7) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal() + scale_fill_gradient(low = "lightgrey",high = "coral4")
```

